### 2018-01-12 draft.tex
Added manuscript draft version

\documentclass[wes, manuscript]{copernicus}

%% \usepackage commands included in the copernicus.cls:
%\usepackage[german, english]{babel}
%\usepackage{tabularx}
%\usepackage{cancel}
%\usepackage{multirow}
%\usepackage{supertabular}
%\usepackage{algorithmic}
%\usepackage{algorithm}
%\usepackage{amsthm}
%\usepackage{float}
%\usepackage{subfig}
%\usepackage{rotating}

\usepackage{tabularx}
\usepackage{mathtools}
\usepackage{mathrsfs}
\usepackage{amsmath}
%\usepackage[short]{optidef}
\usepackage{xcolor}
\usepackage{amsfonts}

% Colors
%---------------------------------------------------------------------------------
\definecolor{C1}{RGB}{31,119,180}
\definecolor{C2}{RGB}{255,127,14}
\definecolor{C3}{RGB}{44,160,44}
\definecolor{C4}{RGB}{214,39,40}

% Some  auxiliary commands for ease of mathematics notations
%---------------------------------------------------------------------------------
% Derivatives and differentials
\newcommand{\ds}{~\text{d}\boldsymbol{s}}
\newcommand{\dzeta}{~\text{d}\boldsymbol{\zeta}}
\newcommand{\pder}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\dx}{\text{d}\boldsymbol{x}}
\newcommand{\dt}{\text{d}t}
\newcommand{\ddt}[1]{\frac{\text{d} #1}{\text{d} t}}
% Integrals
\newcommand{\stint}{\int_{0}^{T} \int_{\Omega}}
\newcommand{\sint}{\int_{\Omega}}
\newcommand{\Tint}{\int_{0}^{T}}
\newcommand{\tint}{\int_{0}^{t}}
% Filtered variables
\newcommand{\utilde}{\widetilde{\bs{u}}}
\newcommand{\ptilde}{\widetilde{p}}
\newcommand{\ctnhat}[1]{\widehat{C}_{T,#1}'}
\newcommand{\ctihat}{\widehat{C}_{T,i}'}
\newcommand{\cthat}{\widehat{C}_T'}
% Other symbols
\newcommand{\ctmax}{C_{T,\text{max}}'}
\newcommand{\cti}{C_{T,i}'}
\newcommand{\ctn}[1]{C_{T,#1}'}
\newcommand{\R}{\mathscr{R}}
\newcommand{\J}{\mathscr{J}}
\newcommand{\Jtilde}{\tilde{\mathscr{J}}}
\newcommand{\Jgrad}{\nabla \Jtilde}
\newcommand{\Lagr}{\mathscr{L}}
\newcommand{\eperp}{\bs{e}_\perp}
\newcommand{\eperpi}{\bs{e}_{\perp,i}}
\newcommand{\etransi}{\bs{e}_{\parallel,i}}
\newcommand{\ex}{\bs{e}_x}
\newcommand{\ey}{\bs{e}_y}
\newcommand{\ez}{\bs{e}_z}
\newcommand{\eperpn}[1]{\bs{e}_{\perp,#1}}
\newcommand{\vi}{\frac{1}{A_i} \sint \R_i (\bs{s})~\utilde \cdot \eperpi \ds}
% Operators
\newcommand{\innerproduct}[2]{\bigg( #1, #2 \bigg)}
\newcommand{\innerproductsmall}[2]{\big( #1, #2 \big)}
\newcommand{\sumturbines}{\sum_{i=1}^{N_t}}
\newcommand{\diracdelta}{{\delta}}
\DeclareMathOperator*{\argmin}{arg\,min}

\begin{document}

\title{Towards practical dynamic induction control of wind farms: analysis of optimal control dynamics and mimicry of first-row behavior}
\runningtitle{Towards practical dynamic induction control of wind farms}

\Author[]{Wim}{Munters}
\Author[]{Johan}{Meyers}
\affil[]{Department of Mechanical Engineering, KU Leuven, Celestijnenlaan 300A, 3001 Leuven, Belgium}
\runningauthor{Munters and Meyers}
\correspondence{wim.munters@kuleuven.be}

\received{}
\pubdiscuss{} %% only important for two-stage journals
\revised{}
\accepted{}
\published{}

\firstpage{1}
\maketitle

\begin{abstract}
	Wake interactions between wind turbines in wind farms entail reduced energy extraction in downstream rows, inciting research into coordinated wind-farm control strategies for optimal power extraction. In recent work, optimization and large-eddy simulation were combined to study optimal dynamic induction control of wind farms with power gains up to 20\% (Munters \& Meyers 2017 \emph{Phil Trans R Soc A} \textbf{375}, 20160100, doi:10.1098/rsta.2016.0100). However, computational costs associated with the optimal control simulations impede practical implementation of this work. Furthermore, the resulting optimal control signals are tuned to specific flow events and hence cannot be used in practice, and the complexity of these signals complicated the derivation of simplified control strategies. The current work focuses on further analysis of the optimization results of Munters \& Meyers. The analysis shows that wind-farm controls are optimized in a uni-directional manner with little upstream propagation of information, and that turbines can be classified into first-row, intermediate-row, and last-row turbines based on their optimal control dynamics. Furthermore, it is found that optimally-controlled first-row turbines periodically shed vortex rings leading to enhanced wake mixing. This behavior is mimicked with a simple sinusoidal thrust control strategy, resulting in robust power gains for turbines in the entrance region of the farm.
\end{abstract}

\copyrightstatement{TEXT}

\introduction  %% \introduction[modified heading if necessary]
Wake interactions between turbines within a wind farm cause reduced power extraction and increased turbine loading in downstream rows. The current control paradigm in such farms optimizes performance at the wind-turbine level and does not account for these interactions, resulting in sub-optimal wind-farm efficiency. In contrast, control strategies at the farm level allow to influence wake interactions and promise to improve overall wind-farm performance: by redirecting propagating wakes (yaw control, see e.g. Refs YYY) or by affecting the induced wake velocity deficits (axial induction control, see e.g. Refs XXX) wind conditions for downstream turbines can be improved significantly. In recent work, Munters \& Meyers \cite{munters2017optimal} ...

Korte beschrijving van methodologie, sleutelresultaat, problemen, zie abstract...

The current paper presents efforts on understanding optimal control dynamics and identifying the flow mechanisms behind the power increase observed in the optimal induction control simulations discussed in Chapter~\ref{ch:opt_induction}, with the aim of harnessing them in tractable dynamic induction wind-farm controllers. Throughout the analysis, the focus will lie on the overinductive fast-response case C3t5, as it was observed to exhibit similar flow statistics as the maximum yield case C3t0, but with smoother thrust force dynamics.

The analysis presented in the current chapter will lead to the identification of a coherent flow mechanism in the wake of optimally controlled first-row turbines, responsible for part of the power increase in the optimal control simulations. This resulted in a new derived simplified dynamic axial induction control strategy for first-row turbines producing a robust power increase in the second row, irrespective of precise local flow details. However, it will be shown that the controls responsible for the remainder of the power gains in the optimal control simulations are tuned to specific flow conditions. This complicates the analysis of downstream rows, and has thus far impeded the derivation of similar control strategies for the latter.

The outline of the chapter is as follows: first, Section~\ref{sec:analysis_controls_analysis} presents an analysis of the control and thrust force dynamics, and performs some numerical experiments to elucidate characteristics of the optimal controls. As expected from the flow field statistics in previous chapter, it will be shown that the coherent behavior of turbines situated in the first row of the wind farm stands out from their downstream counterparts. Hereafter, Section~\ref{sec:analysis_flow_vis} identifies the shedding of vortex rings from the first row based on a flow visualization. Further, a simple sinusoidal thrust control approach is presented that successfully mimics this process with a robust increase in power extraction extraction in the second row. Next, Section~\ref{sec:analysis_intermediate} shortly discusses the behavior of the intermediate rows, i.e. turbines which have both upstream and downstream neighbors, for which similar simple control strategies have not been identified thus far. In conclusion, Section~\ref{sec:analysis_summ} summarizes the main findings of this chapter.

\section{Description of optimal control simulations}\label{sec:optimal_control_sims}
The current section describes the optimal control simulations performed by Munters \& Meyers \cite{munters2017optimal}, of which the results are further analyzed in the current paper. First, the methodology and numerical setup is detailed. Afterwards, the simulation results are discussed.

\subsection{Methodology \& numerical setup}

Optimal control based on LES and optimization

Receding-horizon framework (maak 1 gecombineerde figuur)

Optimal control problem

\begin{alignat}{4}
& \underset{\bs{\varphi}, \bs{q}}{\text{minimize}}  & \qquad  \J(\bs{\varphi}, \bs{q}) &= - \Tint \sum_{i=1}^{N_t} P_i ~\dt  & \label{eq:costfunction}\\
& \text{s.t.}                      			&         \frac{\partial \utilde}{\partial t} + \big(\utilde \cdot \nabla \big)~ \utilde &= - \nabla (\ptilde + \ptilde_\infty) / \rho - \nabla \cdot \boldsymbol{\tau}_{sgs} + \sum_{i=1}^{N_t} \bs{f}_i \qquad  & \text{in } \Omega \times (0,T], \label{eq:NSmomentum_constraint} \\
&                                                   &        \nabla \cdot \utilde&=0 									        & \text{in } \Omega \times (0,T], \label{eq:NScontinuity_constraint}\\
&                                                   &        \tau \ddt{\ctihat}&=\cti - \ctihat 								& i=1\dots N_t~\text{in } (0,T],  \label{eq:ctihat_constraint}\\
&                                                   &        C_{T,\text{min}}' \leq~ &\cti \leq C_{T,\text{max}}'				& i=1\dots N_t~\text{in } (0,T],  \label{eq:boxct_constraint}
\end{alignat}
Define turbine forces and powers!

Focus on C3t5 case.

\subsection{Simulation results}

\section{Thrust coefficient analysis and numerical experiments}\label{sec:analysis_controls_analysis}

The current section focuses on the optimal thrust coefficients generated by the optimal control simulations in the previous chapter, and performs numerical experiments to uncover some of the characteristics of these controls and their dependence on the local flow conditions.

First, the thrust coefficient signals themselves are analyzed in Section~\ref{sec:analysis_controls_analysis}. Second, the optimized thrust coefficients are applied only to subsets of turbine rows in Section~\ref{sec:application_subset}. In this way, the interdependency of optimized thrust coefficients in different rows can be evaluated. Third, additional optimal control simulations, in which only one single active row is optimized, are discussed in Section~\ref{sec:optimization_single}. These optimizations provide an indication on how increased power potential is distributed among the rows, and allows to compare the resulting single-row optimized thrust coefficients with the fully cooperative coefficients from case C3t5. Fourth, Section~\ref{sec:modified_thrust} evaluates the dependency of optimized thrust coefficients on specific flow conditions. This is done by running test simulations in which the coefficients are modified to eliminate any possible correlations to specific flow events at given times. Finally, Section~\ref{sec:analysis_discussion} discusses the main conclusions from the abovementioned sections, and summarizes the lessons learned.

\subsection{Analysis of thrust coefficient signals}\label{sec:analysis_thrust}

Figure~\ref{fig:controls} illustrates the time evolution of some of  the thrust coefficients $\cthat$ in the C3t5 case. The figure illustrates that, for all rows but the last one (i.e. R12), $\cthat$ varies significantly in time, and that the amplitudes and frequencies of these variations are somewhat higher in the upstream rows of the farm. In contrast, row 12 features only minor unsteadiness at lower frequencies, and has an increased mean value of $\cthat$. This relatively steady behavior of the last row can be explained by the fact that there are no further downstream turbines that could benefit from row 12 actively influencing local flow conditions, hence the row optimizes its own power greedily. The increase in mean $\cthat$ can be explained based on Figure~\ref{fig:ct_sweep}, which illustrates the power extraction as a function of steady $\cthat$ for unwaked turbines, subject to identical turbulent inlet as in case C3t5. Although momentum theory predicts maximal power extraction for steady uniform inflow at $\cthat = 2$, the actual optimal steady value for the ADM at the current spatial resolution lies somewhat higher at $\cthat \approx 2.4$, for which power extraction is about 1.4\% higher than at $\cthat = 2$. This behavior is related to the overprediction in ADM power due to the diffuse turbine representation on typical simulation grids discussed in Section~\ref{sec:meth_adm}: the mass flow through the rotor disk at $\cthat > 2$ is slightly too high compared to momentum theory, resulting in a  shift of optimal $\cthat$ towards somewhat higher values. Although the linear fit $C_P' = a \cthat$ introduced in Appendix~\ref{ch:app_adm} eliminates the error in maximal power extraction, it does not correct the value of the optimal $\cthat$ (note that this could be achieved through a more complex relation between $C_P'$ and $\cthat$). Even though this explains the slight increase in mean $\cthat$ for greedily optimized last-row turbines, this does not undermine the overall power gains of the overinductive cases from previous chapter, which all feature power gains well over 10\%. Returning to the more complex thrust coefficients in the other rows it is worth noting that, based on the current dataset, no statistically significant correlations between thrust coefficients of different turbines could be found. Furthermore, attempts towards linking thrust coefficient dynamics to upstream flow measurements (e.g. velocities, shear or kinetic energy) through linear regression models and random forest regressors have been unsuccessful to date.

\begin{figure}
	\includegraphics[width=\textwidth]{controls.eps}
	\caption{Time evolution of the thrust coefficient $\cthat$ for a selection of optimally controlled turbines  of case C3t5. \emph{Left: } Total time horizon. \emph{Right: } Zoomed view, including setpoint $C_T'$ in gray. \label{fig:controls}}
\end{figure}

\begin{figure}
	\includegraphics[width=\textwidth]{sweep_ct.eps}
	\caption{Normalized power extraction as a function of steady thrust coefficient $\cthat$ for wind turbines subject to the same freestream turbulent inflow as in case C3t5. Every dot corresponds to one LES.\label{fig:ct_sweep}}
\end{figure}

\begin{figure}
	\includegraphics[width=\textwidth]{spectra_thrust_cthat.eps}
	\caption{Power spectral density (PSD) estimates of the row-averaged thrust force $f_T$ (\emph{a}) and thrust coefficient $\cthat$ (\emph{b}) as a function of frequency $f$ (bottom axis) and non-dimensional Strouhal number $St$ (top axis). \label{fig:spectra_thrust_cthat}}
\end{figure}

Figure~\ref{fig:spectra_thrust_cthat}a,b illustrate power spectral densities of the row-averaged thrust forces and thrust coefficients respectively. The figure illustrates that the variances of both the thrust coefficients and their resulting forces are highest in row 1. Further downstream, rows 2 to 11 have very similar spectral behavior, and row 12 shows significantly lower variability. The high-frequency slopes of around $-5$ observed both for $f_T$ and $\cthat$ indicate that force variability on short timescales is caused mainly by thrust coefficient variations, whereas the slower thrust force dynamics tend more to a $-5/3$ slope, suggesting that these are governed by the unsteadiness in the turbulence instead. Note that, even though the spectra for all rows except row 12 collapse at frequencies below $0.05$ Hz, the first-row spectrum shows a small yet significant peak at $f \approx 0.02 \dots 0.03$ Hz ($f D / U_\infty \approx 0.2 \dots 0.3$) as indicated by the purple arrow, which also translates to an increased spectral content of the first row thrust force at these frequencies. It will be shown later in this chapter that variations in the thrust coefficient around this frequency can be harnessed to increase power extraction in the second row.

\subsection{Application of optimal thrust coefficients to subsets of turbine rows}\label{sec:application_subset}

In order to further study how the optimal controls increase overall wind-farm power, Figure~\ref{fig:multirow} illustrates power extraction resulting from applying a subset of the optimal controls to specific turbine rows only. Figure~\ref{fig:multirow}a depicts simulation results for which the  optimized controls are applied \emph{only to one specific row}, with the thrust coefficient in all other rows kept at the reference value of $\cthat = 2$. From the figure it can be seen that only for the controls of the first row (R1) this results in a significant power increase in rows 2 and 3. This indicates that the optimal controls, as generated by the optimization at the wind-farm level, depend on the precise flow conditions caused by upstream control actions, and can hence only be applied independently for the first row, which has no upstream dependence by definition.

Figure~\ref{fig:multirow}b shows results from simulations in which the controls are applied \emph{for all rows up to a certain row}, i.e. R1--R3 indicates the application of optimized controls generated by case C3t5 to rows 1, 2, and 3. An interesting observation from this figure is that , for any row $i$ except the last one, the power potential as observed in case C3t5 is almost fully recovered by only applying the optimal controls up to row $i-1$. This suggests that self-optimization is very limited: the optimal controls for a given turbine are designed to create favorable flow conditions in the downstream rows instead of increasing local power. Furthermore, although the discussion in the previous paragraph has shown that downstream controls are optimized with the upstream actions in mind, the converse is not true: upstream control actions do not require a specific downstream response in order to increase power.

\begin{figure}
	\includegraphics[width=\textwidth]{multirow.eps}
	\caption{Normalized row-averaged power extraction for the reference case, optimal control case C3t5, and subset control cases up to optimization window 3. \emph{a)} Subset control cases with optimal controls applied \emph{only} in one specific row. \emph{b)} Subset control cases with optimal controls applied \emph{for all rows up to} specific row. \label{fig:multirow}}
\end{figure}

\subsection{Optimization of single active turbine rows}\label{sec:optimization_single}
The previous section has shown that, based on the full-farm optimization case, the first-row controls can be applied independently, whereas this does not work for the downstream rows. To further quantify the potential for increasing wind-farm power in each row of turbines, the current section considers a set of additional optimal control cases in which \emph{only a single active row is optimized}, with all other rows remaining passive. Furthermore, by comparing the optimized controls of these cases with the full-farm optimization case C3t5, the degree of cooperation between turbines can be assessed. Note that the current single-row optimal control is not equivalent to greedy control: the optimizer still aims to increase aggregate farm power by taking into account wake interactions with downstream turbines. Furthermore, in contrast to the single-row control simulations from the previous section (i.e. in Figure~\ref{fig:multirow}a), the current optimizations will yield controls that are \emph{explicitly designed} to increase power given that all other rows are passive. To limit computational costs, the additional optimizations are only performed for a single time window.

Figure~\ref{fig:single_row_opt} illustrates the relative increase in wind-farm power extraction for each of the twelve individually optimized control cases. The optimization is run until the continuous adjoint gradient accuracy prevents further progress in the vicinity of a local minimum (see discussion accompanying Figure~\ref{fig:window_opt}). Upon interpreting the actual values from the figure, it is important to note that the reported power gain covers the full optimization horizon $T$, and is hence affected by finite-horizon effects. Furthermore, the first window of an optimal control simulation, as considered here, contains an initial dead zone corresponding to the wake advection lag before upstream turbines start influencing their downstream neighbors. This tends to reduce gains compared to later time windows. Nevertheless, the relative order of the different cases still provides information that can be generalized to full optimal control studies with multiple windows and longer time horizons.

The figure shows that the first row (R1) holds by far the most promise for optimizing wind-farm power. This is not surprising as R1 produces the most power of all rows, and typically leaves the deepest wakes, causing second-row turbines to perform  poorly in aligned wind-farm layouts (see, e.g., \citealp{porte2013numerical,nilsson2015large}; and \citealp{stevens2016effects}). At the other end of the spectrum, the last row (R12) is the least fruitful: it can only increase power by increasing $\cthat$ to harness the imperfections in the ADM for the given spatial resolution illustrated in Figure~\ref{fig:ct_sweep} (see discussion in Section~\ref{sec:analysis_thrust}), or by exploiting finite-horizon effects in which it attempts to completely halt the flow near the end of the time horizon. The intermediate rows (R2--R11) lie closer together, with the general trend being that potential is somewhat decreased with downstream distance into the wind farm, although this decrease is not monotonous.

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{single_row_opt.eps}
	\caption{Increase in wind-farm power extraction for the first (and only) timewindow of the optimal control cases in which only a single row is optimized. \label{fig:single_row_opt}}
\end{figure}

Figure~\ref{fig:gains_singlerow} illustrates the row-wise relative power increase matrix for each of the single-row optimization cases. The figure indicates that, for each of the optimization cases, the largest power increase is observed in the first row downstream of the active turbine (i.e. R$_{\rm i+1}$), and that the influence on row R$_{\rm i+3}$ is limited. This is explained by the fact that the finite optimization horizon $T = 240$ s allows for more interactions with directly neighboring turbines than with those located further downstream. Furthermore, except for the optimal control case of the last row (R12), self-optimization is virtually non-existent: power gains are achieved by modifying the flow to yield more favorable conditions in the downstream rows.

\begin{figure}
	\centering
	\includegraphics[width=0.9\textwidth]{individual_gains.eps}
	\caption{Relative power increase matrix for downstream rows in the single-row optimization case for row R$_{\rm i}$, indicated in the horizontal axis. N/A indicates non-existing downstream rows. Finite-horizon effects are eliminated by only reporting power increase up until $t = T_A$ for the active row R$_i$. \label{fig:gains_singlerow}}
\end{figure}

Turning to the thrust dynamics, Figure~\ref{fig:control_C3t5_singlerow} shows the optimized thrust coefficients for the entire optimization horizon $T$ in several turbine rows for the single-row (gray) and full-farm (black) optimizations. Interestingly, the first-row thrust coefficient is virtually identical, regardless of whether downstream turbines are active or not. Although similarities can also be observed for the downstream rows, the correspondence is much weaker than in the first row. This can be explained by the fact that, in contrast to the first row, downstream turbines experience different local flow conditions based on whether the upstream turbines are active or not.

\begin{figure}
	\centering
	\includegraphics[width=0.9\textwidth]{fullfarm_singlerow.eps}
	\caption{Optimized thrust coefficients in several rows for case C3t5 (black) and single-row optimal control cases (gray) for the full optimization horizon $T = 240$ s.\label{fig:control_C3t5_singlerow}}
\end{figure}

The current observations further confirm the conclusions from previous section: optimal control actions depend to some degree on the local flow conditions. Furthermore, previous section showed that first-row controls still lead to power gains in the absence of downstream control response. Here, as indicated by the similarity in first-row controls for single-row and full-farm optimization cases, it is shown that, for the first row, the optimizer is actually indifferent to whether or not downstream rows are actively controlled.

\subsection{Modification of thrust coefficient signals}\label{sec:modified_thrust}
The observations from previous sections illustrate that, at least to some degree, the optimized thrust coefficients are tuned to local flow conditions. In the current section, the possibility of whether the coefficients contain traits that are independent of flow conditions is nonetheless investigated. To this end, the optimized thrust coefficients are modified in such a way that correlations between them and specific flow events are eliminated. This is done in two independent test cases.

In the first case, the controls, which were specifically generated for selected turbines, are reassigned to other turbines by randomly swapping the control sets of different turbine columns. In doing so, each turbine will receive controls that were specifically designed for another turbine in the same row. To avoid erroneous conclusions based on coincidence, the column swap is performed in 2 random independent ways. Row-averaged power for these cases is shown in Figure~\ref{fig:scrambled}a.

In the second case, controls remain assigned to their original turbines, but are shuffled in time by randomly swapping optimal controls generated for different control windows. In this way, the spectral thrust characteristics for timescales smaller that the control horizon $T_A = 120$ s remain unchanged, whereas the time synchronization of control actions to specific flow events is eliminated. Similar to the first case, this is done in 2 random ways, and Figure~\ref{fig:scrambled}b illustrates the row-averaged power for these cases.

The figure shows similar behavior for each of the modified control cases: only in the second row (R2), a consistent increase in power extraction can be observed. Combining this observation with the findings of previous sections, i.e. that power gains in a given row are caused by upstream control action instead of self-optimization, the current test cases suggest the presence of flow-invariant	features in the control signals of the first row. Note however that the full power gain in the second row is only partially attained, indicating that also the first turbine row reacts to the specific flow conditions. The small power increase in the last row (R12) can again be explained by the slightly higher mean value in $\cthat$. Based on the current data, no consistent gains in any of the other rows are found.

\begin{figure}
	\includegraphics[width=\textwidth]{scrambled.eps}
	\caption{Normalized row-averaged power extraction for the reference case, optimal control case C3t5, and modified control cases. \emph{a)} Modified control cases with controls swapped between wind-turbine columns. \emph{b)} Modified control cases with controls swapped randomly by time window. \label{fig:scrambled}}
\end{figure}

\subsection{Discussion}\label{sec:analysis_discussion}

The observations and experiments from previous sections have revealed information that increases the understanding the optimized thrust coefficients, and can be used as a starting point towards designing practical wind-farm controllers that do not require computationally prohibitive LES-based optimal control simulations. A first conclusion that can be drawn from the analysis above is that wind turbines can be classified into three distinct categories, based on their position within the farm: first-row turbines, last-row turbines, and intermediate turbines.

The most salient behavior can be found in the first-row turbines (R1). Firstly, Figure~\ref{fig:spectra_thrust_cthat} showed that these turbines exhibit the largest variability in thrust coefficients and resulting thrust forces. Second, Figure~\ref{fig:multirow}a has shown that the optimized thrust coefficients in R1 are the only ones that can be applied independently from the other rows, and Figure~\ref{fig:single_row_opt} illustrated that the first row holds the greatest potential for power optimization of all rows in the farm. Third, the first row is the only one that has been shown to retain part of its power increase after eliminating the correlation between the optimized thrust coefficients and the flow field in Figure~\ref{fig:scrambled}, suggesting a flow-invariant component in the optimized coefficients. Finally, it is worth remarking the observed peak around $fD/U_\infty \approx 0.2 \dots 0.3$ from Figure~\ref{fig:spectra_thrust_cthat}. It will be shown below that variations in the thrust coefficient are related to specific flow events in the wake of R1.

The characteristics of the last row turbines (R12) also stand out from the rest. This is mainly caused by the fact that, by definition, the last row has no downstream turbines for which it can generate more favorable flow conditions, causing them to operate greedily in an optimal control setting. This explains the relatively quiescent behavior of optimized thrust coefficients in Figure~\ref{fig:controls}, and the limited potential for power increase in Figure~\ref{fig:single_row_opt}.

The remaining intermediate rows (R2--R11) have similar spectral thrust characteristics (Figure~\ref{fig:spectra_thrust_cthat}) and potential for power increase (Figure~\ref{fig:single_row_opt}), situated in between but clearly separated from first- and last-row turbines. Further, it is worth noting that the behavior and analysis of control actions in these turbines is most complex: not only do they influence downstream turbines, they in turn are dependent on controls of upstream turbines.

A second conclusion is that, whether or not the wind-farm is controlled with the possibility of active response and cooperation between turbines, the resulting power and thrust characteristics are very similar. Firstly, both Figure~\ref{fig:multirow}b (full-farm optimization) and Figure~\ref{fig:gains_singlerow} (single-row optimization) suggest that self-optimization of turbines is negligible, and that power gains are mainly achieved by generating more favorable flow conditions for downstream turbines. Furthermore, the remarkable match between optimized thrust coefficients for full-farm and single-row optimal control cases from Figure~\ref{fig:control_C3t5_singlerow} indicates that, at least for the first row, upstream controls do not rely on active downstream repose Instead, Figure~\ref{fig:multirow}b shows that, for any row $i$, the full potential in power increase is virtually attained by applying controls only for the upstream turbines up until row $i-1$. Furthermore, Figure~\ref{fig:multirow}a shows that optimized thrust coefficients in downstream rows by themselves do not result in power increase without also applying optimized coefficients in their upstream neighbors. This indicates that controls are tuned to specific flow events. This also explains the discrepancies between full-farm and single-row optimized thrust coefficients for downstream rows in Figure~\ref{fig:control_C3t5_singlerow}. 
These observations strongly suggest that the optimized thrust coefficients are designed in a parabolic manner, i.e. with a unidirectional propagation of control information from the first row to the last, and very little upstream influence of downstream turbine actions.

With this in mind, the remainder of this chapter will focus on the first and most promising link of the control chain: the turbines situated in the first row of the wind farm.

\section{Flow visualization and mimicry of first-row behavior}\label{sec:analysis_flow_vis}
As shown in the previous section, the behavior of turbines within the first upstream row stands out from the others. Firstly, they are not influenced by any dynamic actions of the upstream turbines which makes their behavior somewhat easier to study. Secondly, they are unwaked and hence subjected to completely different flow characteristics than their downstream counterparts leading to much larger power extraction in the first row than downstream. Third, it was shown that the first-row controls show some inherent characteristics that are not tuned to specific flow events, and hence hold promise for simple and universally applicable control strategies. Therefore, the current section further focuses on the analysis of the first row.

First, a qualitative analysis of the instantaneous flow field is performed in Section~\ref{sec:opt_visualization}, resulting in the observation of vortex rings being shed from first-row turbines. Thereafter, this mechanisms is mimicked by imposing sinusoidally varying thrust coefficients in these turbines in Section~\ref{sec:opt_sinus}, with the aim of increasing power through similar mechanisms as in the computationally expensive optimal control cases.

\subsection{Flow field visualization}\label{sec:opt_visualization}
With the aim of visualizing instantaneous flow field differences between controlled and reference cases, Figure~\ref{fig:vorticity_windfarm} illustrates snapshots of the velocity field at t = 300 s for the reference case (left) and the optimal control case C3t5 (right). Figures~\ref{fig:vorticity_windfarm}a,b show isosurfaces of vorticity magnitude, colored by streamwise velocity $\widetilde{u}_x$. Further, volumes of very low streamwise velocity (at $\widetilde{u}_x < 4.5 $ m s$^{-1}$) are rendered in black. Figure~\ref{fig:vorticity_windfarm}a shows that, in the reference case, the first-row turbines shed relatively stable vortex sheets that demarcate the wake from the freestream flow. The sheets destabilize and break up as they are advected downstream, resulting in complex three-dimensional vortical structures. Furthermore, as also shown in Figure~\ref{fig:vorticity_windfarm}c, wake mixing is limited, and downstream turbines experience reduced velocities. In contrast, the optimized case shows coherent vortex rings being shed from the first-row turbine. As indicated in Figure~\ref{fig:vorticity_windfarm}b, the locations of the rings in the controlled case coincide with naturally occurring bulges in the vortex sheet of the reference case: the controlled turbines further destabilize the sheet by well-timed temporal variations in its thrust coefficient. As illustrated in Figure~\ref{fig:vorticity_windfarm}c, this results in smaller velocity deficits in the wake region. Note that, downstream of the second turbine, the vorticity field becomes much more complex and differences in the flow fields are less coherent.

The observed shedding of ring vortices seems to occur at specific flow-synchronized times to exploit natural instabilities in the original vortex sheets. Therefore, the remainder of this chapter will attempt to accomplish the same effect by simple sinusoidal thrust variations.

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{vorticity.eps}
	\caption{Instantaneous snapshots at t = 300 s of portion of wind-farm flow field for the reference (left) and optimized (right) case. \emph{a,b) } Isosurface of vorticity magnitude, colored by streamwise velocity $\widetilde{u}_x$. Deep wake regions (with $\widetilde{u}_x < 4.5$ m s$^{-1}$) are rendered in black. \emph{c)} Contours of streamwise velocity $\widetilde{u}_x$. Coloring is in units of m s$^{-1}$. Wind turbines are represented as gray disks. \label{fig:vorticity_windfarm}}
\end{figure}

\subsection{Sinusoidal thrust variations}\label{sec:opt_sinus}
The aim of the current section is to mimic the quasi-periodic shedding of vortex rings by upstream turbines as observed above through the use of simple periodic variations in the thrust coefficient. To this end, the Betz-optimal coefficient $\cthat = 2$ is perturbed by a sinusoidal signal, parametrized by its amplitude $A$, and a non-dimensional frequency in the form of the Strouhal number $St = f D/ U_\infty$:
\begin{equation}
\cthat(t) = 2 + A \sin \bigg(2\pi St \frac{t U_\infty}{D} \bigg).\label{eq:define_thrust}
\end{equation}

\subsubsection{Parameter sweep}
A parameter sweep is performed for both $A$ and $St$ on a reduced-size wind-farm LES, as illustrated in Figure~\ref{fig:sinus_setup}. The farm consists of $4 \times 4$ turbines in an aligned layout with $S = 6D$ in both streamwise and spanwise directions, geometrically equivalent to the optimally controlled wind farm from Chapter~\ref{ch:opt_induction}. The simulation is performed on a domain of $4 \times 2.4 \times 1$ km$^3$ with a simulation grid of $192 \times 256 \times 144$ gridpoints. Inflow conditions differ from those applied in the aforementioned simulations, and are generated by running an auxiliary shifted-periodic precursor simulation on a domain identical to the main simulation domain. All other simulation setup parameters are the same as those defined in Chapter~\ref{ch:opt_induction}. A set of wind-farm flow simulations is advanced in time by 30 minutes, during which the front row is controlled using a sinusoidally varying thrust coefficient $\cthat$, as defined in Equation \eqref{eq:define_thrust}. Within this set, the amplitude $A$ is varied between 0.5 and 2, with increments of 0.5. Furthermore, the Strouhal number $St$ is ranged between 0.05 and 0.6, with increments of 0.05. In total, this leads to 48 LES cases within the set.

\begin{figure}
	\centering
	\includegraphics[width=0.75\textwidth]{setup_sinusoidal.eps}
	\caption{Reduced $4 \times 4$ wind-farm simulation setup for sinusoidal variation parameter study showing instantaneous contours of streamwise velocity $\widetilde{u}_x$. Coloring is in units of m s$^{-1}$. The dashed line indicates the start of the fringe region for imposition of unwaked inflow conditions. \label{fig:sinus_setup}}
\end{figure}

Figure~\ref{fig:sinus_baseline} illustrates the power extraction for all cases considered. Figure~\ref{fig:sinus_baseline}a illustrates the relative power gains over the reference case. From the figure it can be seen that there is a well-defined range of values for $A$ and $St$ for which wind-farm power can be increased substantially through upstream sinusoidal thrust variations, with a maximal power increase of $\approx$ 5\% at $(St^\star, A^\star) = (0.25, 1.5)$. In order to put the technical feasibility of these parameters into context, note that a Strouhal number $St = 0.25$ corresponds here to a sine wave period of $\approx$ 50 s for a turbine with diameter $D = 100$ m and a freestream velocity $U_\infty = 8.5$ m s$^{-1}$. Furthermore, although higher than the $\ctmax$ applied in the optimal control cases of the previous chapter, the resulting maximum thrust coefficient of 3.5 could e.g. be attained by an NREL 5MW rotor with a 50\% increase in chord length (see \cite{goit2015optimal}, Appendix A). Figure~\ref{fig:sinus_baseline}b illustrates normalized power extraction by row for the reference case, the best sinusoidal case, and the first four rows of the optimal control case C3t5 from previous chapter. The figure shows that the power gain in the sinusoidal cases originates mostly from the second row, and that power in the first row is decreased by approximately 5\%. In contrast, optimal control case C3t5, in which all rows are active, also increases power in rows 3 and 4, and reduces first-row power by only 1\%. 
\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{gains_turbulent_6D_6D_6D_wide_inphase2.eps}
	\caption{Power extraction of baseline sinusoidal thrust case ($S = 6D$, $z_0/H = 10^{-4}$). \emph{a) } Relative gain in mean wind-farm power extraction over reference case as a function of sine amplitude $A$ and frequency $St$. \emph{b) } Row-averaged mean power extraction for the best sinusoidal case, the optimal control case C3t5 from Chapter~\ref{ch:opt_induction}, and the reference case, normalized by first-row reference power.\label{fig:sinus_baseline} }
\end{figure}

Figure~\ref{fig:vorticity_parametersweep} illustrates instantaneous vorticity and axial velocities for a set of wind turbines of the aforementioned reference case (a), the best sinusoidal case with $(St, A) = (St^\star, A^\star)= (0.25, 1.5)$ (b), and a sinusoidal case which does not lead to increased power extraction with $(St, A) = (0.6, 2)$ (c). The figure illustrates that sinusoidal variations in the first-row thrust coefficient indeed cause periodic shedding of vortex rings. Figure~\ref{fig:vorticity_parametersweep}b shows that, at the optimal frequency, this leads to increased wake mixing, providing the second-row turbine with a higher incoming velocity. In contrast, Figure~\ref{fig:vorticity_parametersweep}c shows that, even though higher frequency thrust oscillations also result in periodic shedding of vortex rings, this does not automatically lead to more favorable flow conditions for downstream turbines. 
\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{pumpcases2.eps}
	\caption{Snapshots of wind-farm flow fields at t = 1 800 s. \emph{Left: } Isocontours of vorticity, colored by streamwise velocity. \emph{Right: } Contours of streamwise velocity. \emph{a)} Reference. \emph{b)} Best sinusoidal case, with $(St , A) = (St^\star, A^\star) = (0.25, 1.5)$. \emph{c)} Sinusoidal case with $(St, A) = (0.6, 2)$ \label{fig:vorticity_parametersweep}  }
\end{figure}

In order to assess whether the same strategy can be used in the downstream turbines as well, Figure~\ref{fig:sinus_row2} illustrates the results from an identical parameter sweep as discussed above, except that here the second turbine row is controlled using a sinusoidal thrust coefficient. Figure~\ref{fig:sinus_row2}a indicates that sinusoidal actuation of the second row invariably leads to losses in wind-farm power. Figure ~\ref{fig:sinus_row2}b  shows that, for the optimal combination of parameters of  $(St, A) = (0.25, 1.5)$ as reported for first-row actuation above, the minor power increase in the third row does not compensate for additional losses in the second and fourth row. This shows that the proposed simple control strategy does not work when applied to waked turbines, and that more elaborate control strategies are required to harness the gains achieved by the optimal control simulation in the downstream regions of the farm. 
\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{gains_row22.eps}
	\caption{Power extraction of second-row sinusoidal thrust case . \emph{a) } Relative gain in mean wind-farm power extraction over reference case as a function of sine amplitude $A$ and frequency $St$. \emph{b) } Row-averaged mean power extraction for the sinusoidal case with the optimal parameters as for row 1 sinusoidal thrust, and the reference case, normalized by first-row reference power.\label{fig:sinus_row2} }
\end{figure}

It is interesting to note that, even though the current parameter sweep is performed using different initial and inlet conditions, the optimal frequency of sinusoidal variations in $\cthat$ at $St^\star = f D / U_\infty = 0.25$ corresponds to the location of the peak in the first-row thrust coefficient spectrum in Figure~\ref{fig:spectra_thrust_cthat}. This suggests that the current vortex-shedding mechanism is related to the flow-invariant characteristics for increasing power reported in Figure~\ref{fig:scrambled}. 
In the following paragraphs, the robustness of the best parameter pair for first-row thrust variations, i.e. $(St, A) = (0.25, 1.5)$, is investigated with the aim of assessing the general applicability of this control strategy. To this end, similar parameter sweeps are performed for cases with varying turbine spacings and turbulence intensities.

\subsubsection{Robustness with respect to turbine spacing and turbulence intensity}

Figure~\ref{fig:sinus_spacing} illustrates the power extraction resulting from two parameter sweeps with streamwise turbine spacings of $5D$ and $7D$ respectively. Results are promising: for the given cases, $St^\star$ and $A^\star$  do not depend on streamwise turbine spacing. Furthermore, even in the $7D$ spacing case (Figure~\ref{fig:sinus_spacing}c-d), which naturally features lower overall power losses in downstream rows, power extraction in the second row can be significantly increased through sinusoidal variations in the first-row thrust coefficient. 
\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{gains_combined2.eps}
	\caption{Power extraction of sinusoidal thrust cases with varying streamwise spacing. \emph{Top} (\emph{a,b}): Decreased spacing at $S = 5D$. \emph{Bottom} (\emph{c,d}): Increased spacing at $S = 7D$.  \emph{Left} (\emph{a,c}): Relative gain in mean wind-farm power extraction over reference case as a function of sine amplitude $A$ and frequency $St$. \emph{Right} (\emph{b,d}): Row-averaged mean power extraction for the best sinusoidal case and the reference case, normalized by first-row reference power. \label{fig:sinus_spacing} }
\end{figure}

Figure~\ref{fig:sinus_roughness} depicts the power extraction results from a parameter sweep with the same wind-farm layout as in the baseline case, but with a tenfold increase in roughness length, i.e. $z_0 = 10^{-3}H$. This results in a turbulence intensity of $\approx$ 16\%, compared to $\approx$ 10\% in the baseline case. Again, the best parameter combination of $(St^\star, A^\star) = (0.25, 1.5)$ remains unchanged. Further, even for this higher turbulence case, in which downstream losses are lower due to naturally better wake mixing, power is significantly increased in the second row, leading to a relative gain in wind-farm power of around 2\%. 
\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{gains_turbulent_6D_6D_6D_wide_antiphase_z0_1m2.eps}
	\caption{Power extraction of sinusoidal thrust case with increased wall roughness $z_0/H = 10^{-3}$. \emph{a) } Relative gain in mean wind-farm power extraction over reference case as a function of sine amplitude $A$ and frequency $St$. \emph{b) } Row-averaged mean power extraction for the best sinusoidal case and the reference case, normalized by first-row reference power.\label{fig:sinus_roughness} }
\end{figure}

%\begin{figure}
%	\centering
%	\includegraphics[width=\textwidth]{pump.eps}
%	\caption{a = reference, b = row 1, c= row 2\label{fig:vorticity_pump}}
%\end{figure}

As evidenced above, periodic sinusoidal variations of first-row thrust coefficients substantially increase power extraction in the second row, resulting in a net gain in total power for the considered $4\times4$ wind farm. Moreover, different simulation sets indicate that, at least for the range considered here, the best values for Strouhal number and amplitude of these variations, i.e. $(St^\star, A^\star) = (0.25, 1.5)$, are robust with respect to turbine spacing and turbulence intensity.

\subsubsection{Full-scale wind-farm test}
In the remainder of this section, the sinusoidal variation strategy will be tested in a full-scale wind farm, corresponding to the $12 \times 6$ aligned wind farm of Chapter~\ref{ch:opt_induction}. Simulations are performed for a reduced range of amplitudes and Strouhal numbers, corresponding to the most favorable region identified in the parameter sweeps above. In order to increase statistical convergence, the time horizon for each simulation is extended to a physical time of 10 hours.

Figure~\ref{fig:sinus_fullscale} shows the power extraction of the full-scale test. Figure~\ref{fig:sinus_fullscale}a and~\ref{fig:sinus_fullscale}b show the relative power gains over the reference case for the full wind farm and the first four turbine rows respectively. A first observation is that the optimal pair $(St^\star, A^\star)$ differs for both measures, suggesting that the increased power in front rows entails losses in the downstream. This is confirmed by the row-wise power shown in Figure~\ref{fig:sinus_fullscale}c: from the fifth row onwards, power is slightly reduced in the sinusoidal cases. This explains the negligible increase in total farm power observed in Figure~\ref{fig:sinus_fullscale}a.

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{gains_fullscale.eps}
	\caption{Power extraction of full-scale sinusoidal thrust case ($S = 6D$, $z_0/H = 10^{-4}$). \emph{a) } Relative gain in mean wind-farm power extraction over reference case as a function of sine amplitude $A$ and frequency $St$. \emph{b) } relative gain in mean power extraction for the first four rows. \emph{c) } Row-averaged mean power extraction for the best sinusoidal case, the optimal control case C3t5 from Chapter~\ref{ch:opt_induction}, and the reference case, normalized by first-row reference power.\label{fig:sinus_fullscale} }
\end{figure}

The top panel of Figure~\ref{fig:cross_section_sinus} shows cross sections of time-averaged axial velocities $\widetilde{U}_x$ at the rotor disk locations for the reference case. Further, the middle and bottom panel illustrate deviations from the reference velocity for the $(St^\star, A^\star)$ sinusoidal case and the optimal control case C3t5 respectively.  The figure shows that both controlled cases show similar characteristics at the second turbine row, with an increased axial velocity at the rotor disk, accompanied by decreased velocities above and below. Downstream, it can be seen that the passive turbines of the sinusoidal case fail to retain increased velocities at the rotor disks, instead resulting in slightly lower disk velocities starting from the fifth row. In contrast, case C3t5, in which all turbines are actively controlled, succeeds to attain similar cross section characteristics with higher rotor velocities in the downstream as well.

The current strategy hence succeeds in increasing total farm power for relatively short sets of aligned wind turbines. In contrast, larger wind farms require activity in downstream turbines as well to achieve significant gains in power extraction. However, it was found that the simple sinusoidal control strategy cannot be used in the intermediate rows. They are further discussed below in Section~\ref{sec:analysis_intermediate}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{frontviews_allc3t5.eps}
	\caption{Cross section of time-averaged axial velocity $\widetilde{U}_x$ at rotor locations in rows 1, 2, 3, 4, 5, and 12. \emph{Top: } Reference case. \emph{Middle: } Difference between best sinusoidal perturbation case (with $A = 1.5$, $St = 0.25$) and reference case. \emph{Bottom: } Difference between optimal control case C3t5 and reference case. Coloring is in units of m s$^{-1}$. \label{fig:cross_section_sinus}}
\end{figure}

As shown throughout the current section, a qualitative analysis of instantaneous flow features in the optimal induction control wind farm from Chapter~\ref{ch:opt_induction} has led to the identification of a sinusoidal thrust control strategy for first-row turbines, resulting in increased power extraction in the second row. In contrast to earlier work on steady axial induction control, for which power gains have often been marginal or non-existent, the current analysis resulted in a robust axial induction control strategy with significant power gains, independent of specific local flow conditions. To the best knowledge of the author, this is the first time that optimal control simulations of wind farms are used to come up with simple control strategies, hence marking an important step towards practical dynamic axial induction control for increased power extraction. However, important comments should be made. First, sustained sinusoidal thrust variations with a large amplitude could contribute significantly to turbine fatigue loading. Hence, structural aspects should be taken into account upon evaluating the practical viability of the approach. Second, even though experiments have shown that, for practically relevant tip speed ratios, wind turbines shed vortices in a similar way as disk-like bluff bodies \citep{medici2006measurements}, the current behavior could still be an artifact of the relatively simple ADM used throughout this study. Further verification using higher fidelity wind turbine models, such as actuator line models, and wind-tunnel testing is hence necessary.

\section{Intermediate-row behavior}\label{sec:analysis_intermediate}
The current section discusses the intermediate rows. It was shown that, without active participation in these rows, upstream gains are lost in downstream rows, and only full optimal control succeeds in achieving significant gains in downstream rows as well (see Figures~\ref{fig:sinus_fullscale},~\ref{fig:cross_section_sinus}). It was already mentioned that the analysis and behavior of turbines within intermediate rows is more complex than in the first row: they aim to influence the flow to the benefit of downstream rows but are also dependent on the actions of upstream rows. The remainder of the current section aims to illustrate the additional difficulty for power increase in downstream rows, and speculates on possible future paths for the identification of simplified control strategies as found for the first row.

First, as shown in the top panels of Figure~\ref{fig:cross_section_sinus}, even in the uncontrolled case, the kinetic energy of the flow in the vicinity of the turbine rotor is depleted more and more in the downstream rows. This complicates control strategies for these rows as the opportunity for increased mixing with high energy winds is decreased. Furthermore, intermediate turbines are subjected to increased turbulence levels and more complex vorticity dynamics, as illustrated in Figure~\ref{fig:vorticity_windfarm}. This could explain why the sinusoidal thrust control for the first row did not lead to increase power when applied to the second row: whereas the first row produces increased mixing by destabilizing relatively stable vortex sheets into vortex rings, the second row is already continuously immersed in complex vorticity patterns for which this simple approach does not work. 
However, note that for instance R7 in Figure~\ref{fig:controls} also seems to show quasi-periodic sinusoidal variations in $\cthat$ at a time period of approximately 50~s. This could be an indication that, also for intermediate rows, vortex ring shedding could amount to part of the power increase of observed in the optimal control simulations, albeit at specific moments in time, synchronized with the local flow conditions.

Second, it is important to note that the vortex ring shedding mechanism constitutes only part of the power increase caused by the first row. Figure~\ref{fig:multirow} illustrates that the first-row optimized thrust coefficient also results in a significant power increase in the third row, which is not observed using the sinusoidal thrust strategy. Furthermore, the analysis of the modified control cases in Figure~\ref{fig:scrambled} proves that also the first-row controls are partially synchronized with the flow. This shows that other mechanisms, dependent on specific flow events for increasing wind-farm power, are at play as well. Even though the application of regression algorithms in an attempt to link turbine actions to low-dimensional flow measurements (e.g. local velocity, shear and kinetic energy) has been unsuccessful thus far, similar analysis based upon more complex flow features (e.g. vorticity structures, high-speed turbulent streaks, or downdrafts) might be more promising. However, this requires further optimal control simulations over an extended time, as the total control time horizon of 30 minutes in the current dataset is expected to be insufficient for robust statistics in this kind of analysis. This is an important remaining challenge to be addressed in future research.

%\section{Summary}\label{sec:analysis_summ}
\conclusions \label{sec:analysis_summ}
The current chapter provided a further analysis of the thrust coefficient control characteristics for the C3t5 optimal control case from Chapter~\ref{ch:opt_induction}.

Analysis of the thrust coefficients and numerical experiments have shown a clear distinction between first-row turbines, last-row turbines, and intermediate turbines. Furthermore, observations strongly suggest that the optimization works in a unidirectional way: upstream turbines influence the flow field resulting in favorable conditions for their downstream neighbors, yet information on the possibility of active response and cooperation in the latter has very little influence on upstream control actions.

Qualitative analysis of instantaneous flow fields led to the observation of quasi-periodic shedding of vortex rings from first-row turbines in the optimal control case. This flow feature was succesfully mimicked using simple sinusoidal thrust actuation of the first row. The best parameter set for these sinusoidal variations proved robust to both wind-turbine spacing and turbulence intensity, with a non-dimensional frequency at $St^\star = 0.25$. Interestingly, this frequency corresponds to the peak at $St = 0.2 \dots 0.3$ observed in the first-row thrust coefficient spectra. Since the optimal value $St^\star$ was found using a different realization of the turbulent wind-farm flow, this indicates that the vortex-shedding mechanism is linked to the flow-invariant traits of the first-row optimal controls of case C3t5. Although the first-row sinusoidal control led to a robust increase in total power for a reduced-size $4\times4$ wind farm, a full-scale test indicated that downstream turbine activity is required to obtain increased power at larger farm scales. Further, it was shown that the simple sinusoidal strategy does not lead to increase power extraction when applied to downstream intermediate turbines. Identifying the mechanisms for power increase in these turbines hence remains an important open research question.

\dataavailability{TEXT} %% use this section when having only data sets available

%\appendix
%\section{}    %% Appendix A
%\subsection{}     %% Appendix A1, A2, etc.
%\noappendix       %% use this to mark the end of the appendix section

\competinginterests{The authors declare no conflict of interest.} %% this section is mandatory even if you declare that no competing interests are present

\begin{acknowledgements}
The authors acknowledge funding by the European Research Council (grant no. 306471). The computational resources and services used in this work were provided by the VSC (Flemish Supercomputer Center), funded by the Research Foundation - Flanders (FWO) and the Flemish Government ? department EWI.
\end{acknowledgements}

%% REFERENCES
%\begin{thebibliography}{}
%
%\bibitem[AUTHOR(YEAR)]{LABEL}
%REFERENCE 1
%
%\bibitem[AUTHOR(YEAR)]{LABEL}
%REFERENCE 2
%
%\end{thebibliography}

\bibliographystyle{copernicus}
\bibliography{bib_induction_energies}

%% LITERATURE CITATIONS
%%
%% command                        & example result
%% \citet{jones90}|               & Jones et al. (1990)
%% \citep{jones90}|               & (Jones et al., 1990)
%% \citep{jones90,jones93}|       & (Jones et al., 1990, 1993)
%% \citep[p.~32]{jones90}|        & (Jones et al., 1990, p.~32)
%% \citep[e.g.,][]{jones90}|      & (e.g., Jones et al., 1990)
%% \citep[e.g.,][p.~32]{jones90}| & (e.g., Jones et al., 1990, p.~32)
%% \citeauthor{jones90}|          & Jones et al.
%% \citeyear{jones90}|            & 1990

\end{document}

### 2018-01-12 getfiles.sh
Added figures

@@ -1,0 +1,15 @@
+cp -v ../induction_control_wes/sweep_ct.eps .
+cp -v ../induction_control_wes/spectra_thrust_cthat.eps .
+cp -v ../induction_control_wes/multirow.eps .
+cp -v ../induction_control_wes/single_row_opt.eps .
+cp -v ../induction_control_wes/individual_gains.eps .
+cp -v ../induction_control_wes/fullfarm_singlerow.eps .
+cp -v ../induction_control_wes/scrambled.eps .
+cp -v ../induction_control_wes/vorticity.eps .
+cp -v ../induction_control_wes/setup_sinusoidal.eps .
+cp -v ../induction_control_wes/gains_turbulent_6D_6D_6D_wide_inphase2.eps .
+cp -v ../induction_control_wes/gains_turbulent_6D_6D_6D_wide_antiphase_z0_1m2.eps .
+cp -v ../induction_control_wes/pumpcases2.eps .
+cp -v ../induction_control_wes/gains_row22.eps .
+cp -v ../induction_control_wes/gains_fullscale.eps .
+cp -v ../induction_control_wes/frontviews_allc3t5.eps .

### 2018-01-14 draft.tex
Update chapter -> paper

@@
 
-The analysis presented in the current chapter will lead to the identification of a coherent flow mechanism in the wake of optimally controlled first-row turbines, responsible for part of the power increase in the optimal control simulations.
+The analysis presented in the current paper will lead to the identification of a coherent flow mechanism in the wake of optimally controlled first-row turbines, responsible for part of the power increase in the optimal control simulations.
 This resulted in a new derived simplified dynamic axial induction control strategy for first-row turbines producing a robust power increase in the second row, irrespective of precise local flow details.
@@
 
-The outline of the chapter is as follows: first, Section~\ref{sec:analysis_controls_analysis} presents an analysis of the control and thrust force dynamics, and performs some numerical experiments to elucidate characteristics of the optimal controls.
-As expected from the flow field statistics in previous chapter, it will be shown that the coherent behavior of turbines situated in the first row of the wind farm stands out from their downstream counterparts.
+The outline of the paper is as follows: first, Section~\ref{sec:analysis_controls_analysis} presents an analysis of the control and thrust force dynamics, and performs some numerical experiments to elucidate characteristics of the optimal controls.
+It will be shown that the coherent behavior of turbines situated in the first row of the wind farm stands out from their downstream counterparts.
 Hereafter, Section~\ref{sec:analysis_flow_vis} identifies the shedding of vortex rings from the first row based on a flow visualization.
@@
 turbines which have both upstream and downstream neighbors, for which similar simple control strategies have not been identified thus far.
-In conclusion, Section~\ref{sec:analysis_summ} summarizes the main findings of this chapter.
+In conclusion, Section~\ref{sec:analysis_summ} summarizes the main findings of this paper.
 
@@
 \subsection{Methodology \& numerical setup}
+
+\begin{figure}
+	\includegraphics[width=\textwidth]{figure1.eps}
+	\caption{Schematic overview of wind-farm optimal control methodology from MM17.
+\emph{a)} Control block diagram with adjoint gradient-based optimization and LES flow models illustrating data flow of (optimal) controls $\bs{\varphi}^{(\bullet)}$, system state $\bs{q}$, cost functional $\J$ and its gradient $\nabla \J$.
+\emph{b)} Receding horizon framework subdividing time into discrete flow advancement windows of length $T_A$ with prediction horizon $T$.
+Each arrow represents a forward or adjoint LES.
+Every window consists of an optimization stage (blue and red lines) follow by a flow advancement stage with optimal controls $\bs{\varphi}^{\bullet}$ (green lines).
+\label{fig:optimization_meth}}
+\end{figure}
 
@@
 
+${\J} \nabla\J \bs{\varphi}^{\bullet}$
+
 \subsection{Simulation results}
@@
 
-The current section focuses on the optimal thrust coefficients generated by the optimal control simulations in the previous chapter, and performs numerical experiments to uncover some of the characteristics of these controls and their dependence on the local flow conditions.
+The current section focuses on the optimal thrust coefficients generated by the optimal control simulations in MM17, and performs numerical experiments to uncover some of the characteristics of these controls and their dependence on the local flow conditions.
 
@@
 Although the linear fit $C_P' = a \cthat$ introduced in Appendix~\ref{ch:app_adm} eliminates the error in maximal power extraction, it does not correct the value of the optimal $\cthat$ (note that this could be achieved through a more complex relation between $C_P'$ and $\cthat$).
-Even though this explains the slight increase in mean $\cthat$ for greedily optimized last-row turbines, this does not undermine the overall power gains of the overinductive cases from previous chapter, which all feature power gains well over 10\%.
+Even though this explains the slight increase in mean $\cthat$ for greedily optimized last-row turbines, this does not undermine the overall power gains of the optimal control cases in MM17, which all feature power gains well over 10\%.
 Returning to the more complex thrust coefficients in the other rows it is worth noting that, based on the current dataset, no statistically significant correlations between thrust coefficients of different turbines could be found.
@@
 Note that, even though the spectra for all rows except row 12 collapse at frequencies below $0.05$ Hz, the first-row spectrum shows a small yet significant peak at $f \approx 0.02 \dots 0.03$ Hz ($f D / U_\infty \approx 0.2 \dots 0.3$) as indicated by the purple arrow, which also translates to an increased spectral content of the first row thrust force at these frequencies.
-It will be shown later in this chapter that variations in the thrust coefficient around this frequency can be harnessed to increase power extraction in the second row.
+It will be shown later in this paper that variations in the thrust coefficient around this frequency can be harnessed to increase power extraction in the second row.
 
@@
 
-With this in mind, the remainder of this chapter will focus on the first and most promising link of the control chain: the turbines situated in the first row of the wind farm.
+With this in mind, the remainder of this paper will focus on the first and most promising link of the control chain: the turbines situated in the first row of the wind farm.
 
@@
 The observed shedding of ring vortices seems to occur at specific flow-synchronized times to exploit natural instabilities in the original vortex sheets.
-Therefore, the remainder of this chapter will attempt to accomplish the same effect by simple sinusoidal thrust variations.
+Therefore, the remainder of this paper will attempt to accomplish the same effect by simple sinusoidal thrust variations.
 
@@
 In order to put the technical feasibility of these parameters into context, note that a Strouhal number $St = 0.25$ corresponds here to a sine wave period of $\approx$ 50 s for a turbine with diameter $D = 100$ m and a freestream velocity $U_\infty = 8.5$ m s$^{-1}$.
-Furthermore, although higher than the $\ctmax$ applied in the optimal control cases of the previous chapter, the resulting maximum thrust coefficient of 3.5 could e.g.
+The resulting maximum thrust coefficient of 3.5 could e.g.
 be attained by an NREL 5MW rotor with a 50\% increase in chord length (see \cite{goit2015optimal}, Appendix A).
-Figure~\ref{fig:sinus_baseline}b illustrates normalized power extraction by row for the reference case, the best sinusoidal case, and the first four rows of the optimal control case C3t5 from previous chapter.
+Figure~\ref{fig:sinus_baseline}b illustrates normalized power extraction by row for the reference case, the best sinusoidal case, and the first four rows of the optimal control case C3t5 from MM17.
 The figure shows that the power gain in the sinusoidal cases originates mostly from the second row, and that power in the first row is decreased by approximately 5\%.
@@
 \conclusions \label{sec:analysis_summ}
-The current chapter provided a further analysis of the thrust coefficient control characteristics for the C3t5 optimal control case from Chapter~\ref{ch:opt_induction}.
+The current paper provided a further analysis of the thrust coefficient control characteristics for the C3t5 optimal control case from Chapter~\ref{ch:opt_induction}.
 

### 2018-01-14 draft.tex
Finished draft introduction

@@
 
-\title{Towards practical dynamic induction control of wind farms: analysis of optimal control dynamics and mimicry of first-row behavior}
+\title{Towards practical dynamic induction control of wind farms: analysis of optimal control dynamics and sinusoidal induction control of first-row turbines}
 \runningtitle{Towards practical dynamic induction control of wind farms}
@@
 The current control paradigm in such farms optimizes performance at the wind-turbine level and does not account for these interactions, resulting in sub-optimal wind-farm efficiency.
-In contrast, control strategies at the farm level allow to influence wake interactions and promise to improve overall wind-farm performance: by redirecting propagating wakes (yaw control, see e.g.
-Refs YYY) or by affecting the induced wake velocity deficits (axial induction control, see e.g.
-Refs XXX) wind conditions for downstream turbines can be improved significantly.
-In recent work, Munters \& Meyers \cite{munters2017optimal} ...
+In contrast, control strategies at the farm level allow to influence wake interactions and promise to improve overall wind-farm performance by improving wind conditions for downstream turbines, i.e.
+by redirecting propagating wakes (yaw control, see e.g.
+\citealp{fleming2014evaluating, gebraad2016wind, campagnolo2016wind}) or by affecting the induced wake velocity deficits (axial induction control, see e.g.
+\citealp{nilsson2015large, annoni2016analysis, bartl2016experimental}).
+A more exhaustive survey of wind-farm control in a broader context can be found in \cite{knudsen2015survey} and \cite{boersma2017tutorial}.
 
-Korte beschrijving van methodologie, sleutelresultaat, problemen, zie abstract...
+In contrast to the studies cited above, which focus on static setpoint control of wind farms, \cite{goit2015optimal} introduced a dynamic induction control approach based on large-eddy simulations and adjoint gradient optimization.
+The methodology was applied to the asymptotic case of a fully-developed `infinite' aligned wind farm, and power gains of about 16\% were quantified.
+Later, this approach was also used in induction control studies of wind farms with entrance effects in \cite{goit2016optimal} and, more recently, in \cite{munters2017optimal}, where similar gains in the order of 15\%--20\% were achieved.
+It is important to note that the computational cost of this dynamic induction control methodology is orders of magnitude too high for direct implementation, and the complex optimal control signals generated by the simulations are highly tuned to specific flow events and cannot be translated to practical flow conditions.
+However, the methodology allows to assess the theoretical potential for wind-farm control, and an increased understanding of the flow physics of controlled wind farms could lead to simplified control strategies that could be applied in practice.
 
-The current paper presents efforts on understanding optimal control dynamics and identifying the flow mechanisms behind the power increase observed in the optimal induction control simulations discussed in Chapter~\ref{ch:opt_induction}, with the aim of harnessing them in tractable dynamic induction wind-farm controllers.
-Throughout the analysis, the focus will lie on the overinductive fast-response case C3t5, as it was observed to exhibit similar flow statistics as the maximum yield case C3t0, but with smoother thrust force dynamics.
+Very recently, the methodology of \cite{goit2015optimal} was generalized to include dynamic yaw control in \cite{munters2018dynamic}.
+In this study, induction control and yaw control were compared for a relatively small aligned wind farm, and yaw control was found to yield higher power gains for this setup.
+Furthermore, the high potential of combined induction and yaw control was quantified, and analysis of the yaw control signals allowed to identify practical simplified dynamic yaw control strategies.
+The search for similar practical control strategies for yaw control has remained unsuccessful to date.
 
-The analysis presented in the current paper will lead to the identification of a coherent flow mechanism in the wake of optimally controlled first-row turbines, responsible for part of the power increase in the optimal control simulations.
+The current paper presents efforts on understanding optimal control dynamics and identifying the flow mechanisms behind the power increase observed in the optimal induction control simulations in \cite{munters2017optimal} (further denoted as MM17), with the aim of harnessing them in tractable dynamic induction wind-farm controllers.
+The analysis presented here will lead to the identification of a coherent flow mechanism in the wake of optimally controlled first-row turbines, responsible for part of the power increase in the optimal control simulations.
 This resulted in a new derived simplified dynamic axial induction control strategy for first-row turbines producing a robust power increase in the second row, irrespective of precise local flow details.
@@
 
-The outline of the paper is as follows: first, Section~\ref{sec:analysis_controls_analysis} presents an analysis of the control and thrust force dynamics, and performs some numerical experiments to elucidate characteristics of the optimal controls.
+The outline of the paper is as follows: first, Section~\ref{sec:optimal_control_sims} discusses the numerical setup and optimal control simulations of MM17 that will be further analysed in the current paper.
+Section~\ref{sec:analysis_controls_analysis} presents an analysis of the control and thrust force dynamics, and performs some numerical experiments to elucidate characteristics of the optimal controls.
 It will be shown that the coherent behavior of turbines situated in the first row of the wind farm stands out from their downstream counterparts.
@@
 \end{figure}
+
+Throughout the analysis, the focus will lie on the overinductive fast-response case C3t5, as it was observed to exhibit similar flow statistics as the maximum yield case C3t0, but with smoother thrust force dynamics.
 
@@
 306471).
-The computational resources and services used in this work were provided by the VSC (Flemish Supercomputer Center), funded by the Research Foundation - Flanders (FWO) and the Flemish Government ?
-department EWI.
+The computational resources and services used in this work were provided by the VSC (Flemish Supercomputer Center), funded by the Research Foundation - Flanders (FWO) and the Flemish Government, department EWI.
 \end{acknowledgements}

### 2018-02-05 draft.tex
First draft sent to JM

@@
 
+\newcommand{\red}[1]{{\color{red} #1}}
+
 \begin{document}
 
-\title{Towards practical dynamic induction control of wind farms: analysis of optimal control dynamics and sinusoidal induction control of first-row turbines}
+\title{Towards practical dynamic induction control of wind farms: analysis of optimally controlled wind-farm boundary layers and sinusoidal induction control of first-row turbines}
 \runningtitle{Towards practical dynamic induction control of wind farms}
@@
 The analysis shows that wind-farm controls are optimized in a uni-directional manner with little upstream propagation of information, and that turbines can be classified into first-row, intermediate-row, and last-row turbines based on their optimal control dynamics.
-Furthermore, it is found that optimally-controlled first-row turbines periodically shed vortex rings leading to enhanced wake mixing.
-This behavior is mimicked with a simple sinusoidal thrust control strategy, resulting in robust power gains for turbines in the entrance region of the farm.
+Although the control mechanisms for intermediate-row turbines remain unclear based on the current analysis, for first-row turbines it is found that the optimal controls increase wake mixing through periodically shed vortex rings.
+This behavior is mimicked with a simple sinusoidal thrust control strategy for first-row turbines, resulting in robust power gains for turbines in the entrance region of the farm.
 \end{abstract}
@@
 
-In contrast to the studies cited above, which focus on static setpoint control of wind farms, \cite{goit2015optimal} introduced a dynamic induction control approach based on large-eddy simulations and adjoint gradient optimization.
+In contrast to the studies cited above, which focus on static setpoint control of wind farms, \cite{goit2015optimal} introduced a dynamic induction control approach based on large-eddy simulations (LES) and adjoint gradient optimization.
+In this way, individual turbines could be used as dynamic flow actuators that influence the wind-farm boundary layer flow in such a way as to optimize aggregate wind-farm power extraction.
 The methodology was applied to the asymptotic case of a fully-developed `infinite' aligned wind farm, and power gains of about 16\% were quantified.
@@
 
-Very recently, the methodology of \cite{goit2015optimal} was generalized to include dynamic yaw control in \cite{munters2018dynamic}.
+Recently, the methodology of \cite{goit2015optimal} was generalized to include dynamic yaw control in \cite{munters2018dynamic}.
 In this study, induction control and yaw control were compared for a relatively small aligned wind farm, and yaw control was found to yield higher power gains for this setup.
@@
 
-\section{Description of optimal control simulations}\label{sec:optimal_control_sims}
-The current section describes the optimal control simulations performed by Munters \& Meyers \cite{munters2017optimal}, of which the results are further analyzed in the current paper.
-First, the methodology and numerical setup is detailed.
-Afterwards, the simulation results are discussed.
+\section{Description of optimal control simulations in MM17}\label{sec:optimal_control_sims}
+The current section describes the optimal control simulations performed by MM17, the results of which are further analyzed in the current paper.
+First, the methodology is illustrated.
+Then, the numerical setup is detailed.
+Afterwards, the optimization results on power extraction and time-averaged flow field features are discussed.
 
-\subsection{Methodology \& numerical setup}
+\subsection{Control methodology}
+
+A schematic overview of the wind-farm control methodology is shown in Figure \ref{fig:optimization_meth}.
+Figure \ref{fig:optimization_meth}a illustrates the control block diagram: an iterative optimization loop updates the wind-farm control vector $\bs{\varphi}$ until a set of optimized controls $\bs{\varphi}^{\bullet}$ is found.
+This optimization is based upon an unsteady turbulence-resolving LES wind-farm flow model, and sensitivities of the cost functional $\J$ (i.e.
+the total wind-farm power extraction) are calculated using an adjoint formulation of this flow model.
+In this way, no a priori simplification on the model accuracy are made, and the control signals can be designed in a such a way that turbines can actively tap into the dynamic of the turbulent wind-farm boundary layer.
+The optimization is performed using a receding-horizon control framework, as illustrated in Figure \ref{fig:optimization_meth}b.
+In this framework, wind-farm control is optimized for a finite time horizon $T$, involving a sequential set of LES and adjoint LES simulations.
+Upon convergence of the optimization, optimized control signals are applied in a flow advancement simulation for a time $T_A < T$, after which a new optimization window is initiated.
 
@@
 
-Throughout the analysis, the focus will lie on the overinductive fast-response case C3t5, as it was observed to exhibit similar flow statistics as the maximum yield case C3t0, but with smoother thrust force dynamics.
-
-Optimal control based on LES and optimization
-
-Receding-horizon framework (maak 1 gecombineerde figuur)
-
-Optimal control problem
-
+Within each optimization window, total wind-farm power is optimized by solving the following partial-differential-equation-constrained optimization problem: 
 \begin{alignat}{4}
@@
 &                                                   &        \tau \ddt{\ctihat}&=\cti - \ctihat 								& i=1\dots N_t~\text{in } (0,T],  \label{eq:ctihat_constraint}\\
-&                                                   &        C_{T,\text{min}}' \leq~ &\cti \leq C_{T,\text{max}}'				& i=1\dots N_t~\text{in } (0,T],  \label{eq:boxct_constraint}
+&                                                   &       0 \leq~ &\cti \leq C_{T,\text{max}}'				& i=1\dots N_t~\text{in } (0,T],  \label{eq:boxct_constraint}
 \end{alignat}
-Define turbine forces and powers!
+The cost functional that is minimized in \eqref{eq:costfunction} is the total wind-farm energy extraction over time horizon $T$, with switched sign for convention of notation.
+The control variables are the time-dependent thrust coefficient setpoints $C_{T,i}'$ of every turbine $i$ ($=1 \dots N_t$), i.e.
+$\bs{\varphi} = [C_{T,1}'(t), \dots, C_{T,N_t}'(t) ]$, and the state variables are denoted as  $\bs{q}=[\utilde(\bs{x},t);~ \ptilde(\bs{x},t);~ \widehat{C}_{T,1}'(t), \dots, \widehat{C}_{T,N_t}'(t)]$, with $\utilde$ the filtered velocity, $\ptilde$ the filtered pressure, and $\widehat{C}_{T,i}'$ the (time-filtered) thrust coefficient for turbine $i$ (see below).
 
-Focus on C3t5 case.
+The filtered Navier--Stokes momentum and continuity state equations in \eqref{eq:NSmomentum_constraint} - \eqref{eq:NScontinuity_constraint} are solved using an in-house LES solver (see, e.g.
+\citealp{calaf2010large, meyers2010large, goit2016optimal} for a detailed discussion of the solver).
+The time-filtering state equation in \eqref{eq:ctihat_constraint} applies a one-sided exponential time filter to the thrust coefficient setpoints $C_{T,i}'$ with a characteristic wind-turbine response timescale $\tau$ \citep{munters2016effect}.
+Finally, the box constraints in \eqref{eq:boxct_constraint} limit the variations in the turbine thrust coefficients to technically feasible values.
 
-${\J} \nabla\J \bs{\varphi}^{\bullet}$
+The forces exerted by turbine $i$ on the boundary-layer flow are parametrized using a standard non-rotating actuator disk model as
+%\begin{equation}
+%$	\bs{f}_i(\bs{x},t) = - \frac{1}{2}~ \widehat{C}_{T,i}'(t)~ V_i(t)^2~ \R_i (\bs{x})~ \eperpi,$
+$	\bs{f}_i(\bs{x},t) = - (1/2) \widehat{C}_{T,i}'(t) V_i(t)^2 \R_i (\bs{x}) \eperpi,$
+%\end{equation}
+where $\R_i$ is a smoothed representation of the geometric footprint of the turbine on the LES grid and $\eperpi$ is the rotor-perpendicular vector.
+Further, the disk-averaged velocity is defined as $V_i = (1/A_i) \sint \R_i(\bs{x}) \utilde \cdot \eperpi ~\dx$, with $A_i$ the rotor disk area.
+Mechanical power captured by the wind turbine is calculated as $P_i = (1/2) C_{P,i}'(t) V_i(t)^3 A_i$, with $C_{P,i}' = 0.9 C_{T,i}'$, resulting from a fit of LES results to momentum theory, partially eliminating the overprediction of wind-turbine power on typical wind-farm LES grid resolutions \citep{munters2017optimal}.
+
+\subsection{Case setup}
+The wind farm considered in MM17 has an aligned pattern of 12 rows by 6 columns, with 6 rotor diameters spacing in axial and tranversal directions.
+The flow is simulated on a domain of $10 \times 3.6 \times 1$ km$^3$, discretized on a simulation grid of $384 \times 192 \times 144$ grid points.
+A snapshot of the streamwise velocity field is shown in Figure \ref{fig:flowfield_MM17}.
+The wind farm was controlled over a total of $N_A = 15$ time windows with a prediction horizon $T = 240 s$ and a flow advancement time of $T_A = T/2 = 120 s$, resulting in a total control time $T_{\rm tot} = N_A T_A = 30$ minutes.
+
+\begin{figure}
+	\includegraphics[width=0.8\textwidth]{flowfield_MM17}
+	\caption{Instantaneous streamwise velocity $\widetilde{u}_x$ for the $12 \times 6$ aligned wind farm considered in MM17.
+Black lines indicate wind-turbine locations.
+The black dashed line near the end of the domain indicates a buffer region used for imposition of inflow boundary conditions.
+Figure originally published in \cite{munters2017optimal} under a~CC-BY 4.0 license.\label{fig:flowfield_MM17}}
+\end{figure}
+
+A conventionally (greedily) controlled wind farm with steady $C_T' = 2$ (corresponding to the Betz limit in momentum theory) was defined as a reference case.
+Several different optimal control cases were defined, based on the wind-turbine response time $\tau = $0, 5, or 30 s (instantaneous, fast, or slow response, see Equation \ref{eq:ctihat_constraint}) and the maximal thrust coefficient $\ctmax = 2$ or 3 (underinductive or overinductive, see Equation \ref{eq:boxct_constraint}).
+Cases are denoted as C<X>t<Y>, where X and Y represent $\ctmax$ and $\tau$ respectively, e.g.
+C3t30  for the case with $\ctmax = 3$ and $\tau = 30$ s.
 
@@
 
+Figure \ref{fig:power_MM17} illustrates the energy extraction of the optimally controlled wind farm cases, normalized by the greedy reference control case.
+Figure \ref{fig:power_MM17}a shows that the adjoint LES-based control approaches achieves energy gains ranging from a minor 2\% in the most restrictive C2t30 case to over 20\% in the C3t0 case.
+From Figure \ref{fig:power_MM17}b it can be seen that, for all cases except C3t30, power is curtailed in the first row to a limited degree, whereas the downstream rows compensate for this loss by extracting significantly more energy.
+Furthermore, not taking into account the first row, the last row achieves the highest energy extraction in every case, as it can act greedily without compromising power extraction in downstream neighbors.
+
+\begin{figure}
+	\includegraphics[width=\textwidth]{power_MM17}
+	\caption{Energy extraction $E_O$ of optimally controlled wind-farm cases from MM17, normalized by a greedy reference case $E_R$.
+\emph{a)} Total energy extraction.
+Error bars indicate confidence intervals of $\pm$ 2 standard deviations.
+\emph{b)} Energy extraction by row, normalized by first-row reference power.
+C3t0, red line; C3t5, yellow line; C3t30, blue line; C2t0, red dashed line; C2t5, yellow dashed line; C2t30, blue dashed line.
+Figure originally published in \cite{munters2017optimal} under a~CC-BY 4.0 license.\label{fig:power_MM17}}
+\end{figure}
+
+In the remainder of this section, time-averaged wind-farm flow properties will be investigated.
+Here and throughout the remainder of this paper, the focus will lie on case C3t5, as it produces similar energy gains as the highest-yield case C3t0 (see Figure \ref{fig:power_MM17}a), but achieves this with smooth thrust coefficient signals.
+In the following discussion, the time averaging operation is denoted by an overline, and flow field variables are decomposed into mean and fluctuating components as $u = \overline{u} + u' \equiv U + u'  $.
+Figure \ref{fig:flowfield_statistics} illustrates time-averaged flow field quantities of the reference case (left panels, a1 -- g1) and the differences between the optimized C3t5 case and the reference case (right panels, a2 -- g2).
+Simulation results are averaged over the different columns and are shown as either topviews at hub height (Figures \ref{fig:flowfield_statistics}b,f,g) or sideviews through a turbine column (Figures \ref{fig:flowfield_statistics}a,c,d,e).
+
+Figures \ref{fig:flowfield_statistics}a and \ref{fig:flowfield_statistics}b illustrate sideviews and topviews of the axial velocity throughout the wind farm.
+It can be seen that downstream turbines are consistently subjected to higher incoming velocities, which explains the increased energy extraction discussed above.
+Furthermore, a larger drop in streamwise velocity over the turbine disk can be observed, most notably in the first-row turbines, indicating deeper wakes with enhanced recovery before hitting the next row of turbines.
+Furthermore, it can be observed that the axial velocity in the flow above and besides the wind-turbine column is reduced, indicating that the mean-flow kinetic energy is depleted in these regions, to the benefit of the flow passing through the wind turbines.
+
+Figures \ref{fig:flowfield_statistics}c shows sideviews of turbulence kinetic energy.
+The figure shows an increase in turbulence throughout the entire wind farm, spreading to the internal boundary layer above the turbines.
+Note specifically the sharp increase in turbulence in the core wake region behind the first-row turbine, for which an enhanced recovery was found as discussed above.
+
+Figure~\ref{fig:flowfield_statistics}d and \ref{fig:flowfield_statistics}e show sideviews through the rotor centerline of top-down turbulence and mean-flow transport of axial momemtum, i.e.
+$-\overline{\widetilde{u}_x' \widetilde{u}_z'}$ and $- \widetilde{U}_x \widetilde{U}_z$ respectively.
+It can be seen that, although mean-flow vertical transport is virtually unaffected, turbulence top-down transport of axial momentum is increased significantly in the upper part of the wakes, indicating increased turbulent mixing with the internal boundary layer above the wind-farm canopy.
+The effect is somewhat more pronounced in the wake behind the first row of turbines, for which also a slight increase in upwards transport of momentum can be observed in the lower part of the wake.
+
+Figure~\ref{fig:flowfield_statistics}f and \ref{fig:flowfield_statistics}g show planviews at hub height of transversal turbulence and mean-flow transport of axial momemtum, i.e.
+$\overline{\widetilde{u}_x' \widetilde{u}_y'}$ and $\widetilde{U}_x \widetilde{U}_y$ respectively.
+The sign convention is such that positive values correspond to transport in the positive $y$ direction in the figure.
+A slight increase in turbulent transversal transport towards the wake centerline can be observed behind every row.
+On the other hand, mean-flow transversal momentum transport into the wake region is increased significantly behind the first two turbine rows only.
+Downstream of these rows, the difference between these cases and the reference case is far less coherent.
+The latter can be explained by the fact that it occurs in the region in which $\widetilde{U}_x$ in the intercolumn channels starts to deviate significantly from the reference case as shown in Figure~\ref{fig:flowfield_statistics}b.
+
+The analysis of flow features given above indicate that the optimal controls in case C3t5 coherently influence the wind-farm flow field in such a way as to provide better flow conditions for downstream turbines.
+Increased axial velocities are observed for all downstream turbines, and enhanced momentum transport towards the turbine region is achieved.
+Furthermore, many of the observed flow features are most salient for the first row turbines.
+In the following section, the optimized thrust coefficients themselves will be investigated.
+It will be shown that, also from a controls perspective, first-row turbines stand out from their downstream counterparts.
+
+\begin{figure}
+	\includegraphics[width=\textwidth]{statfig_initial.eps}
+	\caption{Time-averaged flow field quantities of simulation results from MM17.
+\emph{Left (1):} Time averages of reference simulations with $C_T' = 2$ for all turbines.
+\emph{Right (2):} Difference $\Delta$ between reference and optimal control (C3t5) simulation, defined as $\Delta = X_{\rm C3t5} - X_{\rm REF}$ for any variable $X$.
+\emph{a):} Axial velocity $U_x$ (planview at hub height).
+\emph{b):} Axial velocity $U_x$ (sideview through turbine column).
+\emph{c):} Turbulence kinetic energy $k = (1/2)(\overline{\widetilde{u}'_x \widetilde{u}'_x} + \overline{\widetilde{u}'_y \widetilde{u}'_y} + \overline{\widetilde{u}'_z \widetilde{u}'_z})$ (sideview).
+\emph{c):} Turbulence top-down transport $ - \overline{\widetilde{u}'_x\widetilde{u}'_z}$.
+\emph{c):} Mean-flow top-down transport $ - \widetilde{U}_x \widetilde{U}_z$ (sideview).
+\emph{c):} Turbulence horizontal transport $\overline{\widetilde{u}'_x \widetilde{u}'_y}$ (planview).
+\emph{c):} Mean-flow horizontal transport $\widetilde{U}_x \widetilde{U}_y$ (planview).
+Black lines indicate wind turbine locations.
+Simulation results are averaged over the six different wind turbine columns.
+\label{fig:flowfield_statistics}}
+\end{figure}
+
 \section{Thrust coefficient analysis and numerical experiments}\label{sec:analysis_controls_analysis}
 
-The current section focuses on the optimal thrust coefficients generated by the optimal control simulations in MM17, and performs numerical experiments to uncover some of the characteristics of these controls and their dependence on the local flow conditions.
+The current section focuses on the optimal thrust coefficients generated by the optimal control simulations in MM17, and performs numerical experiments to uncover some of the characteristics of these control signals.
+The aim of this analysis is to increase our overall understanding of how power is increased in optimally controlled wind-farm boundary layers, i.e.
+how the potential for power gains depends on the location of a turbine within the farm, whether optimal controls are fully tuned to specific flow events or contain flow-invariant features, and whether the control signals in neighboring turbines are correlated or can be used independently.
+This increased understanding will be used as a stepping stone towards the design of practical control strategies that do not require the computationally prohibitive LES-based methodology of MM17.
 
-First, the thrust coefficient signals themselves are analyzed in Section~\ref{sec:analysis_controls_analysis}.
+First, the thrust coefficient signals themselves are analyzed in Section~\ref{sec:analysis_thrust}.
 Second, the optimized thrust coefficients are applied only to subsets of turbine rows in Section~\ref{sec:application_subset}.
@@
 Third, additional optimal control simulations, in which only one single active row is optimized, are discussed in Section~\ref{sec:optimization_single}.
-These optimizations provide an indication on how increased power potential is distributed among the rows, and allows to compare the resulting single-row optimized thrust coefficients with the fully cooperative coefficients from case C3t5.
+These optimizations provide an indication on how increased power potential is distributed among the rows, and allows to compare the resulting single-row optimized thrust coefficients with the fully cooperative coefficients from case MM17.
 Fourth, Section~\ref{sec:modified_thrust} evaluates the dependency of optimized thrust coefficients on specific flow conditions.
@@
 Although momentum theory predicts maximal power extraction for steady uniform inflow at $\cthat = 2$, the actual optimal steady value for the ADM at the current spatial resolution lies somewhat higher at $\cthat \approx 2.4$, for which power extraction is about 1.4\% higher than at $\cthat = 2$.
-This behavior is related to the overprediction in ADM power due to the diffuse turbine representation on typical simulation grids discussed in Section~\ref{sec:meth_adm}: the mass flow through the rotor disk at $\cthat > 2$ is slightly too high compared to momentum theory, resulting in a  shift of optimal $\cthat$ towards somewhat higher values.
-Although the linear fit $C_P' = a \cthat$ introduced in Appendix~\ref{ch:app_adm} eliminates the error in maximal power extraction, it does not correct the value of the optimal $\cthat$ (note that this could be achieved through a more complex relation between $C_P'$ and $\cthat$).
-Even though this explains the slight increase in mean $\cthat$ for greedily optimized last-row turbines, this does not undermine the overall power gains of the optimal control cases in MM17, which all feature power gains well over 10\%.
+This behavior is related to the overprediction in ADM power due to the diffuse turbine representation on typical simulation grids: the mass flow through the rotor disk at $\cthat > 2$ is slightly too high compared to momentum theory, resulting in a  shift of optimal $\cthat$ towards somewhat higher values.
+Although the linear fit $C_P' = a \cthat$ introduced in \cite{munters2017optimal}, Appendix~A eliminates the error in maximal power extraction, it does not correct the value of the optimal $\cthat$ (note that this could be achieved through a more complex relation between $C_P'$ and $\cthat$).
+Even though this explains the slight increase in mean $\cthat$ for greedily optimized last-row turbines, this does not undermine the overall power gains of the overinductive optimal control cases in MM17, which all feature increases well over 10\%.
 Returning to the more complex thrust coefficients in the other rows it is worth noting that, based on the current dataset, no statistically significant correlations between thrust coefficients of different turbines could be found.
@@
 \begin{figure}
-	\includegraphics[width=\textwidth]{controls.eps}
+	\includegraphics[width=0.8\textwidth]{controls.eps}
 	\caption{Time evolution of the thrust coefficient $\cthat$ for a selection of optimally controlled turbines  of case C3t5.
@@
 \begin{figure}
-	\includegraphics[width=\textwidth]{sweep_ct.eps}
+	\includegraphics[width=0.8\textwidth]{sweep_ct.eps}
 	\caption{Normalized power extraction as a function of steady thrust coefficient $\cthat$ for wind turbines subject to the same freestream turbulent inflow as in case C3t5.
@@
 \begin{figure}
-	\includegraphics[width=\textwidth]{spectra_thrust_cthat.eps}
+	\includegraphics[width=0.8\textwidth]{spectra_thrust_cthat.eps}
 	\caption{Power spectral density (PSD) estimates of the row-averaged thrust force $f_T$ (\emph{a}) and thrust coefficient $\cthat$ (\emph{b}) as a function of frequency $f$ (bottom axis) and non-dimensional Strouhal number $St$ (top axis).
@@
 The high-frequency slopes of around $-5$ observed both for $f_T$ and $\cthat$ indicate that force variability on short timescales is caused mainly by thrust coefficient variations, whereas the slower thrust force dynamics tend more to a $-5/3$ slope, suggesting that these are governed by the unsteadiness in the turbulence instead.
-Note that, even though the spectra for all rows except row 12 collapse at frequencies below $0.05$ Hz, the first-row spectrum shows a small yet significant peak at $f \approx 0.02 \dots 0.03$ Hz ($f D / U_\infty \approx 0.2 \dots 0.3$) as indicated by the purple arrow, which also translates to an increased spectral content of the first row thrust force at these frequencies.
-It will be shown later in this paper that variations in the thrust coefficient around this frequency can be harnessed to increase power extraction in the second row.
+Note that, even though the spectra for all rows except row 12 collapse at frequencies below $0.05$ Hz, the first-row spectrum shows a small peak at $f \approx 0.02 \dots 0.03$ Hz ($f D / U_\infty \approx 0.2 \dots 0.3$) as indicated by the purple arrow, which also translates to an increased spectral content of the first row thrust force at these frequencies.
+It will be shown later in this paper that variations in the thrust coefficient around this frequency are directly related to increased power extraction.
 
@@
 This suggests that self-optimization is very limited: the optimal controls for a given turbine are designed to create favorable flow conditions in the downstream rows instead of increasing local power.
-Furthermore, although the discussion in the previous paragraph has shown that downstream controls are optimized with the upstream actions in mind, the converse is not true: upstream control actions do not require a specific downstream response in order to increase power.
+Furthermore, although the discussion in the previous paragraph has shown that downstream controls are optimized with the upstream actions in mind, the converse is not true: upstream control actions do not require a specific downstream response in order to increase power in downstream rows.
 
@@
 Figure~\ref{fig:single_row_opt} illustrates the relative increase in wind-farm power extraction for each of the twelve individually optimized control cases.
-The optimization is run until the continuous adjoint gradient accuracy prevents further progress in the vicinity of a local minimum (see discussion accompanying Figure~\ref{fig:window_opt}).
-Upon interpreting the actual values from the figure, it is important to note that the reported power gain covers the full optimization horizon $T$, and is hence affected by finite-horizon effects.
+The optimization is run until the continuous adjoint gradient accuracy prevents further progress in the vicinity of a local minimum.
+\red{Upon interpreting the actual values from the figure, it is important to note that the reported power gain covers the full optimization horizon $T$, and is hence affected by finite-horizon effects.
 Furthermore, the first window of an optimal control simulation, as considered here, contains an initial dead zone corresponding to the wake advection lag before upstream turbines start influencing their downstream neighbors.
@@
 Nevertheless, the relative order of the different cases still provides information that can be generalized to full optimal control studies with multiple windows and longer time horizons.
+(WM: dit eventueel weglaten?)}
 
@@
 This is not surprising as R1 produces the most power of all rows, and typically leaves the deepest wakes, causing second-row turbines to perform  poorly in aligned wind-farm layouts (see, e.g., \citealp{porte2013numerical,nilsson2015large}; and \citealp{stevens2016effects}).
-At the other end of the spectrum, the last row (R12) is the least fruitful: it can only increase power by increasing $\cthat$ to harness the imperfections in the ADM for the given spatial resolution illustrated in Figure~\ref{fig:ct_sweep} (see discussion in Section~\ref{sec:analysis_thrust}), or by exploiting finite-horizon effects in which it attempts to completely halt the flow near the end of the time horizon.
+At the other end of the spectrum, the last row (R12) is the least fruitful.
 The intermediate rows (R2--R11) lie closer together, with the general trend being that potential is somewhat decreased with downstream distance into the wind farm, although this decrease is not monotonous.
@@
 	\centering
-	\includegraphics[width=\textwidth]{single_row_opt.eps}
+	\includegraphics[width=0.8\textwidth]{single_row_opt.eps}
 	\caption{Increase in wind-farm power extraction for the first (and only) timewindow of the optimal control cases in which only a single row is optimized. \label{fig:single_row_opt}}
@@
 R$_{\rm i+1}$), and that the influence on row R$_{\rm i+3}$ is limited.
-This is explained by the fact that the finite optimization horizon $T = 240$ s allows for more interactions with directly neighboring turbines than with those located further downstream.
+This is explained by the fact that the finite optimization horizon used in MM17 (i.e.
+$T = 240$ s) allows for more interactions with directly neighboring turbines than with those located further downstream.
 Furthermore, except for the optimal control case of the last row (R12), self-optimization is virtually non-existent: power gains are achieved by modifying the flow to yield more favorable conditions in the downstream rows.
@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{individual_gains.eps}
+	\includegraphics[width=0.75\textwidth]{individual_gains.eps}
 	\caption{Relative power increase matrix for downstream rows in the single-row optimization case for row R$_{\rm i}$, indicated in the horizontal axis.
@@
 	\centering
-	\includegraphics[width=0.9\textwidth]{fullfarm_singlerow.eps}
+	\includegraphics[width=0.8\textwidth]{fullfarm_singlerow.eps}
 	\caption{Optimized thrust coefficients in several rows for case C3t5 (black) and single-row optimal control cases (gray) for the full optimization horizon $T = 240$ s.\label{fig:control_C3t5_singlerow}}
@@
 Firstly, both Figure~\ref{fig:multirow}b (full-farm optimization) and Figure~\ref{fig:gains_singlerow} (single-row optimization) suggest that self-optimization of turbines is negligible, and that power gains are mainly achieved by generating more favorable flow conditions for downstream turbines.
-Furthermore, the remarkable match between optimized thrust coefficients for full-farm and single-row optimal control cases from Figure~\ref{fig:control_C3t5_singlerow} indicates that, at least for the first row, upstream controls do not rely on active downstream repose Instead, Figure~\ref{fig:multirow}b shows that, for any row $i$, the full potential in power increase is virtually attained by applying controls only for the upstream turbines up until row $i-1$.
-Furthermore, Figure~\ref{fig:multirow}a shows that optimized thrust coefficients in downstream rows by themselves do not result in power increase without also applying optimized coefficients in their upstream neighbors.
+Furthermore, the remarkable match between optimized thrust coefficients for full-farm and single-row optimal control cases from Figure~\ref{fig:control_C3t5_singlerow} indicates that, at least for the first row, upstream controls do not rely on active downstream response.
+Instead, Figure~\ref{fig:multirow}b shows that, for any row $i$, the full potential in power increase is virtually attained by applying controls only for the upstream turbines up until row $i-1$.
+Furthermore, Figure~\ref{fig:multirow}a illustrates that optimized thrust coefficients in downstream rows by themselves do not result in power increase without also applying optimized coefficients in their upstream neighbors.
 This indicates that controls are tuned to specific flow events.
@@
 
-With this in mind, the remainder of this paper will focus on the first and most promising link of the control chain: the turbines situated in the first row of the wind farm.
+With this in mind, the following section of this paper will focus on the first and most promising link of the control chain: the turbines situated in the first row of the wind farm.
 
-\section{Flow visualization and mimicry of first-row behavior}\label{sec:analysis_flow_vis}
+\section{First-row turbine behavior}\label{sec:analysis_flow_vis}
 As shown in the previous section, the behavior of turbines within the first upstream row stands out from the others.
@@
 First, a qualitative analysis of the instantaneous flow field is performed in Section~\ref{sec:opt_visualization}, resulting in the observation of vortex rings being shed from first-row turbines.
-Thereafter, this mechanisms is mimicked by imposing sinusoidally varying thrust coefficients in these turbines in Section~\ref{sec:opt_sinus}, with the aim of increasing power through similar mechanisms as in the computationally expensive optimal control cases.
+Thereafter, this mechanism is mimicked by imposing sinusoidally varying thrust coefficients in these turbines in Section~\ref{sec:opt_sinus}, with the aim of increasing power through similar mechanisms as in the computationally expensive optimal control cases.
 
@@
 Figures~\ref{fig:vorticity_windfarm}a,b show isosurfaces of vorticity magnitude, colored by streamwise velocity $\widetilde{u}_x$.
-Further, volumes of very low streamwise velocity (at $\widetilde{u}_x < 4.5 $ m s$^{-1}$) are rendered in black.
 Figure~\ref{fig:vorticity_windfarm}a shows that, in the reference case, the first-row turbines shed relatively stable vortex sheets that demarcate the wake from the freestream flow.
@@
 In contrast, the optimized case shows coherent vortex rings being shed from the first-row turbine.
-As indicated in Figure~\ref{fig:vorticity_windfarm}b, the locations of the rings in the controlled case coincide with naturally occurring bulges in the vortex sheet of the reference case: the controlled turbines further destabilize the sheet by well-timed temporal variations in its thrust coefficient.
-As illustrated in Figure~\ref{fig:vorticity_windfarm}c, this results in smaller velocity deficits in the wake region.
+As indicated by the black arrows in Figure~\ref{fig:vorticity_windfarm}b, the locations of the rings in the controlled case coincide with naturally occurring bulges in the vortex sheet of the reference case: the controlled turbines further destabilize the sheet by well-timed temporal variations in its thrust coefficient.
+Figure~\ref{fig:vorticity_windfarm}c shows that this results in smaller velocity deficits in the wake region.
 Note that, downstream of the second turbine, the vorticity field becomes much more complex and differences in the flow fields are less coherent.
@@
 Deep wake regions (with $\widetilde{u}_x < 4.5$ m s$^{-1}$) are rendered in black.
+Black arrows indicate the naturally occuring unstable bulges in the reference case, and accompanying vortex rings in the optimized case.
 \emph{c)} Contours of streamwise velocity $\widetilde{u}_x$.
@@
 The aim of the current section is to mimic the quasi-periodic shedding of vortex rings by upstream turbines as observed above through the use of simple periodic variations in the thrust coefficient.
-To this end, the Betz-optimal coefficient $\cthat = 2$ is perturbed by a sinusoidal signal, parametrized by its amplitude $A$, and a non-dimensional frequency in the form of the Strouhal number $St = f D/ U_\infty$:
+Instead of optimizing a high-dimensional control signal that can evolve freely in time as in MM17, we impose a sinusoidal perturbation on the Betz-optimal coefficient $\cthat = 2$, parametrized by its amplitude $A$, and its frequency in the form of a non-dimensional Strouhal number $St = f D/ U_\infty$, with $f$ the dimensional frequency, $D$ the turbine diameter, and $U_\infty$ the unperturbed time-averaged upstream velocity:
 \begin{equation}
@@
 \subsubsection{Parameter sweep}
-A parameter sweep is performed for both $A$ and $St$ on a reduced-size wind-farm LES, as illustrated in Figure~\ref{fig:sinus_setup}.
-The farm consists of $4 \times 4$ turbines in an aligned layout with $S = 6D$ in both streamwise and spanwise directions, geometrically equivalent to the optimally controlled wind farm from Chapter~\ref{ch:opt_induction}.
+Instead of optimizing $A$ and $St$ using a similar gradient-based optimization setup as in MM17, we perform a parameter sweep to find optimal parameter combinations.
+The reason for this is twofold.
+On the one hand we would need a rather long optimization horizon $T$ to find a parameter combination which is invariable of specific flow events.
+However, the chaotic nature of turbulent flow fields makes long-time optimization using adjoint LES significantly more complex (see, e.g.
+\cite{wang2014least}).
+On the other hand the fact that we have only two parameters renders a parameter sweep computationally feasible.
+The sweep is performed for a reduced-size wind-farm LES, as illustrated in Figure~\ref{fig:sinus_setup}.
+The farm consists of $4 \times 4$ turbines in an aligned layout with $S = 6D$ in both streamwise and spanwise directions, geometrically equivalent to the optimally controlled wind farm in MM17.
 The simulation is performed on a domain of $4 \times 2.4 \times 1$ km$^3$ with a simulation grid of $192 \times 256 \times 144$ gridpoints.
-Inflow conditions differ from those applied in the aforementioned simulations, and are generated by running an auxiliary shifted-periodic precursor simulation on a domain identical to the main simulation domain.
-All other simulation setup parameters are the same as those defined in Chapter~\ref{ch:opt_induction}.
+A wall roughness length $z_0 = 10^{-1}$ m is used.
 A set of wind-farm flow simulations is advanced in time by 30 minutes, during which the front row is controlled using a sinusoidally varying thrust coefficient $\cthat$, as defined in Equation \eqref{eq:define_thrust}.
@@
 	\centering
-	\includegraphics[width=0.75\textwidth]{setup_sinusoidal.eps}
+	\includegraphics[width=0.7\textwidth]{setup_sinusoidal.eps}
 	\caption{Reduced $4 \times 4$ wind-farm simulation setup for sinusoidal variation parameter study showing instantaneous contours of streamwise velocity $\widetilde{u}_x$.
@@
 The resulting maximum thrust coefficient of 3.5 could e.g.
-be attained by an NREL 5MW rotor with a 50\% increase in chord length (see \cite{goit2015optimal}, Appendix A).
+be attained by an NREL 5MW rotor with a 50\% increase in chord length (see \citealp{goit2015optimal}, Appendix A).
 Figure~\ref{fig:sinus_baseline}b illustrates normalized power extraction by row for the reference case, the best sinusoidal case, and the first four rows of the optimal control case C3t5 from MM17.
@@
 	\includegraphics[width=\textwidth]{gains_turbulent_6D_6D_6D_wide_inphase2.eps}
-	\caption{Power extraction of baseline sinusoidal thrust case ($S = 6D$, $z_0/H = 10^{-4}$).
+	\caption{Power extraction of baseline sinusoidal thrust case ($S = 6D$, $z_0 = 10^{-1}$ m).
 \emph{a) } Relative gain in mean wind-farm power extraction over reference case as a function of sine amplitude $A$ and frequency $St$.
-\emph{b) } Row-averaged mean power extraction for the best sinusoidal case, the optimal control case C3t5 from Chapter~\ref{ch:opt_induction}, and the reference case, normalized by first-row reference power.\label{fig:sinus_baseline} }
+\emph{b) } Row-averaged mean power extraction for the best sinusoidal case, the optimal control case C3t5 from MM17, and the reference case, normalized by first-row reference power.\label{fig:sinus_baseline} }
 \end{figure}
@@
 In contrast, Figure~\ref{fig:vorticity_parametersweep}c shows that, even though higher frequency thrust oscillations also result in periodic shedding of vortex rings, this does not automatically lead to more favorable flow conditions for downstream turbines.
+Therefore, it can be concluded that a correct timing and spacing of vortex rings is essential for increased wake mixing.
 \begin{figure}
@@
 
-It is interesting to note that, even though the current parameter sweep is performed using different initial and inlet conditions, the optimal frequency of sinusoidal variations in $\cthat$ at $St^\star = f D / U_\infty = 0.25$ corresponds to the location of the peak in the first-row thrust coefficient spectrum in Figure~\ref{fig:spectra_thrust_cthat}.
-This suggests that the current vortex-shedding mechanism is related to the flow-invariant characteristics for increasing power reported in Figure~\ref{fig:scrambled}.
+It is interesting to note that, even though the current parameter sweep is performed using different initial and inlet conditions than those applied in MM17, the optimal frequency of sinusoidal variations in $\cthat$ at $St^\star = f D / U_\infty = 0.25$ corresponds to the location of the peak in the first-row thrust coefficient spectrum of C3t5 in Figure~\ref{fig:spectra_thrust_cthat}.
+This also suggests that the current vortex-shedding mechanism is related to the flow-invariant characteristics for increasing power reported in Figure~\ref{fig:scrambled}.
 In the following paragraphs, the robustness of the best parameter pair for first-row thrust variations, i.e.
@@
 Figure~\ref{fig:sinus_roughness} depicts the power extraction results from a parameter sweep with the same wind-farm layout as in the baseline case, but with a tenfold increase in roughness length, i.e.
-$z_0 = 10^{-3}H$.
+$z_0 = 1$ m.
 This results in a turbulence intensity of $\approx$ 16\%, compared to $\approx$ 10\% in the baseline case.
@@
 	\includegraphics[width=\textwidth]{gains_turbulent_6D_6D_6D_wide_antiphase_z0_1m2.eps}
-	\caption{Power extraction of sinusoidal thrust case with increased wall roughness $z_0/H = 10^{-3}$.
+	\caption{Power extraction of sinusoidal thrust case with increased wall roughness $z_0 = 1$ m.
 \emph{a) } Relative gain in mean wind-farm power extraction over reference case as a function of sine amplitude $A$ and frequency $St$.
@@
 \subsubsection{Full-scale wind-farm test}
-In the remainder of this section, the sinusoidal variation strategy will be tested in a full-scale wind farm, corresponding to the $12 \times 6$ aligned wind farm of Chapter~\ref{ch:opt_induction}.
+In the remainder of this section, the sinusoidal variation strategy will be tested in a full-scale wind farm, corresponding to the full $12 \times 6$ aligned wind farm of MM17.
 Simulations are performed for a reduced range of amplitudes and Strouhal numbers, corresponding to the most favorable region identified in the parameter sweeps above.
@@
 	\includegraphics[width=\textwidth]{gains_fullscale.eps}
-	\caption{Power extraction of full-scale sinusoidal thrust case ($S = 6D$, $z_0/H = 10^{-4}$).
+	\caption{Power extraction of full-scale sinusoidal thrust case ($S = 6D$, $z_0 = 10^{-1}$ m).
 \emph{a) } Relative gain in mean wind-farm power extraction over reference case as a function of sine amplitude $A$ and frequency $St$.
-\emph{b) } relative gain in mean power extraction for the first four rows.
-\emph{c) } Row-averaged mean power extraction for the best sinusoidal case, the optimal control case C3t5 from Chapter~\ref{ch:opt_induction}, and the reference case, normalized by first-row reference power.\label{fig:sinus_fullscale} }
+\emph{b) } Relative gain in mean power extraction for the first four rows.
+\emph{c) } Row-averaged mean power extraction for the best sinusoidal case, the optimal control case C3t5 from MM17, and the reference case, normalized by first-row reference power.\label{fig:sinus_fullscale} }
 \end{figure}
@@
 In contrast, case C3t5, in which all turbines are actively controlled, succeeds to attain similar cross section characteristics with higher rotor velocities in the downstream as well.
-
-The current strategy hence succeeds in increasing total farm power for relatively short sets of aligned wind turbines.
-In contrast, larger wind farms require activity in downstream turbines as well to achieve significant gains in power extraction.
-However, it was found that the simple sinusoidal control strategy cannot be used in the intermediate rows.
-They are further discussed below in Section~\ref{sec:analysis_intermediate}
 
@@
 
-As shown throughout the current section, a qualitative analysis of instantaneous flow features in the optimal induction control wind farm from Chapter~\ref{ch:opt_induction} has led to the identification of a sinusoidal thrust control strategy for first-row turbines, resulting in increased power extraction in the second row.
-In contrast to earlier work on steady axial induction control, for which power gains have often been marginal or non-existent, the current analysis resulted in a robust axial induction control strategy with significant power gains, independent of specific local flow conditions.
-To the best knowledge of the author, this is the first time that optimal control simulations of wind farms are used to come up with simple control strategies, hence marking an important step towards practical dynamic axial induction control for increased power extraction.
-However, important comments should be made.
+As shown throughout the current section, a qualitative analysis of instantaneous flow features in the optimal induction control wind farm from MM17 has led to the identification of a sinusoidal thrust control strategy for first-row turbines, resulting in increased power extraction in the second row.
+In contrast to earlier work on steady axial induction control, for which power gains have often been marginal or non-existent (see, e.g.
+\citealp{annoni2016analysis}), the current analysis resulted in a robust axial induction control strategy with significant power gains, independent of specific local flow conditions.
+\red{To the best knowledge of the authors, this is the first time that optimal control simulations of wind farms are used to come up with simple induction control strategies, hence marking an important step towards practical dynamic axial induction control for increased power extraction.
+WM: dit is misschien een beetje obvious aangezien we voorlopig de enige zijn die echt LES-based optimal control doen..?} However, important comments should be made.
 First, sustained sinusoidal thrust variations with a large amplitude could contribute significantly to turbine fatigue loading.
@@
 Further verification using higher fidelity wind turbine models, such as actuator line models, and wind-tunnel testing is hence necessary.
+The current strategy hence allows to increase total power extraction for relatively short sets of aligned wind turbines.
+In contrast, larger wind farms require activity in downstream turbines as well to achieve significant gains in power extraction.
+However, it was found that the simple sinusoidal control strategy cannot be used in the intermediate rows.
+They are further discussed below in Section~\ref{sec:analysis_intermediate}.
 
-\section{Intermediate-row behavior}\label{sec:analysis_intermediate}
+\section{Intermediate-row turbine behavior}\label{sec:analysis_intermediate}
 The current section discusses the intermediate rows.
@@
 \conclusions \label{sec:analysis_summ}
-The current paper provided a further analysis of the thrust coefficient control characteristics for the C3t5 optimal control case from Chapter~\ref{ch:opt_induction}.
+The current paper provided a further analysis of the thrust coefficient control characteristics for the C3t5 optimal control case featured in \cite{munters2017optimal}.
 
@@
 This flow feature was succesfully mimicked using simple sinusoidal thrust actuation of the first row.
-The best parameter set for these sinusoidal variations proved robust to both wind-turbine spacing and turbulence intensity, with a non-dimensional frequency at $St^\star = 0.25$.
+The best parameter set for these sinusoidal variations proved robust to both wind-turbine spacing and turbulence intensity, with an amplitude $A^\star = 1.5$ and a non-dimensional frequency at $St^\star = 0.25$.
 Interestingly, this frequency corresponds to the peak at $St = 0.2 \dots 0.3$ observed in the first-row thrust coefficient spectra.
@@
 Although the first-row sinusoidal control led to a robust increase in total power for a reduced-size $4\times4$ wind farm, a full-scale test indicated that downstream turbine activity is required to obtain increased power at larger farm scales.
-Further, it was shown that the simple sinusoidal strategy does not lead to increase power extraction when applied to downstream intermediate turbines.
+It was also shown that the simple sinusoidal strategy does not lead to increase power extraction when applied to downstream intermediate turbines.
 Identifying the mechanisms for power increase in these turbines hence remains an important open research question.
+Finally, it is important to remark that all current simulations were performed using a standard non-rotating actuator disk model without the inclusion of mechanical turbine loading.
+Therefore, wind tunnel testing and/or simulations with more advanced turbine models (such as the actuator line model) including assessment of turbine loading are essential to evaluate the real-life applicability of the sinusoidal thrust strategy.
 

### 2018-02-06 draft.tex
Comments JM: abstract

@@
 \begin{abstract}
-	Wake interactions between wind turbines in wind farms entail reduced energy extraction in downstream rows, inciting research into coordinated wind-farm control strategies for optimal power extraction.
-In recent work, optimization and large-eddy simulation were combined to study optimal dynamic induction control of wind farms with power gains up to 20\% (Munters \& Meyers 2017 \emph{Phil Trans R Soc A} \textbf{375}, 20160100, doi:10.1098/rsta.2016.0100).
-However, computational costs associated with the optimal control simulations impede practical implementation of this work.
-Furthermore, the resulting optimal control signals are tuned to specific flow events and hence cannot be used in practice, and the complexity of these signals complicated the derivation of simplified control strategies.
-The current work focuses on further analysis of the optimization results of Munters \& Meyers.
-The analysis shows that wind-farm controls are optimized in a uni-directional manner with little upstream propagation of information, and that turbines can be classified into first-row, intermediate-row, and last-row turbines based on their optimal control dynamics.
-Although the control mechanisms for intermediate-row turbines remain unclear based on the current analysis, for first-row turbines it is found that the optimal controls increase wake mixing through periodically shed vortex rings.
-This behavior is mimicked with a simple sinusoidal thrust control strategy for first-row turbines, resulting in robust power gains for turbines in the entrance region of the farm.
+	Wake interactions between wind turbines in wind farms lead to reduced energy extraction in downstream rows.
+In recent work, optimization and large-eddy simulation were combined with optimal dynamic induction control of wind farms to study the mitigation of these effects, showing potential power gains up to 20\% (Munters \& Meyers 2017 \emph{Phil Trans R Soc A} \textbf{375}, 20160100, doi:10.1098/rsta.2016.0100).
+However, the computational cost associated with the optimal control simulations impedes practical implementation of this approach.
+Furthermore, the resulting control signals optimally react to the specific instantaneous turbulent flow realizations in the simulations, so that they cannot be simply be used in general.
+The current work focuses on the detailed analysis of the optimization results of Munters \& Meyers, with the aim to identify simplified control strategies that mimic the optimal control results and can be used in practice.
+The analysis shows that wind-farm controls are optimized in a parabolic manner with little upstream propagation of information.
+Moreover, turbines can be classified into first-row, intermediate-row, and last-row turbines based on their optimal control dynamics.
+At the moment, the control mechanisms for intermediate-row turbines remain unclear, but for first-row turbines we find that the optimal controls increase wake mixing by periodic shedding of vortex rings.
+This behavior can be mimicked with a simple sinusoidal thrust control strategy for first-row turbines, resulting in robust power gains for turbines in the entrance region of the farm.
 \end{abstract}

### 2018-02-06 draft.tex
Comments JM: introduction

@@
 The current control paradigm in such farms optimizes performance at the wind-turbine level and does not account for these interactions, resulting in sub-optimal wind-farm efficiency.
-In contrast, control strategies at the farm level allow to influence wake interactions and promise to improve overall wind-farm performance by improving wind conditions for downstream turbines, i.e.
-by redirecting propagating wakes (yaw control, see e.g.
+In contrast, control strategies at the farm level allow to influence wake interaction and promise to improve overall wind-farm performance by improving wind conditions for downstream turbines.
+This can be achieved by redirecting propagating wakes (yaw control, see e.g.
 \citealp{fleming2014evaluating, gebraad2016wind, campagnolo2016wind}) or by affecting the induced wake velocity deficits (axial induction control, see e.g.
@@
 
-In contrast to the studies cited above, which focus on static setpoint control of wind farms, \cite{goit2015optimal} introduced a dynamic induction control approach based on large-eddy simulations (LES) and adjoint gradient optimization.
-In this way, individual turbines could be used as dynamic flow actuators that influence the wind-farm boundary layer flow in such a way as to optimize aggregate wind-farm power extraction.
+In contrast to the studies cited above, that all focus on static setpoint optimization of wind farms, \cite{goit2015optimal} introduced a dynamic induction control approach based on large-eddy simulations (LES) and adjoint gradient optimization.
+In this study, individual turbines were used as dynamic flow actuators that influence the wind-farm boundary layer flow in such a way as to optimize aggregate wind-farm power extraction.
 The methodology was applied to the asymptotic case of a fully-developed `infinite' aligned wind farm, and power gains of about 16\% were quantified.
 Later, this approach was also used in induction control studies of wind farms with entrance effects in \cite{goit2016optimal} and, more recently, in \cite{munters2017optimal}, where similar gains in the order of 15\%--20\% were achieved.
-It is important to note that the computational cost of this dynamic induction control methodology is orders of magnitude too high for direct implementation, and the complex optimal control signals generated by the simulations are highly tuned to specific flow events and cannot be translated to practical flow conditions.
-However, the methodology allows to assess the theoretical potential for wind-farm control, and an increased understanding of the flow physics of controlled wind farms could lead to simplified control strategies that could be applied in practice.
+It is important to note that the computational cost of this LES-based dynamic induction control methodology is orders of magnitude too high for direct implementation.
+However, the methodology allows to assess the theoretical potential for wind-farm control, and an increased understanding of the flow physics of controlled wind farms can lead to simplified control strategies that can be applied in practice.
 
@@
 Furthermore, the high potential of combined induction and yaw control was quantified, and analysis of the yaw control signals allowed to identify practical simplified dynamic yaw control strategies.
-The search for similar practical control strategies for yaw control has remained unsuccessful to date.
+The search for similar practical control strategies for induction control has remained unsuccessful to date.
 
-The current paper presents efforts on understanding optimal control dynamics and identifying the flow mechanisms behind the power increase observed in the optimal induction control simulations in \cite{munters2017optimal} (further denoted as MM17), with the aim of harnessing them in tractable dynamic induction wind-farm controllers.
-The analysis presented here will lead to the identification of a coherent flow mechanism in the wake of optimally controlled first-row turbines, responsible for part of the power increase in the optimal control simulations.
-This resulted in a new derived simplified dynamic axial induction control strategy for first-row turbines producing a robust power increase in the second row, irrespective of precise local flow details.
-However, it will be shown that the controls responsible for the remainder of the power gains in the optimal control simulations are tuned to specific flow conditions.
-This complicates the analysis of downstream rows, and has thus far impeded the derivation of similar control strategies for the latter.
-
+The current paper presents efforts on understanding optimal control dynamics observed in the optimal induction control simulations by \cite{munters2017optimal} (further denoted as MM17).
 The outline of the paper is as follows: first, Section~\ref{sec:optimal_control_sims} discusses the numerical setup and optimal control simulations of MM17 that will be further analysed in the current paper.
@@
 Next, Section~\ref{sec:analysis_intermediate} shortly discusses the behavior of the intermediate rows, i.e.
-turbines which have both upstream and downstream neighbors, for which similar simple control strategies have not been identified thus far.
+turbines which have both upstream and downstream neighbors, for which similar simple control strategies remain elusive thus far.
 In conclusion, Section~\ref{sec:analysis_summ} summarizes the main findings of this paper.

### 2018-02-06 draft.tex
Comments JM: section 2

@@
 The current section describes the optimal control simulations performed by MM17, the results of which are further analyzed in the current paper.
-First, the methodology is illustrated.
+First, the methodology is introduced.
 Then, the numerical setup is detailed.
@@
 A schematic overview of the wind-farm control methodology is shown in Figure \ref{fig:optimization_meth}.
-Figure \ref{fig:optimization_meth}a illustrates the control block diagram: an iterative optimization loop updates the wind-farm control vector $\bs{\varphi}$ until a set of optimized controls $\bs{\varphi}^{\bullet}$ is found.
+Figure \ref{fig:optimization_meth}a illustrates the control block diagram: an iterative optimization loop updates the wind-farm control vector $\bs{\varphi}(t)$ until a set of optimized controls $\bs{\varphi}^{\bullet}(t)$ is found.
 This optimization is based upon an unsteady turbulence-resolving LES wind-farm flow model, and sensitivities of the cost functional $\J$ (i.e.
 the total wind-farm power extraction) are calculated using an adjoint formulation of this flow model.
-In this way, no a priori simplification on the model accuracy are made, and the control signals can be designed in a such a way that turbines can actively tap into the dynamic of the turbulent wind-farm boundary layer.
+In this way, a priori simplifications to the turbulent boundary layer and wake representation are avoided as much as possible, and the control signals are designed in a such a way that turbines actively tap into the dynamic of the turbulent flow.
 The optimization is performed using a receding-horizon control framework, as illustrated in Figure \ref{fig:optimization_meth}b.
-In this framework, wind-farm control is optimized for a finite time horizon $T$, involving a sequential set of LES and adjoint LES simulations.
+In this framework, wind-farm controls $\bs{\varphi}(t)$ are optimized for a finite time horizon $T$, involving a sequential set of LES and adjoint LES simulations.
 Upon convergence of the optimization, optimized control signals are applied in a flow advancement simulation for a time $T_A < T$, after which a new optimization window is initiated.
@@
 
-Within each optimization window, total wind-farm power is optimized by solving the following partial-differential-equation-constrained optimization problem: 
+Within each optimization window, the total wind-farm power is optimized by solving the following partial-differential-equation-constrained optimization problem: 
 \begin{alignat}{4}
@@
 \end{alignat}
-The cost functional that is minimized in \eqref{eq:costfunction} is the total wind-farm energy extraction over time horizon $T$, with switched sign for convention of notation.
+The cost functional that is optimized in \eqref{eq:costfunction} is the total wind-farm energy extraction over time horizon $T$.
 The control variables are the time-dependent thrust coefficient setpoints $C_{T,i}'$ of every turbine $i$ ($=1 \dots N_t$), i.e.
@@
 Further, the disk-averaged velocity is defined as $V_i = (1/A_i) \sint \R_i(\bs{x}) \utilde \cdot \eperpi ~\dx$, with $A_i$ the rotor disk area.
-Mechanical power captured by the wind turbine is calculated as $P_i = (1/2) C_{P,i}'(t) V_i(t)^3 A_i$, with $C_{P,i}' = 0.9 C_{T,i}'$, resulting from a fit of LES results to momentum theory, partially eliminating the overprediction of wind-turbine power on typical wind-farm LES grid resolutions \citep{munters2017optimal}.
+Mechanical power captured by the wind turbine is calculated as $P_i = (1/2) C_{P,i}'(t) V_i(t)^3 A_i$, with $C_{P,i}' = 0.9 C_{T,i}'$, resulting from a fit of LES results to momentum theory, eliminating the overprediction of wind-turbine power on typical wind-farm LES grid resolutions \citep{munters2017optimal}.
 
@@
 In the remainder of this section, time-averaged wind-farm flow properties will be investigated.
-Here and throughout the remainder of this paper, the focus will lie on case C3t5, as it produces similar energy gains as the highest-yield case C3t0 (see Figure \ref{fig:power_MM17}a), but achieves this with smooth thrust coefficient signals.
-In the following discussion, the time averaging operation is denoted by an overline, and flow field variables are decomposed into mean and fluctuating components as $u = \overline{u} + u' \equiv U + u'  $.
+Here and throughout the remainder of this paper, we focus on case C3t5, as it produces similar energy gains as the highest-yield case C3t0 (see Figure \ref{fig:power_MM17}a), but achieves this with smooth thrust coefficient signals.
+In the following discussion, the time averaging operation is denoted by an overline, and flow field variables are decomposed into mean and fluctuating components as $\widetilde{u} = \overline{\widetilde{u}} + \widetilde{u}' \equiv \widetilde{U} + \widetilde{u}'  $.
 Figure \ref{fig:flowfield_statistics} illustrates time-averaged flow field quantities of the reference case (left panels, a1 -- g1) and the differences between the optimized C3t5 case and the reference case (right panels, a2 -- g2).
@@
 Figures \ref{fig:flowfield_statistics}a and \ref{fig:flowfield_statistics}b illustrate sideviews and topviews of the axial velocity throughout the wind farm.
-It can be seen that downstream turbines are consistently subjected to higher incoming velocities, which explains the increased energy extraction discussed above.
+It can be seen that downstream turbines in the controlled case experience consistently higher incoming velocities, which explains the increased energy extraction discussed above.
 Furthermore, a larger drop in streamwise velocity over the turbine disk can be observed, most notably in the first-row turbines, indicating deeper wakes with enhanced recovery before hitting the next row of turbines.
@@
 A slight increase in turbulent transversal transport towards the wake centerline can be observed behind every row.
-On the other hand, mean-flow transversal momentum transport into the wake region is increased significantly behind the first two turbine rows only.
+The mean-flow transversal momentum transport into the wake region is increased significantly behind the first two turbine rows.
 Downstream of these rows, the difference between these cases and the reference case is far less coherent.
-The latter can be explained by the fact that it occurs in the region in which $\widetilde{U}_x$ in the intercolumn channels starts to deviate significantly from the reference case as shown in Figure~\ref{fig:flowfield_statistics}b.
+The latter can be explained by the fact that $\widetilde{U}_x$ in the intercolumn channels starts to deviate significantly from the reference case as shown in Figure~\ref{fig:flowfield_statistics}b.
 
-The analysis of flow features given above indicate that the optimal controls in case C3t5 coherently influence the wind-farm flow field in such a way as to provide better flow conditions for downstream turbines.
+The analysis of flow features given above indicates that the optimal controls in case C3t5 influence the wind-farm flow field in such a way as to provide better flow conditions for downstream turbines.
 Increased axial velocities are observed for all downstream turbines, and enhanced momentum transport towards the turbine region is achieved.

### 2018-02-06 draft.tex
Comments JM: section 3 (minus 3.5)

@@
 The current section focuses on the optimal thrust coefficients generated by the optimal control simulations in MM17, and performs numerical experiments to uncover some of the characteristics of these control signals.
-The aim of this analysis is to increase our overall understanding of how power is increased in optimally controlled wind-farm boundary layers, i.e.
-how the potential for power gains depends on the location of a turbine within the farm, whether optimal controls are fully tuned to specific flow events or contain flow-invariant features, and whether the control signals in neighboring turbines are correlated or can be used independently.
-This increased understanding will be used as a stepping stone towards the design of practical control strategies that do not require the computationally prohibitive LES-based methodology of MM17.
 
@@
 These optimizations provide an indication on how increased power potential is distributed among the rows, and allows to compare the resulting single-row optimized thrust coefficients with the fully cooperative coefficients from case MM17.
-Fourth, Section~\ref{sec:modified_thrust} evaluates the dependency of optimized thrust coefficients on specific flow conditions.
-This is done by running test simulations in which the coefficients are modified to eliminate any possible correlations to specific flow events at given times.
+Fourth, Section~\ref{sec:modified_thrust} evaluates the dependency of optimized thrust coefficients on the actual turbulent flow realization.
 Finally, Section~\ref{sec:analysis_discussion} discusses the main conclusions from the abovementioned sections, and summarizes the lessons learned.
@@
 Figure~\ref{fig:controls} illustrates the time evolution of some of  the thrust coefficients $\cthat$ in the C3t5 case.
-The figure illustrates that, for all rows but the last one (i.e.
+The figure shows that, for all rows but the last one (i.e.
 R12), $\cthat$ varies significantly in time, and that the amplitudes and frequencies of these variations are somewhat higher in the upstream rows of the farm.
 In contrast, row 12 features only minor unsteadiness at lower frequencies, and has an increased mean value of $\cthat$.
-This relatively steady behavior of the last row can be explained by the fact that there are no further downstream turbines that could benefit from row 12 actively influencing local flow conditions, hence the row optimizes its own power greedily.
-The increase in mean $\cthat$ can be explained based on Figure~\ref{fig:ct_sweep}, which illustrates the power extraction as a function of steady $\cthat$ for unwaked turbines, subject to identical turbulent inlet as in case C3t5.
+This relatively steady behavior of the last row can be explained by the fact that there are no further downstream turbines that can benefit from row 12 actively influencing local flow conditions, hence the row optimizes its own power only.
+The increase in mean $\cthat$ in row 12 can be explained based on Figure~\ref{fig:ct_sweep}, which shows the power extraction as a function of steady $\cthat$ for unwaked turbines, subject to identical turbulent inlet as in case C3t5.
 Although momentum theory predicts maximal power extraction for steady uniform inflow at $\cthat = 2$, the actual optimal steady value for the ADM at the current spatial resolution lies somewhat higher at $\cthat \approx 2.4$, for which power extraction is about 1.4\% higher than at $\cthat = 2$.
@@
 Although the linear fit $C_P' = a \cthat$ introduced in \cite{munters2017optimal}, Appendix~A eliminates the error in maximal power extraction, it does not correct the value of the optimal $\cthat$ (note that this could be achieved through a more complex relation between $C_P'$ and $\cthat$).
-Even though this explains the slight increase in mean $\cthat$ for greedily optimized last-row turbines, this does not undermine the overall power gains of the overinductive optimal control cases in MM17, which all feature increases well over 10\%.
 Returning to the more complex thrust coefficients in the other rows it is worth noting that, based on the current dataset, no statistically significant correlations between thrust coefficients of different turbines could be found.
@@
 
-Figure~\ref{fig:spectra_thrust_cthat}a,b illustrate power spectral densities of the row-averaged thrust forces and thrust coefficients respectively.
-The figure illustrates that the variances of both the thrust coefficients and their resulting forces are highest in row 1.
+Figure~\ref{fig:spectra_thrust_cthat}a,b shows row-averaged power spectral densities of the thrust forces and thrust coefficients respectively.
+The figure shows that the variances of both the thrust coefficients and their resulting forces are highest in row 1.
 Further downstream, rows 2 to 11 have very similar spectral behavior, and row 12 shows significantly lower variability.
 The high-frequency slopes of around $-5$ observed both for $f_T$ and $\cthat$ indicate that force variability on short timescales is caused mainly by thrust coefficient variations, whereas the slower thrust force dynamics tend more to a $-5/3$ slope, suggesting that these are governed by the unsteadiness in the turbulence instead.
-Note that, even though the spectra for all rows except row 12 collapse at frequencies below $0.05$ Hz, the first-row spectrum shows a small peak at $f \approx 0.02 \dots 0.03$ Hz ($f D / U_\infty \approx 0.2 \dots 0.3$) as indicated by the purple arrow, which also translates to an increased spectral content of the first row thrust force at these frequencies.
+Note that, even though the spectra for all rows except row 12 collapse at frequencies below $0.05$ Hz, the first-row spectrum shows a small peak at $f \approx 0.02 \dots 0.03$ Hz ($f D / U_\infty \approx 0.2 \dots 0.3$) as indicated by the purple arrow.
 It will be shown later in this paper that variations in the thrust coefficient around this frequency are directly related to increased power extraction.
@@
 
-In order to further study how the optimal controls increase overall wind-farm power, Figure~\ref{fig:multirow} illustrates power extraction resulting from applying a subset of the optimal controls to specific turbine rows only.
+In order to further study how the optimal controls increase overall wind-farm power, Figure~\ref{fig:multirow} shows power extraction resulting from applying a subset of the optimal controls to specific turbine rows only.
 Figure~\ref{fig:multirow}a depicts simulation results for which the  optimized controls are applied \emph{only to one specific row}, with the thrust coefficient in all other rows kept at the reference value of $\cthat = 2$.
 From the figure it can be seen that only for the controls of the first row (R1) this results in a significant power increase in rows 2 and 3.
-This indicates that the optimal controls, as generated by the optimization at the wind-farm level, depend on the precise flow conditions caused by upstream control actions, and can hence only be applied independently for the first row, which has no upstream dependence by definition.
+This indicates that the optimal controls, as generated by the optimization at the wind-farm level, react to the precise flow conditions caused by upstream control actions, and can hence only be applied independently for the first row, which has no upstream dependence by definition.
 
@@
 This suggests that self-optimization is very limited: the optimal controls for a given turbine are designed to create favorable flow conditions in the downstream rows instead of increasing local power.
-Furthermore, although the discussion in the previous paragraph has shown that downstream controls are optimized with the upstream actions in mind, the converse is not true: upstream control actions do not require a specific downstream response in order to increase power in downstream rows.
+Furthermore, although the discussion in the previous paragraph has shown that downstream controls are optimized with the upstream actions in mind, the converse is not true: upstream control actions do not require a specific downstream response in order to increase power in that downstream row.
 
@@
 \subsection{Optimization of single active turbine rows}\label{sec:optimization_single}
-The previous section has shown that, based on the full-farm optimization case, the first-row controls can be applied independently, whereas this does not work for the downstream rows.
+The previous section has shown that, based on the full-farm optimization case, the first-row controls can be applied independently from other turbine controls, whereas this does not work for the downstream rows.
 To further quantify the potential for increasing wind-farm power in each row of turbines, the current section considers a set of additional optimal control cases in which \emph{only a single active row is optimized}, with all other rows remaining passive.
@@
 
-Figure~\ref{fig:single_row_opt} illustrates the relative increase in wind-farm power extraction for each of the twelve individually optimized control cases.
-The optimization is run until the continuous adjoint gradient accuracy prevents further progress in the vicinity of a local minimum.
-\red{Upon interpreting the actual values from the figure, it is important to note that the reported power gain covers the full optimization horizon $T$, and is hence affected by finite-horizon effects.
+Figure~\ref{fig:single_row_opt} shows the relative increase in wind-farm power extraction for each of the twelve individually optimized control cases.
+The optimization is run until the continuous adjoint gradient accuracy prevents further progress in the optimization.
+Upon interpreting the actual values from the figure, it is important to note that the reported power gain covers the full optimization horizon $T$, and is hence affected by finite-horizon effects.
 Furthermore, the first window of an optimal control simulation, as considered here, contains an initial dead zone corresponding to the wake advection lag before upstream turbines start influencing their downstream neighbors.
@@
 Nevertheless, the relative order of the different cases still provides information that can be generalized to full optimal control studies with multiple windows and longer time horizons.
-(WM: dit eventueel weglaten?)}
 
@@
 This is not surprising as R1 produces the most power of all rows, and typically leaves the deepest wakes, causing second-row turbines to perform  poorly in aligned wind-farm layouts (see, e.g., \citealp{porte2013numerical,nilsson2015large}; and \citealp{stevens2016effects}).
-At the other end of the spectrum, the last row (R12) is the least fruitful.
-The intermediate rows (R2--R11) lie closer together, with the general trend being that potential is somewhat decreased with downstream distance into the wind farm, although this decrease is not monotonous.
+At the other end of the spectrum, the last row (R12) is the least useful.
+The intermediate rows (R2--R11) lie closer together, with the general trend being that the potential is somewhat decreased with downstream distance into the wind farm, although this decrease is not monotonous.
 
@@
 $T = 240$ s) allows for more interactions with directly neighboring turbines than with those located further downstream.
-Furthermore, except for the optimal control case of the last row (R12), self-optimization is virtually non-existent: power gains are achieved by modifying the flow to yield more favorable conditions in the downstream rows.
+Furthermore, except for the optimal control case of the last row (R12), self-optimization is virtually non-existent: power gains are achieved by modifying the flow to yield more favorable conditions for downstream rows.
 
@@
 
-Turning to the thrust dynamics, Figure~\ref{fig:control_C3t5_singlerow} shows the optimized thrust coefficients for the entire optimization horizon $T$ in several turbine rows for the single-row (gray) and full-farm (black) optimizations.
-Interestingly, the first-row thrust coefficient is virtually identical, regardless of whether downstream turbines are active or not.
-Although similarities can also be observed for the downstream rows, the correspondence is much weaker than in the first row.
-This can be explained by the fact that, in contrast to the first row, downstream turbines experience different local flow conditions based on whether the upstream turbines are active or not.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.8\textwidth]{fullfarm_singlerow.eps}
-	\caption{Optimized thrust coefficients in several rows for case C3t5 (black) and single-row optimal control cases (gray) for the full optimization horizon $T = 240$ s.\label{fig:control_C3t5_singlerow}}
-\end{figure}
-
-The current observations further confirm the conclusions from previous section: optimal control actions depend to some degree on the local flow conditions.
-Furthermore, previous section showed that first-row controls still lead to power gains in the absence of downstream control response.
-Here, as indicated by the similarity in first-row controls for single-row and full-farm optimization cases, it is shown that, for the first row, the optimizer is actually indifferent to whether or not downstream rows are actively controlled.
-
 \subsection{Modification of thrust coefficient signals}\label{sec:modified_thrust}
 The observations from previous sections illustrate that, at least to some degree, the optimized thrust coefficients are tuned to local flow conditions.
-In the current section, the possibility of whether the coefficients contain traits that are independent of flow conditions is nonetheless investigated.
+In the current section, the possibility of whether the coefficients contain traits that are independent of flow conditions is investigated.
 To this end, the optimized thrust coefficients are modified in such a way that correlations between them and specific flow events are eliminated.
@@
 
-The figure shows similar behavior for each of the modified control cases: only in the second row (R2), a consistent increase in power extraction can be observed.
-Combining this observation with the findings of previous sections, i.e.
-that power gains in a given row are caused by upstream control action instead of self-optimization, the current test cases suggest the presence of flow-invariant	features in the control signals of the first row.
+The figure shows similar behavior for each of the modified control cases: only in the second row (R2), a consistent (though small) increase in power extraction can be observed.
+This suggests the presence of flow-invariant	features in the control signals of the first row.
 Note however that the full power gain in the second row is only partially attained, indicating that also the first turbine row reacts to the specific flow conditions.
-The small power increase in the last row (R12) can again be explained by the slightly higher mean value in $\cthat$.
-Based on the current data, no consistent gains in any of the other rows are found.
 

### 2018-02-06 draft.tex
Comment JM: section 4 and 5 (minus figures)

@@
 \subsection{Flow field visualization}\label{sec:opt_visualization}
-With the aim of visualizing instantaneous flow field differences between controlled and reference cases, Figure~\ref{fig:vorticity_windfarm} illustrates snapshots of the velocity field at t = 300 s for the reference case (left) and the optimal control case C3t5 (right).
+Figure~\ref{fig:vorticity_windfarm} shows snapshots of the vorticity and velocity fields at t = 300 s for the reference case (left) and the optimal control case C3t5 (right).
 Figures~\ref{fig:vorticity_windfarm}a,b show isosurfaces of vorticity magnitude, colored by streamwise velocity $\widetilde{u}_x$.
@@
 The reason for this is twofold.
-On the one hand we would need a rather long optimization horizon $T$ to find a parameter combination which is invariable of specific flow events.
-However, the chaotic nature of turbulent flow fields makes long-time optimization using adjoint LES significantly more complex (see, e.g.
-\cite{wang2014least}).
+On the one hand we would need a rather long optimization horizon $T$ to find a robust parameter combination that is independent of specific flow realizations.
+However, the chaotic nature of turbulent flow fields makes long-time optimization using adjoint LES practically infeasible to date (see, e.g.
+\citealp{wang2014least}).
 On the other hand the fact that we have only two parameters renders a parameter sweep computationally feasible.
@@
 Within this set, the amplitude $A$ is varied between 0.5 and 2, with increments of 0.5.
-Furthermore, the Strouhal number $St$ is ranged between 0.05 and 0.6, with increments of 0.05.
+Furthermore, the Strouhal number $St$ is varied between 0.05 and 0.6, with increments of 0.05.
 In total, this leads to 48 LES cases within the set.
@@
 From the figure it can be seen that there is a well-defined range of values for $A$ and $St$ for which wind-farm power can be increased substantially through upstream sinusoidal thrust variations, with a maximal power increase of $\approx$ 5\% at $(St^\star, A^\star) = (0.25, 1.5)$.
-In order to put the technical feasibility of these parameters into context, note that a Strouhal number $St = 0.25$ corresponds here to a sine wave period of $\approx$ 50 s for a turbine with diameter $D = 100$ m and a freestream velocity $U_\infty = 8.5$ m s$^{-1}$.
-The resulting maximum thrust coefficient of 3.5 could e.g.
-be attained by an NREL 5MW rotor with a 50\% increase in chord length (see \citealp{goit2015optimal}, Appendix A).
+For instance, a Strouhal number $St = 0.25$ corresponds here to a sine wave period of $\approx$ 50 s for a turbine with diameter $D = 100$ m and a freestream velocity $U_\infty = 8.5$ m s$^{-1}$.
+The resulting maximum thrust coefficient of 3.5 can, e.g., be attained by an NREL 5MW rotor with a 50\% increase in chord length (see \citealp{goit2015optimal}, Appendix A).
 Figure~\ref{fig:sinus_baseline}b illustrates normalized power extraction by row for the reference case, the best sinusoidal case, and the first four rows of the optimal control case C3t5 from MM17.
@@
 It is interesting to note that, even though the current parameter sweep is performed using different initial and inlet conditions than those applied in MM17, the optimal frequency of sinusoidal variations in $\cthat$ at $St^\star = f D / U_\infty = 0.25$ corresponds to the location of the peak in the first-row thrust coefficient spectrum of C3t5 in Figure~\ref{fig:spectra_thrust_cthat}.
-This also suggests that the current vortex-shedding mechanism is related to the flow-invariant characteristics for increasing power reported in Figure~\ref{fig:scrambled}.
 In the following paragraphs, the robustness of the best parameter pair for first-row thrust variations, i.e.
@@
 
-Figure~\ref{fig:sinus_spacing} illustrates the power extraction resulting from two parameter sweeps with streamwise turbine spacings of $5D$ and $7D$ respectively.
+Figure~\ref{fig:sinus_spacing} shows the power extraction resulting from two parameter sweeps with streamwise turbine spacings of $5D$ and $7D$ respectively.
 Results are promising: for the given cases, $St^\star$ and $A^\star$  do not depend on streamwise turbine spacing.
@@
 $z_0 = 1$ m.
-This results in a turbulence intensity of $\approx$ 16\%, compared to $\approx$ 10\% in the baseline case.
+This results in a turbulence intensity of approximately 16\%, compared to 10\% in the baseline case.
 Again, the best parameter combination of $(St^\star, A^\star) = (0.25, 1.5)$ remains unchanged.
-Further, even for this higher turbulence case, in which downstream losses are lower due to naturally better wake mixing, power is significantly increased in the second row, leading to a relative gain in wind-farm power of around 2\%.
+Further, even for this higher turbulence case, in which downstream losses are lower due to naturally better wake mixing, power is increased in the second row, leading to a relative gain in wind-farm power of around 2\%.
 \begin{figure}
@@
 Figure~\ref{fig:sinus_fullscale}a and~\ref{fig:sinus_fullscale}b show the relative power gains over the reference case for the full wind farm and the first four turbine rows respectively.
-A first observation is that the optimal pair $(St^\star, A^\star)$ differs for both measures, suggesting that the increased power in front rows entails losses in the downstream.
+A first observation is that the optimal pair $(St^\star, A^\star)$ differs for both measures, suggesting that the increased power in front rows leads to losses in the downstream rows.
 This is confirmed by the row-wise power shown in Figure~\ref{fig:sinus_fullscale}c: from the fifth row onwards, power is slightly reduced in the sinusoidal cases.
@@
 \citealp{annoni2016analysis}), the current analysis resulted in a robust axial induction control strategy with significant power gains, independent of specific local flow conditions.
-\red{To the best knowledge of the authors, this is the first time that optimal control simulations of wind farms are used to come up with simple induction control strategies, hence marking an important step towards practical dynamic axial induction control for increased power extraction.
-WM: dit is misschien een beetje obvious aangezien we voorlopig de enige zijn die echt LES-based optimal control doen..?} However, important comments should be made.
+However, important comments should be made.
 First, sustained sinusoidal thrust variations with a large amplitude could contribute significantly to turbine fatigue loading.
@@
 Further verification using higher fidelity wind turbine models, such as actuator line models, and wind-tunnel testing is hence necessary.
-The current strategy hence allows to increase total power extraction for relatively short sets of aligned wind turbines.
-In contrast, larger wind farms require activity in downstream turbines as well to achieve significant gains in power extraction.
-However, it was found that the simple sinusoidal control strategy cannot be used in the intermediate rows.
-They are further discussed below in Section~\ref{sec:analysis_intermediate}.
 
@@
 First, as shown in the top panels of Figure~\ref{fig:cross_section_sinus}, even in the uncontrolled case, the kinetic energy of the flow in the vicinity of the turbine rotor is depleted more and more in the downstream rows.
-This complicates control strategies for these rows as the opportunity for increased mixing with high energy winds is decreased.
+This complicates control strategies for these rows as the opportunity for increased mixing with high energy flow is decreased.
 Furthermore, intermediate turbines are subjected to increased turbulence levels and more complex vorticity dynamics, as illustrated in Figure~\ref{fig:vorticity_windfarm}.
@@
 However, note that for instance R7 in Figure~\ref{fig:controls} also seems to show quasi-periodic sinusoidal variations in $\cthat$ at a time period of approximately 50~s.
-This could be an indication that, also for intermediate rows, vortex ring shedding could amount to part of the power increase of observed in the optimal control simulations, albeit at specific moments in time, synchronized with the local flow conditions.
+This could be an indication that, also for intermediate rows, vortex ring shedding could amount to part of the power increase observed in the optimal control simulations, albeit at specific moments in time, synchronized with the local flow conditions.
 
@@
 vorticity structures, high-speed turbulent streaks, or downdrafts) might be more promising.
-However, this requires further optimal control simulations over an extended time, as the total control time horizon of 30 minutes in the current dataset is expected to be insufficient for robust statistics in this kind of analysis.
+However, this requires further optimal control simulations over an extended time, as the total control time horizon of 30 minutes in the current dataset is insufficient for robust statistics in this kind of analysis.
 This is an important remaining challenge to be addressed in future research.

### 2018-02-06 draft.tex
Update figure 16 - smaller fonts

@@
 	\centering
-	\includegraphics[width=\textwidth]{pumpcases2.eps}
+	\includegraphics[width=\textwidth]{pumpcases3.eps}
 	\caption{Snapshots of wind-farm flow fields at t = 1 800 s.
@@
 \conclusions \label{sec:analysis_summ}
-The current paper provided a further analysis of the thrust coefficient control characteristics for the C3t5 optimal control case featured in \cite{munters2017optimal}.
+The current paper provided an analysis of the thrust coefficient control characteristics for the C3t5 optimal control case featured in \cite{munters2017optimal}.
 
 Analysis of the thrust coefficients and numerical experiments have shown a clear distinction between first-row turbines, last-row turbines, and intermediate turbines.
-Furthermore, observations strongly suggest that the optimization works in a unidirectional way: upstream turbines influence the flow field resulting in favorable conditions for their downstream neighbors, yet information on the possibility of active response and cooperation in the latter has very little influence on upstream control actions.
+Furthermore, observations strongly suggest that the optimization works in a unidirectional way: upstream turbines influence the flow field resulting in favorable conditions for their downstream neighbors, yet information on the possibility of active response and cooperation in the latter has no influence on upstream control actions.
 
@@
 The best parameter set for these sinusoidal variations proved robust to both wind-turbine spacing and turbulence intensity, with an amplitude $A^\star = 1.5$ and a non-dimensional frequency at $St^\star = 0.25$.
-Interestingly, this frequency corresponds to the peak at $St = 0.2 \dots 0.3$ observed in the first-row thrust coefficient spectra.
-Since the optimal value $St^\star$ was found using a different realization of the turbulent wind-farm flow, this indicates that the vortex-shedding mechanism is linked to the flow-invariant traits of the first-row optimal controls of case C3t5.
+Interestingly, this frequency corresponds to the peak at $St = 0.2 \dots 0.3$ observed in the first-row thrust coefficient spectra of the optimal control case.
 Although the first-row sinusoidal control led to a robust increase in total power for a reduced-size $4\times4$ wind farm, a full-scale test indicated that downstream turbine activity is required to obtain increased power at larger farm scales.
-It was also shown that the simple sinusoidal strategy does not lead to increase power extraction when applied to downstream intermediate turbines.
+It was also shown that the simple sinusoidal strategy does not lead to increased power extraction when applied to downstream intermediate turbines.
 Identifying the mechanisms for power increase in these turbines hence remains an important open research question.

### 2018-02-14 draft.tex
Added all Johans comments, version WM2_JM1

@@
 In the following discussion, the time averaging operation is denoted by an overline, and flow field variables are decomposed into mean and fluctuating components as $\widetilde{u} = \overline{\widetilde{u}} + \widetilde{u}' \equiv \widetilde{U} + \widetilde{u}'  $.
-Figure \ref{fig:flowfield_statistics} illustrates time-averaged flow field quantities of the reference case (left panels, a1 -- g1) and the differences between the optimized C3t5 case and the reference case (right panels, a2 -- g2).
+Figure~\ref{fig:flowfield_statistics} illustrates time-averaged flow field quantities of the reference case (left panels, a1 -- g1) and the differences between the optimized C3t5 case and the reference case (right panels, a2 -- g2).
 Simulation results are averaged over the different columns and are shown as either topviews at hub height (Figures \ref{fig:flowfield_statistics}b,f,g) or sideviews through a turbine column (Figures \ref{fig:flowfield_statistics}a,c,d,e).
@@
 
-The observations and experiments from previous sections have revealed information that increases the understanding the optimized thrust coefficients, and can be used as a starting point towards designing practical wind-farm controllers that do not require computationally prohibitive LES-based optimal control simulations.
-A first conclusion that can be drawn from the analysis above is that wind turbines can be classified into three distinct categories, based on their position within the farm: first-row turbines, last-row turbines, and intermediate turbines.
+The observations and experiments from previous sections have revealed information that increases the understanding the optimized thrust coefficients, and can be used as a starting point towards designing practical wind-farm controllers that do not require computationally expensive LES-based optimal control simulations.
 
+A first conclusion is that wind turbines can be classified into three distinct categories, based on their position within the farm: first-row turbines, last-row turbines, and intermediate turbines.
 The most salient behavior can be found in the first-row turbines (R1).
-Firstly, Figure~\ref{fig:spectra_thrust_cthat} showed that these turbines exhibit the largest variability in thrust coefficients and resulting thrust forces.
-Second, Figure~\ref{fig:multirow}a has shown that the optimized thrust coefficients in R1 are the only ones that can be applied independently from the other rows, and Figure~\ref{fig:single_row_opt} illustrated that the first row holds the greatest potential for power optimization of all rows in the farm.
-Third, the first row is the only one that has been shown to retain part of its power increase after eliminating the correlation between the optimized thrust coefficients and the flow field in Figure~\ref{fig:scrambled}, suggesting a flow-invariant component in the optimized coefficients.
-Finally, it is worth remarking the observed peak around $fD/U_\infty \approx 0.2 \dots 0.3$ from Figure~\ref{fig:spectra_thrust_cthat}.
-It will be shown below that variations in the thrust coefficient are related to specific flow events in the wake of R1.
-
-The characteristics of the last row turbines (R12) also stand out from the rest.
-This is mainly caused by the fact that, by definition, the last row has no downstream turbines for which it can generate more favorable flow conditions, causing them to operate greedily in an optimal control setting.
-This explains the relatively quiescent behavior of optimized thrust coefficients in Figure~\ref{fig:controls}, and the limited potential for power increase in Figure~\ref{fig:single_row_opt}.
-
-The remaining intermediate rows (R2--R11) have similar spectral thrust characteristics (Figure~\ref{fig:spectra_thrust_cthat}) and potential for power increase (Figure~\ref{fig:single_row_opt}), situated in between but clearly separated from first- and last-row turbines.
+It was shown that these turbines exhibit the largest variability in thrust forces and hold the greatest potential for power optimization.
+Furthermore, they are not influenced by upstream turbine control action, and are the only turbines that retain part of the power gains after eliminating possible correlation between controls and specific flow events.
+The characteristics of the last row turbines (R12) also stand out from the rest due to the fact that, by definition, the last row has no downstream turbines and hence holds no further potential for coordinated control.
+The remaining intermediate rows (R2--R11) have similar spectral thrust characteristics and potential for power increase, situated in between but clearly separated from first- and last-row turbines.
 Further, it is worth noting that the behavior and analysis of control actions in these turbines is most complex: not only do they influence downstream turbines, they in turn are dependent on controls of upstream turbines.
@@
 A second conclusion is that, whether or not the wind-farm is controlled with the possibility of active response and cooperation between turbines, the resulting power and thrust characteristics are very similar.
-Firstly, both Figure~\ref{fig:multirow}b (full-farm optimization) and Figure~\ref{fig:gains_singlerow} (single-row optimization) suggest that self-optimization of turbines is negligible, and that power gains are mainly achieved by generating more favorable flow conditions for downstream turbines.
-Furthermore, the remarkable match between optimized thrust coefficients for full-farm and single-row optimal control cases from Figure~\ref{fig:control_C3t5_singlerow} indicates that, at least for the first row, upstream controls do not rely on active downstream response.
-Instead, Figure~\ref{fig:multirow}b shows that, for any row $i$, the full potential in power increase is virtually attained by applying controls only for the upstream turbines up until row $i-1$.
-Furthermore, Figure~\ref{fig:multirow}a illustrates that optimized thrust coefficients in downstream rows by themselves do not result in power increase without also applying optimized coefficients in their upstream neighbors.
-This indicates that controls are tuned to specific flow events.
-This also explains the discrepancies between full-farm and single-row optimized thrust coefficients for downstream rows in Figure~\ref{fig:control_C3t5_singlerow}.
+It was shown that self-optimization is very limited, and that, for any row $i$, the full potential in power increase is virtually attained by applying controls only for the upstream turbines up until row $i-1$.
 These observations strongly suggest that the optimized thrust coefficients are designed in a parabolic manner, i.e.
 with a unidirectional propagation of control information from the first row to the last, and very little upstream influence of downstream turbine actions.
-
 With this in mind, the following section of this paper will focus on the first and most promising link of the control chain: the turbines situated in the first row of the wind farm.
@@
 \section{First-row turbine behavior}\label{sec:analysis_flow_vis}
-As shown in the previous section, the behavior of turbines within the first upstream row stands out from the others.
+%As shown in the previous section, the behavior of turbines within the first upstream row stands out from the others.
 Firstly, they are not influenced by any dynamic actions of the upstream turbines which makes their behavior somewhat easier to study.
@@
 
+The current section further focuses on the analysis of the first row turbines.
 First, a qualitative analysis of the instantaneous flow field is performed in Section~\ref{sec:opt_visualization}, resulting in the observation of vortex rings being shed from first-row turbines.
@@
 Figure~\ref{fig:sinus_fullscale} shows the power extraction of the full-scale test.
-Figure~\ref{fig:sinus_fullscale}a and~\ref{fig:sinus_fullscale}b show the relative power gains over the reference case for the full wind farm and the first four turbine rows respectively.
-A first observation is that the optimal pair $(St^\star, A^\star)$ differs for both measures, suggesting that the increased power in front rows leads to losses in the downstream rows.
-This is confirmed by the row-wise power shown in Figure~\ref{fig:sinus_fullscale}c: from the fifth row onwards, power is slightly reduced in the sinusoidal cases.
-This explains the negligible increase in total farm power observed in Figure~\ref{fig:sinus_fullscale}a.
+Figure~\ref{fig:sinus_fullscale}a shows the relative power gains over the reference case for the full wind farm.
+It can be seen that the total power gain or loss is below 0.5\% for each of the sinusoidal control cases.
+Figure~\ref{fig:sinus_fullscale}b shows the row-wise power extraction for the reference case, the sinusoidal thrust case with $(St, A) = (0.25, 1.5)$, and the optimal control case C3t5.
+It is shown that, although the second row of the sinusoidal thrust case achieves similar power gains as those observed above, from the fifth row onwards power is slightly reduced in the sinusoidal case.
 
@@
 \emph{a) } Relative gain in mean wind-farm power extraction over reference case as a function of sine amplitude $A$ and frequency $St$.
-\emph{b) } Relative gain in mean power extraction for the first four rows.
-\emph{c) } Row-averaged mean power extraction for the best sinusoidal case, the optimal control case C3t5 from MM17, and the reference case, normalized by first-row reference power.\label{fig:sinus_fullscale} }
+\emph{b) } Row-averaged mean power extraction for the best sinusoidal case, the optimal control case C3t5 from MM17, and the reference case, normalized by first-row reference power.\label{fig:sinus_fullscale} }
 \end{figure}

### 2018-02-14 draft.tex
Added H2020 funding

@@
 \begin{acknowledgements}
-The authors acknowledge funding by the European Research Council (grant no.
-306471).
+The authors have received funding from the European Union's Horizon 2020 research and innovation programme under grant agreement No.
+727680 (TotalControl).
+The authors also acknowledge funding by the European Research Council under grant agreement No.
+306471 (ActiveWindFarms).
 The computational resources and services used in this work were provided by the VSC (Flemish Supercomputer Center), funded by the Research Foundation - Flanders (FWO) and the Flemish Government, department EWI.

### 2018-02-14 draft.tex
Include Johan's comments

@@
 	Wake interactions between wind turbines in wind farms lead to reduced energy extraction in downstream rows.
-In recent work, optimization and large-eddy simulation were combined with optimal dynamic induction control of wind farms to study the mitigation of these effects, showing potential power gains up to 20\% (Munters \& Meyers 2017 \emph{Phil Trans R Soc A} \textbf{375}, 20160100, doi:10.1098/rsta.2016.0100).
-However, the computational cost associated with the optimal control simulations impedes practical implementation of this approach.
-Furthermore, the resulting control signals optimally react to the specific instantaneous turbulent flow realizations in the simulations, so that they cannot be simply be used in general.
+In recent work, optimization and large-eddy simulation were combined with optimal dynamic induction control of wind farms to study the mitigation of these effects, showing potential power gains of up to 20\% (Munters \& Meyers 2017 \emph{Phil Trans R Soc A} \textbf{375}, 20160100, doi:10.1098/rsta.2016.0100).
+However, the computational cost associated with these optimal control simulations impedes practical implementation of this approach.
+Furthermore, the resulting control signals optimally react to the specific instantaneous turbulent flow realizations in the simulations, so that they cannot be simply used in general.
 The current work focuses on the detailed analysis of the optimization results of Munters \& Meyers, with the aim to identify simplified control strategies that mimic the optimal control results and can be used in practice.
@@
 Later, this approach was also used in induction control studies of wind farms with entrance effects in \cite{goit2016optimal} and, more recently, in \cite{munters2017optimal}, where similar gains in the order of 15\%--20\% were achieved.
-It is important to note that the computational cost of this LES-based dynamic induction control methodology is orders of magnitude too high for direct implementation.
-However, the methodology allows to assess the theoretical potential for wind-farm control, and an increased understanding of the flow physics of controlled wind farms can lead to simplified control strategies that can be applied in practice.
+It is important to note that the computational cost of this LES-based dynamic induction control methodology is orders of magnitude too high for direct implementation as a practical control strategy.
+However, the methodology allows to assess the theoretical potential for wind-farm control, and increased understanding of the flow physics can lead to simplified control strategies that can be applied in practice.
 
@@
 the total wind-farm power extraction) are calculated using an adjoint formulation of this flow model.
-In this way, a priori simplifications to the turbulent boundary layer and wake representation are avoided as much as possible, and the control signals are designed in a such a way that turbines actively tap into the dynamic of the turbulent flow.
+In this way, a priori simplifications to the turbulent boundary layer and wake representation are avoided as much as possible, and the control signals are designed in a such a way that turbines actively tap into the dynamics of the turbulent flow.
 The optimization is performed using a receding-horizon control framework, as illustrated in Figure \ref{fig:optimization_meth}b.
@@
 In the remainder of this section, time-averaged wind-farm flow properties will be investigated.
-Here and throughout the remainder of this paper, we focus on case C3t5, as it produces similar energy gains as the highest-yield case C3t0 (see Figure \ref{fig:power_MM17}a), but achieves this with smooth thrust coefficient signals.
+Here and throughout the remainder of this paper, we focus on case C3t5, as it produces similar energy gains as the highest-yield case C3t0 (see Figure \ref{fig:power_MM17}a), but achieves this with smoother thrust coefficient signals.
 In the following discussion, the time averaging operation is denoted by an overline, and flow field variables are decomposed into mean and fluctuating components as $\widetilde{u} = \overline{\widetilde{u}} + \widetilde{u}' \equiv \widetilde{U} + \widetilde{u}'  $.
@@
 From the figure it can be seen that only for the controls of the first row (R1) this results in a significant power increase in rows 2 and 3.
-This indicates that the optimal controls, as generated by the optimization at the wind-farm level, react to the precise flow conditions caused by upstream control actions, and can hence only be applied independently for the first row, which has no upstream dependence by definition.
+This indicates that the optimal controls, as generated by the optimization at the wind-farm level, react to the precise flow conditions caused by upstream control actions, and can hence only be applied independently for the first row, which has no upstream dependence on other controls.
 
@@
 Instead of optimizing $A$ and $St$ using a similar gradient-based optimization setup as in MM17, we perform a parameter sweep to find optimal parameter combinations.
-The reason for this is twofold.
-On the one hand we would need a rather long optimization horizon $T$ to find a robust parameter combination that is independent of specific flow realizations.
-However, the chaotic nature of turbulent flow fields makes long-time optimization using adjoint LES practically infeasible to date (see, e.g.
+The reason for this is that we would need a rather long optimization horizon $T$ to find a robust parameter combination that is independent of specific flow realizations.
+Unfortunately, the chaotic nature of turbulent flow fields makes long-time optimization using adjoint LES practically infeasible to date (see, e.g.
 \citealp{wang2014least}).
-On the other hand the fact that we have only two parameters renders a parameter sweep computationally feasible.
+However, the fact that we have only two parameters renders a parameter sweep computationally feasible.
 The sweep is performed for a reduced-size wind-farm LES, as illustrated in Figure~\ref{fig:sinus_setup}.
@@
 A wall roughness length $z_0 = 10^{-1}$ m is used.
-A set of wind-farm flow simulations is advanced in time by 30 minutes, during which the front row is controlled using a sinusoidally varying thrust coefficient $\cthat$, as defined in Equation \eqref{eq:define_thrust}.
+A set of wind-farm flow simulations is advanced in time by 30~minutes, during which the front row is controlled using a sinusoidally varying thrust coefficient $\cthat$, as defined in Equation \eqref{eq:define_thrust}.
 Within this set, the amplitude $A$ is varied between 0.5 and 2, with increments of 0.5.
@@
 For instance, a Strouhal number $St = 0.25$ corresponds here to a sine wave period of $\approx$ 50 s for a turbine with diameter $D = 100$ m and a freestream velocity $U_\infty = 8.5$ m s$^{-1}$.
-The resulting maximum thrust coefficient of 3.5 can, e.g., be attained by an NREL 5MW rotor with a 50\% increase in chord length (see \citealp{goit2015optimal}, Appendix A).
+The resulting maximum thrust coefficient of 3.5 can, e.g., be attained by an NREL 5MW rotor with a 50\% increase in chord length at a tip speed ratio that is 25\% higher than the greedy tip speed ratio (see \citealp{goit2015optimal}, Appendix A).
 Figure~\ref{fig:sinus_baseline}b illustrates normalized power extraction by row for the reference case, the best sinusoidal case, and the first four rows of the optimal control case C3t5 from MM17.
@@
 	\centering
-	\includegraphics[width=\textwidth]{gains_fullscale.eps}
+	\includegraphics[width=\textwidth]{gains_fullscale2.eps}
 	\caption{Power extraction of full-scale sinusoidal thrust case ($S = 6D$, $z_0 = 10^{-1}$ m).
@@
 As shown throughout the current section, a qualitative analysis of instantaneous flow features in the optimal induction control wind farm from MM17 has led to the identification of a sinusoidal thrust control strategy for first-row turbines, resulting in increased power extraction in the second row.
-In contrast to earlier work on steady axial induction control, for which power gains have often been marginal or non-existent (see, e.g.
-\citealp{annoni2016analysis}), the current analysis resulted in a robust axial induction control strategy with significant power gains, independent of specific local flow conditions.
 However, important comments should be made.
@@
 Furthermore, intermediate turbines are subjected to increased turbulence levels and more complex vorticity dynamics, as illustrated in Figure~\ref{fig:vorticity_windfarm}.
-This could explain why the sinusoidal thrust control for the first row did not lead to increase power when applied to the second row: whereas the first row produces increased mixing by destabilizing relatively stable vortex sheets into vortex rings, the second row is already continuously immersed in complex vorticity patterns for which this simple approach does not work.
+This could explain why sinusoidal thrust control did not lead to increased power when applied to the second row: whereas the first row produces increased mixing by destabilizing relatively stable vortex sheets into vortex rings, the second row is already continuously immersed in complex vorticity patterns for which this simple approach does not work.
 However, note that for instance R7 in Figure~\ref{fig:controls} also seems to show quasi-periodic sinusoidal variations in $\cthat$ at a time period of approximately 50~s.
-This could be an indication that, also for intermediate rows, vortex ring shedding could amount to part of the power increase observed in the optimal control simulations, albeit at specific moments in time, synchronized with the local flow conditions.
+This is an indication that, also for intermediate rows, vortex ring shedding could amount to part of the power increase observed in the optimal control simulations, albeit at specific moments in time, synchronized with the local flow conditions.
 
@@
 vorticity structures, high-speed turbulent streaks, or downdrafts) might be more promising.
-However, this requires further optimal control simulations over an extended time, as the total control time horizon of 30 minutes in the current dataset is insufficient for robust statistics in this kind of analysis.
+This requires further optimal control simulations over an extended time, as the total control time horizon of 30 minutes in the current dataset is insufficient for robust statistics in this kind of analysis.
 This is an important remaining challenge to be addressed in future research.

### 2018-02-14 draft.tex
Section --> Sect.

@@
 The current paper presents efforts on understanding optimal control dynamics observed in the optimal induction control simulations by \cite{munters2017optimal} (further denoted as MM17).
-The outline of the paper is as follows: first, Section~\ref{sec:optimal_control_sims} discusses the numerical setup and optimal control simulations of MM17 that will be further analysed in the current paper.
+The outline of the paper is as follows: first, Sect.~\ref{sec:optimal_control_sims} discusses the numerical setup and optimal control simulations of MM17 that will be further analysed in the current paper.
 Section~\ref{sec:analysis_controls_analysis} presents an analysis of the control and thrust force dynamics, and performs some numerical experiments to elucidate characteristics of the optimal controls.
 It will be shown that the coherent behavior of turbines situated in the first row of the wind farm stands out from their downstream counterparts.
-Hereafter, Section~\ref{sec:analysis_flow_vis} identifies the shedding of vortex rings from the first row based on a flow visualization.
+Hereafter, Sect.~\ref{sec:analysis_flow_vis} identifies the shedding of vortex rings from the first row based on a flow visualization.
 Further, a simple sinusoidal thrust control approach is presented that successfully mimics this process with a robust increase in power extraction extraction in the second row.
-Next, Section~\ref{sec:analysis_intermediate} shortly discusses the behavior of the intermediate rows, i.e.
+Next, Sect.~\ref{sec:analysis_intermediate} shortly discusses the behavior of the intermediate rows, i.e.
 turbines which have both upstream and downstream neighbors, for which similar simple control strategies remain elusive thus far.
-In conclusion, Section~\ref{sec:analysis_summ} summarizes the main findings of this paper.
+In conclusion, Sect.~\ref{sec:analysis_summ} summarizes the main findings of this paper.
 
@@
 
-First, the thrust coefficient signals themselves are analyzed in Section~\ref{sec:analysis_thrust}.
-Second, the optimized thrust coefficients are applied only to subsets of turbine rows in Section~\ref{sec:application_subset}.
+First, the thrust coefficient signals themselves are analyzed in Sect.~\ref{sec:analysis_thrust}.
+Second, the optimized thrust coefficients are applied only to subsets of turbine rows in Sect.~\ref{sec:application_subset}.
 In this way, the interdependency of optimized thrust coefficients in different rows can be evaluated.
-Third, additional optimal control simulations, in which only one single active row is optimized, are discussed in Section~\ref{sec:optimization_single}.
+Third, additional optimal control simulations, in which only one single active row is optimized, are discussed in Sect.~\ref{sec:optimization_single}.
 These optimizations provide an indication on how increased power potential is distributed among the rows, and allows to compare the resulting single-row optimized thrust coefficients with the fully cooperative coefficients from case MM17.
-Fourth, Section~\ref{sec:modified_thrust} evaluates the dependency of optimized thrust coefficients on the actual turbulent flow realization.
-Finally, Section~\ref{sec:analysis_discussion} discusses the main conclusions from the abovementioned sections, and summarizes the lessons learned.
+Fourth, Sect.~\ref{sec:modified_thrust} evaluates the dependency of optimized thrust coefficients on the actual turbulent flow realization.
+Finally, Sect.~\ref{sec:analysis_discussion} discusses the main conclusions from the abovementioned sections, and summarizes the lessons learned.
 
@@
 The current section further focuses on the analysis of the first row turbines.
-First, a qualitative analysis of the instantaneous flow field is performed in Section~\ref{sec:opt_visualization}, resulting in the observation of vortex rings being shed from first-row turbines.
-Thereafter, this mechanism is mimicked by imposing sinusoidally varying thrust coefficients in these turbines in Section~\ref{sec:opt_sinus}, with the aim of increasing power through similar mechanisms as in the computationally expensive optimal control cases.
+First, a qualitative analysis of the instantaneous flow field is performed in Sect.~\ref{sec:opt_visualization}, resulting in the observation of vortex rings being shed from first-row turbines.
+Thereafter, this mechanism is mimicked by imposing sinusoidally varying thrust coefficients in these turbines in Sect.~\ref{sec:opt_sinus}, with the aim of increasing power through similar mechanisms as in the computationally expensive optimal control cases.
 

### 2018-02-14 draft.tex
Equation --> Eq.

@@
 A conventionally (greedily) controlled wind farm with steady $C_T' = 2$ (corresponding to the Betz limit in momentum theory) was defined as a reference case.
-Several different optimal control cases were defined, based on the wind-turbine response time $\tau = $0, 5, or 30 s (instantaneous, fast, or slow response, see Equation \ref{eq:ctihat_constraint}) and the maximal thrust coefficient $\ctmax = 2$ or 3 (underinductive or overinductive, see Equation \ref{eq:boxct_constraint}).
+Several different optimal control cases were defined, based on the wind-turbine response time $\tau = $0, 5, or 30 s (instantaneous, fast, or slow response, see Eq.
+\ref{eq:ctihat_constraint}) and the maximal thrust coefficient $\ctmax = 2$ or 3 (underinductive or overinductive, see Eq.
+\ref{eq:boxct_constraint}).
 Cases are denoted as C<X>t<Y>, where X and Y represent $\ctmax$ and $\tau$ respectively, e.g.
@@
 A wall roughness length $z_0 = 10^{-1}$ m is used.
-A set of wind-farm flow simulations is advanced in time by 30~minutes, during which the front row is controlled using a sinusoidally varying thrust coefficient $\cthat$, as defined in Equation \eqref{eq:define_thrust}.
+A set of wind-farm flow simulations is advanced in time by 30~minutes, during which the front row is controlled using a sinusoidally varying thrust coefficient $\cthat$, as defined in Eq.
+\eqref{eq:define_thrust}.
 Within this set, the amplitude $A$ is varied between 0.5 and 2, with increments of 0.5.

### 2018-02-14 draft.tex
Figure --> Fig.

@@
 
-A schematic overview of the wind-farm control methodology is shown in Figure \ref{fig:optimization_meth}.
+A schematic overview of the wind-farm control methodology is shown in Fig.
+\ref{fig:optimization_meth}.
 Figure \ref{fig:optimization_meth}a illustrates the control block diagram: an iterative optimization loop updates the wind-farm control vector $\bs{\varphi}(t)$ until a set of optimized controls $\bs{\varphi}^{\bullet}(t)$ is found.
@@
 In this way, a priori simplifications to the turbulent boundary layer and wake representation are avoided as much as possible, and the control signals are designed in a such a way that turbines actively tap into the dynamics of the turbulent flow.
-The optimization is performed using a receding-horizon control framework, as illustrated in Figure \ref{fig:optimization_meth}b.
+The optimization is performed using a receding-horizon control framework, as illustrated in Fig.
+\ref{fig:optimization_meth}b.
 In this framework, wind-farm controls $\bs{\varphi}(t)$ are optimized for a finite time horizon $T$, involving a sequential set of LES and adjoint LES simulations.
@@
 The flow is simulated on a domain of $10 \times 3.6 \times 1$ km$^3$, discretized on a simulation grid of $384 \times 192 \times 144$ grid points.
-A snapshot of the streamwise velocity field is shown in Figure \ref{fig:flowfield_MM17}.
+A snapshot of the streamwise velocity field is shown in Fig.
+\ref{fig:flowfield_MM17}.
 The wind farm was controlled over a total of $N_A = 15$ time windows with a prediction horizon $T = 240 s$ and a flow advancement time of $T_A = T/2 = 120 s$, resulting in a total control time $T_{\rm tot} = N_A T_A = 30$ minutes.
@@
 Figure \ref{fig:power_MM17}a shows that the adjoint LES-based control approaches achieves energy gains ranging from a minor 2\% in the most restrictive C2t30 case to over 20\% in the C3t0 case.
-From Figure \ref{fig:power_MM17}b it can be seen that, for all cases except C3t30, power is curtailed in the first row to a limited degree, whereas the downstream rows compensate for this loss by extracting significantly more energy.
+From Fig.
+\ref{fig:power_MM17}b it can be seen that, for all cases except C3t30, power is curtailed in the first row to a limited degree, whereas the downstream rows compensate for this loss by extracting significantly more energy.
 Furthermore, not taking into account the first row, the last row achieves the highest energy extraction in every case, as it can act greedily without compromising power extraction in downstream neighbors.
@@
 In the remainder of this section, time-averaged wind-farm flow properties will be investigated.
-Here and throughout the remainder of this paper, we focus on case C3t5, as it produces similar energy gains as the highest-yield case C3t0 (see Figure \ref{fig:power_MM17}a), but achieves this with smoother thrust coefficient signals.
+Here and throughout the remainder of this paper, we focus on case C3t5, as it produces similar energy gains as the highest-yield case C3t0 (see Fig.
+\ref{fig:power_MM17}a), but achieves this with smoother thrust coefficient signals.
 In the following discussion, the time averaging operation is denoted by an overline, and flow field variables are decomposed into mean and fluctuating components as $\widetilde{u} = \overline{\widetilde{u}} + \widetilde{u}' \equiv \widetilde{U} + \widetilde{u}'  $.
 Figure~\ref{fig:flowfield_statistics} illustrates time-averaged flow field quantities of the reference case (left panels, a1 -- g1) and the differences between the optimized C3t5 case and the reference case (right panels, a2 -- g2).
-Simulation results are averaged over the different columns and are shown as either topviews at hub height (Figures \ref{fig:flowfield_statistics}b,f,g) or sideviews through a turbine column (Figures \ref{fig:flowfield_statistics}a,c,d,e).
+Simulation results are averaged over the different columns and are shown as either topviews at hub height (Figs.
+\ref{fig:flowfield_statistics}b,f,g) or sideviews through a turbine column (Figs.
+\ref{fig:flowfield_statistics}a,c,d,e).
 
@@
 Downstream of these rows, the difference between these cases and the reference case is far less coherent.
-The latter can be explained by the fact that $\widetilde{U}_x$ in the intercolumn channels starts to deviate significantly from the reference case as shown in Figure~\ref{fig:flowfield_statistics}b.
+The latter can be explained by the fact that $\widetilde{U}_x$ in the intercolumn channels starts to deviate significantly from the reference case as shown in Fig.~\ref{fig:flowfield_statistics}b.
 
@@
 This relatively steady behavior of the last row can be explained by the fact that there are no further downstream turbines that can benefit from row 12 actively influencing local flow conditions, hence the row optimizes its own power only.
-The increase in mean $\cthat$ in row 12 can be explained based on Figure~\ref{fig:ct_sweep}, which shows the power extraction as a function of steady $\cthat$ for unwaked turbines, subject to identical turbulent inlet as in case C3t5.
+The increase in mean $\cthat$ in row 12 can be explained based on Fig.~\ref{fig:ct_sweep}, which shows the power extraction as a function of steady $\cthat$ for unwaked turbines, subject to identical turbulent inlet as in case C3t5.
 Although momentum theory predicts maximal power extraction for steady uniform inflow at $\cthat = 2$, the actual optimal steady value for the ADM at the current spatial resolution lies somewhat higher at $\cthat \approx 2.4$, for which power extraction is about 1.4\% higher than at $\cthat = 2$.
@@
 
-In order to further study how the optimal controls increase overall wind-farm power, Figure~\ref{fig:multirow} shows power extraction resulting from applying a subset of the optimal controls to specific turbine rows only.
+In order to further study how the optimal controls increase overall wind-farm power, Fig.~\ref{fig:multirow} shows power extraction resulting from applying a subset of the optimal controls to specific turbine rows only.
 Figure~\ref{fig:multirow}a depicts simulation results for which the  optimized controls are applied \emph{only to one specific row}, with the thrust coefficient in all other rows kept at the reference value of $\cthat = 2$.
@@
 Furthermore, in contrast to the single-row control simulations from the previous section (i.e.
-in Figure~\ref{fig:multirow}a), the current optimizations will yield controls that are \emph{explicitly designed} to increase power given that all other rows are passive.
+in Fig.~\ref{fig:multirow}a), the current optimizations will yield controls that are \emph{explicitly designed} to increase power given that all other rows are passive.
 To limit computational costs, the additional optimizations are only performed for a single time window.
@@
 To avoid erroneous conclusions based on coincidence, the column swap is performed in 2 random independent ways.
-Row-averaged power for these cases is shown in Figure~\ref{fig:scrambled}a.
+Row-averaged power for these cases is shown in Fig.~\ref{fig:scrambled}a.
 
@@
 In this way, the spectral thrust characteristics for timescales smaller that the control horizon $T_A = 120$ s remain unchanged, whereas the time synchronization of control actions to specific flow events is eliminated.
-Similar to the first case, this is done in 2 random ways, and Figure~\ref{fig:scrambled}b illustrates the row-averaged power for these cases.
+Similar to the first case, this is done in 2 random ways, and Fig.~\ref{fig:scrambled}b illustrates the row-averaged power for these cases.
 
@@
 The sheets destabilize and break up as they are advected downstream, resulting in complex three-dimensional vortical structures.
-Furthermore, as also shown in Figure~\ref{fig:vorticity_windfarm}c, wake mixing is limited, and downstream turbines experience reduced velocities.
+Furthermore, as also shown in Fig.~\ref{fig:vorticity_windfarm}c, wake mixing is limited, and downstream turbines experience reduced velocities.
 In contrast, the optimized case shows coherent vortex rings being shed from the first-row turbine.
-As indicated by the black arrows in Figure~\ref{fig:vorticity_windfarm}b, the locations of the rings in the controlled case coincide with naturally occurring bulges in the vortex sheet of the reference case: the controlled turbines further destabilize the sheet by well-timed temporal variations in its thrust coefficient.
+As indicated by the black arrows in Fig.~\ref{fig:vorticity_windfarm}b, the locations of the rings in the controlled case coincide with naturally occurring bulges in the vortex sheet of the reference case: the controlled turbines further destabilize the sheet by well-timed temporal variations in its thrust coefficient.
 Figure~\ref{fig:vorticity_windfarm}c shows that this results in smaller velocity deficits in the wake region.
@@
 However, the fact that we have only two parameters renders a parameter sweep computationally feasible.
-The sweep is performed for a reduced-size wind-farm LES, as illustrated in Figure~\ref{fig:sinus_setup}.
+The sweep is performed for a reduced-size wind-farm LES, as illustrated in Fig.~\ref{fig:sinus_setup}.
 The farm consists of $4 \times 4$ turbines in an aligned layout with $S = 6D$ in both streamwise and spanwise directions, geometrically equivalent to the optimally controlled wind farm in MM17.
@@
 Figure~\ref{fig:vorticity_parametersweep}b shows that, at the optimal frequency, this leads to increased wake mixing, providing the second-row turbine with a higher incoming velocity.
-In contrast, Figure~\ref{fig:vorticity_parametersweep}c shows that, even though higher frequency thrust oscillations also result in periodic shedding of vortex rings, this does not automatically lead to more favorable flow conditions for downstream turbines.
+In contrast, Fig.~\ref{fig:vorticity_parametersweep}c shows that, even though higher frequency thrust oscillations also result in periodic shedding of vortex rings, this does not automatically lead to more favorable flow conditions for downstream turbines.
 Therefore, it can be concluded that a correct timing and spacing of vortex rings is essential for increased wake mixing.
@@
 
-In order to assess whether the same strategy can be used in the downstream turbines as well, Figure~\ref{fig:sinus_row2} illustrates the results from an identical parameter sweep as discussed above, except that here the second turbine row is controlled using a sinusoidal thrust coefficient.
+In order to assess whether the same strategy can be used in the downstream turbines as well, Fig.~\ref{fig:sinus_row2} illustrates the results from an identical parameter sweep as discussed above, except that here the second turbine row is controlled using a sinusoidal thrust coefficient.
 Figure~\ref{fig:sinus_row2}a indicates that sinusoidal actuation of the second row invariably leads to losses in wind-farm power.
@@
 
-It is interesting to note that, even though the current parameter sweep is performed using different initial and inlet conditions than those applied in MM17, the optimal frequency of sinusoidal variations in $\cthat$ at $St^\star = f D / U_\infty = 0.25$ corresponds to the location of the peak in the first-row thrust coefficient spectrum of C3t5 in Figure~\ref{fig:spectra_thrust_cthat}.
+It is interesting to note that, even though the current parameter sweep is performed using different initial and inlet conditions than those applied in MM17, the optimal frequency of sinusoidal variations in $\cthat$ at $St^\star = f D / U_\infty = 0.25$ corresponds to the location of the peak in the first-row thrust coefficient spectrum of C3t5 in Fig.~\ref{fig:spectra_thrust_cthat}.
 In the following paragraphs, the robustness of the best parameter pair for first-row thrust variations, i.e.
@@
 Results are promising: for the given cases, $St^\star$ and $A^\star$  do not depend on streamwise turbine spacing.
-Furthermore, even in the $7D$ spacing case (Figure~\ref{fig:sinus_spacing}c-d), which naturally features lower overall power losses in downstream rows, power extraction in the second row can be significantly increased through sinusoidal variations in the first-row thrust coefficient.
+Furthermore, even in the $7D$ spacing case (Fig.~\ref{fig:sinus_spacing}c-d), which naturally features lower overall power losses in downstream rows, power extraction in the second row can be significantly increased through sinusoidal variations in the first-row thrust coefficient.
 \begin{figure}
@@
 
-The top panel of Figure~\ref{fig:cross_section_sinus} shows cross sections of time-averaged axial velocities $\widetilde{U}_x$ at the rotor disk locations for the reference case.
+The top panel of Fig.~\ref{fig:cross_section_sinus} shows cross sections of time-averaged axial velocities $\widetilde{U}_x$ at the rotor disk locations for the reference case.
 Further, the middle and bottom panel illustrate deviations from the reference velocity for the $(St^\star, A^\star)$ sinusoidal case and the optimal control case C3t5 respectively.
@@
 The current section discusses the intermediate rows.
-It was shown that, without active participation in these rows, upstream gains are lost in downstream rows, and only full optimal control succeeds in achieving significant gains in downstream rows as well (see Figures~\ref{fig:sinus_fullscale},~\ref{fig:cross_section_sinus}).
+It was shown that, without active participation in these rows, upstream gains are lost in downstream rows, and only full optimal control succeeds in achieving significant gains in downstream rows as well (see Figs.
+~\ref{fig:sinus_fullscale},~\ref{fig:cross_section_sinus}).
 It was already mentioned that the analysis and behavior of turbines within intermediate rows is more complex than in the first row: they aim to influence the flow to the benefit of downstream rows but are also dependent on the actions of upstream rows.
@@
 
-First, as shown in the top panels of Figure~\ref{fig:cross_section_sinus}, even in the uncontrolled case, the kinetic energy of the flow in the vicinity of the turbine rotor is depleted more and more in the downstream rows.
+First, as shown in the top panels of Fig.~\ref{fig:cross_section_sinus}, even in the uncontrolled case, the kinetic energy of the flow in the vicinity of the turbine rotor is depleted more and more in the downstream rows.
 This complicates control strategies for these rows as the opportunity for increased mixing with high energy flow is decreased.
-Furthermore, intermediate turbines are subjected to increased turbulence levels and more complex vorticity dynamics, as illustrated in Figure~\ref{fig:vorticity_windfarm}.
+Furthermore, intermediate turbines are subjected to increased turbulence levels and more complex vorticity dynamics, as illustrated in Fig.~\ref{fig:vorticity_windfarm}.
 This could explain why sinusoidal thrust control did not lead to increased power when applied to the second row: whereas the first row produces increased mixing by destabilizing relatively stable vortex sheets into vortex rings, the second row is already continuously immersed in complex vorticity patterns for which this simple approach does not work.
-However, note that for instance R7 in Figure~\ref{fig:controls} also seems to show quasi-periodic sinusoidal variations in $\cthat$ at a time period of approximately 50~s.
+However, note that for instance R7 in Fig.~\ref{fig:controls} also seems to show quasi-periodic sinusoidal variations in $\cthat$ at a time period of approximately 50~s.
 This is an indication that, also for intermediate rows, vortex ring shedding could amount to part of the power increase observed in the optimal control simulations, albeit at specific moments in time, synchronized with the local flow conditions.
@@
 Figure~\ref{fig:multirow} illustrates that the first-row optimized thrust coefficient also results in a significant power increase in the third row, which is not observed using the sinusoidal thrust strategy.
-Furthermore, the analysis of the modified control cases in Figure~\ref{fig:scrambled} proves that also the first-row controls are partially synchronized with the flow.
+Furthermore, the analysis of the modified control cases in Fig.~\ref{fig:scrambled} proves that also the first-row controls are partially synchronized with the flow.
 This shows that other mechanisms, dependent on specific flow events for increasing wind-farm power, are at play as well.

### 2018-02-14 draft.tex
Cleanup, figures numbered

@@
 
-\usepackage{tabularx}
 \usepackage{mathtools}
@@
 \usepackage{amsmath}
-%\usepackage[short]{optidef}
-\usepackage{xcolor}
 \usepackage{amsfonts}
-
-% Colors
-%---------------------------------------------------------------------------------
-\definecolor{C1}{RGB}{31,119,180}
-\definecolor{C2}{RGB}{255,127,14}
-\definecolor{C3}{RGB}{44,160,44}
-\definecolor{C4}{RGB}{214,39,40}
-
-% Some  auxiliary commands for ease of mathematics notations
-%---------------------------------------------------------------------------------
-% Derivatives and differentials
-\newcommand{\ds}{~\text{d}\boldsymbol{s}}
-\newcommand{\dzeta}{~\text{d}\boldsymbol{\zeta}}
-\newcommand{\pder}[2]{\frac{\partial #1}{\partial #2}}
-\newcommand{\bs}[1]{\boldsymbol{#1}}
-\newcommand{\dx}{\text{d}\boldsymbol{x}}
-\newcommand{\dt}{\text{d}t}
-\newcommand{\ddt}[1]{\frac{\text{d} #1}{\text{d} t}}
-% Integrals
-\newcommand{\stint}{\int_{0}^{T} \int_{\Omega}}
-\newcommand{\sint}{\int_{\Omega}}
-\newcommand{\Tint}{\int_{0}^{T}}
-\newcommand{\tint}{\int_{0}^{t}}
-% Filtered variables
-\newcommand{\utilde}{\widetilde{\bs{u}}}
-\newcommand{\ptilde}{\widetilde{p}}
-\newcommand{\ctnhat}[1]{\widehat{C}_{T,#1}'}
-\newcommand{\ctihat}{\widehat{C}_{T,i}'}
-\newcommand{\cthat}{\widehat{C}_T'}
-% Other symbols
-\newcommand{\ctmax}{C_{T,\text{max}}'}
-\newcommand{\cti}{C_{T,i}'}
-\newcommand{\ctn}[1]{C_{T,#1}'}
-\newcommand{\R}{\mathscr{R}}
-\newcommand{\J}{\mathscr{J}}
-\newcommand{\Jtilde}{\tilde{\mathscr{J}}}
-\newcommand{\Jgrad}{\nabla \Jtilde}
-\newcommand{\Lagr}{\mathscr{L}}
-\newcommand{\eperp}{\bs{e}_\perp}
-\newcommand{\eperpi}{\bs{e}_{\perp,i}}
-\newcommand{\etransi}{\bs{e}_{\parallel,i}}
-\newcommand{\ex}{\bs{e}_x}
-\newcommand{\ey}{\bs{e}_y}
-\newcommand{\ez}{\bs{e}_z}
-\newcommand{\eperpn}[1]{\bs{e}_{\perp,#1}}
-\newcommand{\vi}{\frac{1}{A_i} \sint \R_i (\bs{s})~\utilde \cdot \eperpi \ds}
-% Operators
-\newcommand{\innerproduct}[2]{\bigg( #1, #2 \bigg)}
-\newcommand{\innerproductsmall}[2]{\big( #1, #2 \big)}
-\newcommand{\sumturbines}{\sum_{i=1}^{N_t}}
-\newcommand{\diracdelta}{{\delta}}
-\DeclareMathOperator*{\argmin}{arg\,min}
-
-\newcommand{\red}[1]{{\color{red} #1}}
 
@@
 \ref{fig:optimization_meth}.
-Figure \ref{fig:optimization_meth}a illustrates the control block diagram: an iterative optimization loop updates the wind-farm control vector $\bs{\varphi}(t)$ until a set of optimized controls $\bs{\varphi}^{\bullet}(t)$ is found.
-This optimization is based upon an unsteady turbulence-resolving LES wind-farm flow model, and sensitivities of the cost functional $\J$ (i.e.
+Figure \ref{fig:optimization_meth}a illustrates the control block diagram: an iterative optimization loop updates the wind-farm control vector $\boldsymbol{\varphi}(t)$ until a set of optimized controls $\boldsymbol{\varphi}^{\bullet}(t)$ is found.
+This optimization is based upon an unsteady turbulence-resolving LES wind-farm flow model, and sensitivities of the cost functional $\mathscr{J}$ (i.e.
 the total wind-farm power extraction) are calculated using an adjoint formulation of this flow model.
@@
 \ref{fig:optimization_meth}b.
-In this framework, wind-farm controls $\bs{\varphi}(t)$ are optimized for a finite time horizon $T$, involving a sequential set of LES and adjoint LES simulations.
+In this framework, wind-farm controls $\boldsymbol{\varphi}(t)$ are optimized for a finite time horizon $T$, involving a sequential set of LES and adjoint LES simulations.
 Upon convergence of the optimization, optimized control signals are applied in a flow advancement simulation for a time $T_A < T$, after which a new optimization window is initiated.
@@
 	\caption{Schematic overview of wind-farm optimal control methodology from MM17.
-\emph{a)} Control block diagram with adjoint gradient-based optimization and LES flow models illustrating data flow of (optimal) controls $\bs{\varphi}^{(\bullet)}$, system state $\bs{q}$, cost functional $\J$ and its gradient $\nabla \J$.
+\emph{a)} Control block diagram with adjoint gradient-based optimization and LES flow models illustrating data flow of (optimal) controls $\boldsymbol{\varphi}^{(\bullet)}$, system state $\boldsymbol{q}$, cost functional $\mathscr{J}$ and its gradient $\nabla \mathscr{J}$.
 \emph{b)} Receding horizon framework subdividing time into discrete flow advancement windows of length $T_A$ with prediction horizon $T$.
 Each arrow represents a forward or adjoint LES.
-Every window consists of an optimization stage (blue and red lines) follow by a flow advancement stage with optimal controls $\bs{\varphi}^{\bullet}$ (green lines).
+Every window consists of an optimization stage (blue and red lines) follow by a flow advancement stage with optimal controls $\boldsymbol{\varphi}^{\bullet}$ (green lines).
 \label{fig:optimization_meth}}
@@
 \begin{alignat}{4}
-& \underset{\bs{\varphi}, \bs{q}}{\text{minimize}}  & \qquad  \J(\bs{\varphi}, \bs{q}) &= - \Tint \sum_{i=1}^{N_t} P_i ~\dt  & \label{eq:costfunction}\\
-& \text{s.t.}                      			&         \frac{\partial \utilde}{\partial t} + \big(\utilde \cdot \nabla \big)~ \utilde &= - \nabla (\ptilde + \ptilde_\infty) / \rho - \nabla \cdot \boldsymbol{\tau}_{sgs} + \sum_{i=1}^{N_t} \bs{f}_i \qquad  & \text{in } \Omega \times (0,T], \label{eq:NSmomentum_constraint} \\
-&                                                   &        \nabla \cdot \utilde&=0 									        & \text{in } \Omega \times (0,T], \label{eq:NScontinuity_constraint}\\
-&                                                   &        \tau \ddt{\ctihat}&=\cti - \ctihat 								& i=1\dots N_t~\text{in } (0,T],  \label{eq:ctihat_constraint}\\
-&                                                   &       0 \leq~ &\cti \leq C_{T,\text{max}}'				& i=1\dots N_t~\text{in } (0,T],  \label{eq:boxct_constraint}
+& \underset{\boldsymbol{\varphi}, \boldsymbol{q}}{\text{minimize}}  & \qquad  \mathscr{J}(\boldsymbol{\varphi}, \boldsymbol{q}) &= - \int_{0}^{T} \sum_{i=1}^{N_t} P_i ~\text{d}t  & \label{eq:costfunction}\\
+& \text{s.t.}                      			&         \frac{\partial \widetilde{\boldsymbol{u}}}{\partial t} + \big(\widetilde{\boldsymbol{u}} \cdot \nabla \big)~ \widetilde{\boldsymbol{u}} &= - \nabla (\widetilde{p} + \widetilde{p}_\infty) / \rho - \nabla \cdot \boldsymbol{\tau}_{sgs} + \sum_{i=1}^{N_t} \boldsymbol{f}_i \qquad  & \text{in } \Omega \times (0,T], \label{eq:NSmomentum_constraint} \\
+&                                                   &        \nabla \cdot \widetilde{\boldsymbol{u}}&=0 									        & \text{in } \Omega \times (0,T], \label{eq:NScontinuity_constraint}\\
+&                                                   &        \tau \frac{\text{d} \widehat{C}_{T,i}'}{\text{d} t} &=C_{T,i}' - \widehat{C}_{T,i}' 								& i=1\dots N_t~\text{in } (0,T],  \label{eq:ctihat_constraint}\\
+&                                                   &       0 \leq~ &C_{T,i}' \leq C_{T,\text{max}}'				& i=1\dots N_t~\text{in } (0,T],  \label{eq:boxct_constraint}
 \end{alignat}
@@
 The control variables are the time-dependent thrust coefficient setpoints $C_{T,i}'$ of every turbine $i$ ($=1 \dots N_t$), i.e.
-$\bs{\varphi} = [C_{T,1}'(t), \dots, C_{T,N_t}'(t) ]$, and the state variables are denoted as  $\bs{q}=[\utilde(\bs{x},t);~ \ptilde(\bs{x},t);~ \widehat{C}_{T,1}'(t), \dots, \widehat{C}_{T,N_t}'(t)]$, with $\utilde$ the filtered velocity, $\ptilde$ the filtered pressure, and $\widehat{C}_{T,i}'$ the (time-filtered) thrust coefficient for turbine $i$ (see below).
+$\boldsymbol{\varphi} = [C_{T,1}'(t), \dots, C_{T,N_t}'(t) ]$, and the state variables are denoted as  $\boldsymbol{q}=[\widetilde{\boldsymbol{u}}(\boldsymbol{x},t);~ \widetilde{p}(\boldsymbol{x},t);~ \widehat{C}_{T,1}'(t), \dots, \widehat{C}_{T,N_t}'(t)]$, with $\widetilde{\boldsymbol{u}}$ the filtered velocity, $\widetilde{p}$ the filtered pressure, and $\widehat{C}_{T,i}'$ the (time-filtered) thrust coefficient for turbine $i$ (see below).
 
@@
 %\begin{equation}
-%$	\bs{f}_i(\bs{x},t) = - \frac{1}{2}~ \widehat{C}_{T,i}'(t)~ V_i(t)^2~ \R_i (\bs{x})~ \eperpi,$
-$	\bs{f}_i(\bs{x},t) = - (1/2) \widehat{C}_{T,i}'(t) V_i(t)^2 \R_i (\bs{x}) \eperpi,$
+%$	\boldsymbol{f}_i(\boldsymbol{x},t) = - \frac{1}{2}~ \widehat{C}_{T,i}'(t)~ V_i(t)^2~ \R_i (\boldsymbol{x})~ \boldsymbol{e}_{\perp,i},$
+$	\boldsymbol{f}_i(\boldsymbol{x},t) = - (1/2) \widehat{C}_{T,i}'(t) V_i(t)^2 \mathscr{R}_i (\boldsymbol{x}) \boldsymbol{e}_{\perp,i},$
 %\end{equation}
-where $\R_i$ is a smoothed representation of the geometric footprint of the turbine on the LES grid and $\eperpi$ is the rotor-perpendicular vector.
-Further, the disk-averaged velocity is defined as $V_i = (1/A_i) \sint \R_i(\bs{x}) \utilde \cdot \eperpi ~\dx$, with $A_i$ the rotor disk area.
+where $\mathscr{R}_i$ is a smoothed representation of the geometric footprint of the turbine on the LES grid and $\boldsymbol{e}_{\perp,i}$ is the rotor-perpendicular vector.
+Further, the disk-averaged velocity is defined as $V_i = (1/A_i) \int_{\Omega} \mathscr{R}_i(\boldsymbol{x}) \widetilde{\boldsymbol{u}} \cdot \boldsymbol{e}_{\perp,i} ~\text{d}\boldsymbol{x}$, with $A_i$ the rotor disk area.
 Mechanical power captured by the wind turbine is calculated as $P_i = (1/2) C_{P,i}'(t) V_i(t)^3 A_i$, with $C_{P,i}' = 0.9 C_{T,i}'$, resulting from a fit of LES results to momentum theory, eliminating the overprediction of wind-turbine power on typical wind-farm LES grid resolutions \citep{munters2017optimal}.
@@
 \begin{figure}
-	\includegraphics[width=0.8\textwidth]{flowfield_MM17}
+	\includegraphics[width=0.8\textwidth]{figure2}
 	\caption{Instantaneous streamwise velocity $\widetilde{u}_x$ for the $12 \times 6$ aligned wind farm considered in MM17.
@@
 Several different optimal control cases were defined, based on the wind-turbine response time $\tau = $0, 5, or 30 s (instantaneous, fast, or slow response, see Eq.
-\ref{eq:ctihat_constraint}) and the maximal thrust coefficient $\ctmax = 2$ or 3 (underinductive or overinductive, see Eq.
+\ref{eq:ctihat_constraint}) and the maximal thrust coefficient $C_{T,\text{max}}' = 2$ or 3 (underinductive or overinductive, see Eq.
 \ref{eq:boxct_constraint}).
-Cases are denoted as C<X>t<Y>, where X and Y represent $\ctmax$ and $\tau$ respectively, e.g.
-C3t30  for the case with $\ctmax = 3$ and $\tau = 30$ s.
+Cases are denoted as C<X>t<Y>, where X and Y represent $C_{T,\text{max}}'$ and $\tau$ respectively, e.g.
+C3t30  for the case with $C_{T,\text{max}}' = 3$ and $\tau = 30$ s.
 
@@
 \begin{figure}
-	\includegraphics[width=\textwidth]{power_MM17}
+	\includegraphics[width=\textwidth]{figure3}
 	\caption{Energy extraction $E_O$ of optimally controlled wind-farm cases from MM17, normalized by a greedy reference case $E_R$.
@@
 \begin{figure}
-	\includegraphics[width=\textwidth]{statfig_initial.eps}
+	\includegraphics[width=\textwidth]{figure4}
 	\caption{Time-averaged flow field quantities of simulation results from MM17.
@@
 
-Figure~\ref{fig:controls} illustrates the time evolution of some of  the thrust coefficients $\cthat$ in the C3t5 case.
+Figure~\ref{fig:controls} illustrates the time evolution of some of  the thrust coefficients $\widehat{C}_{T}'$ in the C3t5 case.
 The figure shows that, for all rows but the last one (i.e.
-R12), $\cthat$ varies significantly in time, and that the amplitudes and frequencies of these variations are somewhat higher in the upstream rows of the farm.
-In contrast, row 12 features only minor unsteadiness at lower frequencies, and has an increased mean value of $\cthat$.
+R12), $\widehat{C}_{T}'$ varies significantly in time, and that the amplitudes and frequencies of these variations are somewhat higher in the upstream rows of the farm.
+In contrast, row 12 features only minor unsteadiness at lower frequencies, and has an increased mean value of $\widehat{C}_{T}'$.
 This relatively steady behavior of the last row can be explained by the fact that there are no further downstream turbines that can benefit from row 12 actively influencing local flow conditions, hence the row optimizes its own power only.
-The increase in mean $\cthat$ in row 12 can be explained based on Fig.~\ref{fig:ct_sweep}, which shows the power extraction as a function of steady $\cthat$ for unwaked turbines, subject to identical turbulent inlet as in case C3t5.
-Although momentum theory predicts maximal power extraction for steady uniform inflow at $\cthat = 2$, the actual optimal steady value for the ADM at the current spatial resolution lies somewhat higher at $\cthat \approx 2.4$, for which power extraction is about 1.4\% higher than at $\cthat = 2$.
-This behavior is related to the overprediction in ADM power due to the diffuse turbine representation on typical simulation grids: the mass flow through the rotor disk at $\cthat > 2$ is slightly too high compared to momentum theory, resulting in a  shift of optimal $\cthat$ towards somewhat higher values.
-Although the linear fit $C_P' = a \cthat$ introduced in \cite{munters2017optimal}, Appendix~A eliminates the error in maximal power extraction, it does not correct the value of the optimal $\cthat$ (note that this could be achieved through a more complex relation between $C_P'$ and $\cthat$).
+The increase in mean $\widehat{C}_{T}'$ in row 12 can be explained based on Fig.~\ref{fig:ct_sweep}, which shows the power extraction as a function of steady $\widehat{C}_{T}'$ for unwaked turbines, subject to identical turbulent inlet as in case C3t5.
+Although momentum theory predicts maximal power extraction for steady uniform inflow at $\widehat{C}_{T}' = 2$, the actual optimal steady value for the ADM at the current spatial resolution lies somewhat higher at $\widehat{C}_{T}' \approx 2.4$, for which power extraction is about 1.4\% higher than at $\widehat{C}_{T}' = 2$.
+This behavior is related to the overprediction in ADM power due to the diffuse turbine representation on typical simulation grids: the mass flow through the rotor disk at $\widehat{C}_{T}' > 2$ is slightly too high compared to momentum theory, resulting in a  shift of optimal $\widehat{C}_{T}'$ towards somewhat higher values.
+Although the linear fit $C_P' = a \widehat{C}_{T}'$ introduced in \cite{munters2017optimal}, Appendix~A eliminates the error in maximal power extraction, it does not correct the value of the optimal $\widehat{C}_{T}'$ (note that this could be achieved through a more complex relation between $C_P'$ and $\widehat{C}_{T}'$).
 Returning to the more complex thrust coefficients in the other rows it is worth noting that, based on the current dataset, no statistically significant correlations between thrust coefficients of different turbines could be found.
@@
 \begin{figure}
-	\includegraphics[width=0.8\textwidth]{controls.eps}
-	\caption{Time evolution of the thrust coefficient $\cthat$ for a selection of optimally controlled turbines  of case C3t5.
+	\includegraphics[width=0.8\textwidth]{figure5}
+	\caption{Time evolution of the thrust coefficient $\widehat{C}_{T}'$ for a selection of optimally controlled turbines  of case C3t5.
 \emph{Left: } Total time horizon.
@@
 \begin{figure}
-	\includegraphics[width=0.8\textwidth]{sweep_ct.eps}
-	\caption{Normalized power extraction as a function of steady thrust coefficient $\cthat$ for wind turbines subject to the same freestream turbulent inflow as in case C3t5.
+	\includegraphics[width=0.8\textwidth]{figure6}
+	\caption{Normalized power extraction as a function of steady thrust coefficient $\widehat{C}_{T}'$ for wind turbines subject to the same freestream turbulent inflow as in case C3t5.
 Every dot corresponds to one LES.\label{fig:ct_sweep}}
@@
 \begin{figure}
-	\includegraphics[width=0.8\textwidth]{spectra_thrust_cthat.eps}
-	\caption{Power spectral density (PSD) estimates of the row-averaged thrust force $f_T$ (\emph{a}) and thrust coefficient $\cthat$ (\emph{b}) as a function of frequency $f$ (bottom axis) and non-dimensional Strouhal number $St$ (top axis).
+	\includegraphics[width=0.8\textwidth]{figure7}
+	\caption{Power spectral density (PSD) estimates of the row-averaged thrust force $f_T$ (\emph{a}) and thrust coefficient $\widehat{C}_{T}'$ (\emph{b}) as a function of frequency $f$ (bottom axis) and non-dimensional Strouhal number $St$ (top axis).
 \label{fig:spectra_thrust_cthat}}
@@
 Further downstream, rows 2 to 11 have very similar spectral behavior, and row 12 shows significantly lower variability.
-The high-frequency slopes of around $-5$ observed both for $f_T$ and $\cthat$ indicate that force variability on short timescales is caused mainly by thrust coefficient variations, whereas the slower thrust force dynamics tend more to a $-5/3$ slope, suggesting that these are governed by the unsteadiness in the turbulence instead.
+The high-frequency slopes of around $-5$ observed both for $f_T$ and $\widehat{C}_{T}'$ indicate that force variability on short timescales is caused mainly by thrust coefficient variations, whereas the slower thrust force dynamics tend more to a $-5/3$ slope, suggesting that these are governed by the unsteadiness in the turbulence instead.
 Note that, even though the spectra for all rows except row 12 collapse at frequencies below $0.05$ Hz, the first-row spectrum shows a small peak at $f \approx 0.02 \dots 0.03$ Hz ($f D / U_\infty \approx 0.2 \dots 0.3$) as indicated by the purple arrow.
@@
 In order to further study how the optimal controls increase overall wind-farm power, Fig.~\ref{fig:multirow} shows power extraction resulting from applying a subset of the optimal controls to specific turbine rows only.
-Figure~\ref{fig:multirow}a depicts simulation results for which the  optimized controls are applied \emph{only to one specific row}, with the thrust coefficient in all other rows kept at the reference value of $\cthat = 2$.
+Figure~\ref{fig:multirow}a depicts simulation results for which the  optimized controls are applied \emph{only to one specific row}, with the thrust coefficient in all other rows kept at the reference value of $\widehat{C}_{T}' = 2$.
 From the figure it can be seen that only for the controls of the first row (R1) this results in a significant power increase in rows 2 and 3.
@@
 \begin{figure}
-	\includegraphics[width=\textwidth]{multirow.eps}
+	\includegraphics[width=\textwidth]{figure8}
 	\caption{Normalized row-averaged power extraction for the reference case, optimal control case C3t5, and subset control cases up to optimization window 3.
@@
 	\centering
-	\includegraphics[width=0.8\textwidth]{single_row_opt.eps}
+	\includegraphics[width=0.8\textwidth]{figure9}
 	\caption{Increase in wind-farm power extraction for the first (and only) timewindow of the optimal control cases in which only a single row is optimized. \label{fig:single_row_opt}}
@@
 	\centering
-	\includegraphics[width=0.75\textwidth]{individual_gains.eps}
+	\includegraphics[width=0.75\textwidth]{figure10}
 	\caption{Relative power increase matrix for downstream rows in the single-row optimization case for row R$_{\rm i}$, indicated in the horizontal axis.
@@
 \begin{figure}
-	\includegraphics[width=\textwidth]{scrambled.eps}
+	\includegraphics[width=\textwidth]{figure11}
 	\caption{Normalized row-averaged power extraction for the reference case, optimal control case C3t5, and modified control cases.
@@
 	\centering
-	\includegraphics[width=\textwidth]{vorticity.eps}
+	\includegraphics[width=\textwidth]{figure12}
 	\caption{Instantaneous snapshots at t = 300 s of portion of wind-farm flow field for the reference (left) and optimized (right) case.
@@
 The aim of the current section is to mimic the quasi-periodic shedding of vortex rings by upstream turbines as observed above through the use of simple periodic variations in the thrust coefficient.
-Instead of optimizing a high-dimensional control signal that can evolve freely in time as in MM17, we impose a sinusoidal perturbation on the Betz-optimal coefficient $\cthat = 2$, parametrized by its amplitude $A$, and its frequency in the form of a non-dimensional Strouhal number $St = f D/ U_\infty$, with $f$ the dimensional frequency, $D$ the turbine diameter, and $U_\infty$ the unperturbed time-averaged upstream velocity:
+Instead of optimizing a high-dimensional control signal that can evolve freely in time as in MM17, we impose a sinusoidal perturbation on the Betz-optimal coefficient $\widehat{C}_{T}' = 2$, parametrized by its amplitude $A$, and its frequency in the form of a non-dimensional Strouhal number $St = f D/ U_\infty$, with $f$ the dimensional frequency, $D$ the turbine diameter, and $U_\infty$ the unperturbed time-averaged upstream velocity:
 \begin{equation}
-\cthat(t) = 2 + A \sin \bigg(2\pi St \frac{t U_\infty}{D} \bigg).\label{eq:define_thrust}
+\widehat{C}_{T}'(t) = 2 + A \sin \bigg(2\pi St \frac{t U_\infty}{D} \bigg).\label{eq:define_thrust}
 \end{equation}
@@
 A wall roughness length $z_0 = 10^{-1}$ m is used.
-A set of wind-farm flow simulations is advanced in time by 30~minutes, during which the front row is controlled using a sinusoidally varying thrust coefficient $\cthat$, as defined in Eq.
+A set of wind-farm flow simulations is advanced in time by 30~minutes, during which the front row is controlled using a sinusoidally varying thrust coefficient $\widehat{C}_{T}'$, as defined in Eq.
 \eqref{eq:define_thrust}.
@@
 	\centering
-	\includegraphics[width=0.7\textwidth]{setup_sinusoidal.eps}
+	\includegraphics[width=0.7\textwidth]{figure13}
 	\caption{Reduced $4 \times 4$ wind-farm simulation setup for sinusoidal variation parameter study showing instantaneous contours of streamwise velocity $\widetilde{u}_x$.
@@
 	\centering
-	\includegraphics[width=\textwidth]{gains_turbulent_6D_6D_6D_wide_inphase2.eps}
+	\includegraphics[width=\textwidth]{figure14}
 	\caption{Power extraction of baseline sinusoidal thrust case ($S = 6D$, $z_0 = 10^{-1}$ m).
@@
 	\centering
-	\includegraphics[width=\textwidth]{pumpcases3.eps}
+	\includegraphics[width=\textwidth]{figure15}
 	\caption{Snapshots of wind-farm flow fields at t = 1 800 s.
@@
 	\centering
-	\includegraphics[width=\textwidth]{gains_row22.eps}
+	\includegraphics[width=\textwidth]{figure16}
 	\caption{Power extraction of second-row sinusoidal thrust case .
@@
 
-It is interesting to note that, even though the current parameter sweep is performed using different initial and inlet conditions than those applied in MM17, the optimal frequency of sinusoidal variations in $\cthat$ at $St^\star = f D / U_\infty = 0.25$ corresponds to the location of the peak in the first-row thrust coefficient spectrum of C3t5 in Fig.~\ref{fig:spectra_thrust_cthat}.
+It is interesting to note that, even though the current parameter sweep is performed using different initial and inlet conditions than those applied in MM17, the optimal frequency of sinusoidal variations in $\widehat{C}_{T}'$ at $St^\star = f D / U_\infty = 0.25$ corresponds to the location of the peak in the first-row thrust coefficient spectrum of C3t5 in Fig.~\ref{fig:spectra_thrust_cthat}.
 In the following paragraphs, the robustness of the best parameter pair for first-row thrust variations, i.e.
@@
 	\centering
-	\includegraphics[width=\textwidth]{gains_combined2.eps}
+	\includegraphics[width=\textwidth]{figure17}
 	\caption{Power extraction of sinusoidal thrust cases with varying streamwise spacing.
@@
 	\centering
-	\includegraphics[width=\textwidth]{gains_turbulent_6D_6D_6D_wide_antiphase_z0_1m2.eps}
+	\includegraphics[width=\textwidth]{figure18}
 	\caption{Power extraction of sinusoidal thrust case with increased wall roughness $z_0 = 1$ m.
@@
 	\centering
-	\includegraphics[width=\textwidth]{gains_fullscale2.eps}
+	\includegraphics[width=\textwidth]{figure19}
 	\caption{Power extraction of full-scale sinusoidal thrust case ($S = 6D$, $z_0 = 10^{-1}$ m).
 \emph{a) } Relative gain in mean wind-farm power extraction over reference case as a function of sine amplitude $A$ and frequency $St$.
-\emph{b) } Row-averaged mean power extraction for the best sinusoidal case, the optimal control case C3t5 from MM17, and the reference case, normalized by first-row reference power.\label{fig:sinus_fullscale} }
+\emph{b) } Row-averaged mean power extraction for the sinusoidal case with parameters from previous sections, the optimal control case C3t5 from MM17, and the reference case, normalized by first-row reference power.\label{fig:sinus_fullscale} }
 \end{figure}
@@
 	\centering
-	\includegraphics[width=\textwidth]{frontviews_allc3t5.eps}
+	\includegraphics[width=\textwidth]{figure20}
 	\caption{Cross section of time-averaged axial velocity $\widetilde{U}_x$ at rotor locations in rows 1, 2, 3, 4, 5, and 12.
@@
 This could explain why sinusoidal thrust control did not lead to increased power when applied to the second row: whereas the first row produces increased mixing by destabilizing relatively stable vortex sheets into vortex rings, the second row is already continuously immersed in complex vorticity patterns for which this simple approach does not work.
-However, note that for instance R7 in Fig.~\ref{fig:controls} also seems to show quasi-periodic sinusoidal variations in $\cthat$ at a time period of approximately 50~s.
+However, note that for instance R7 in Fig.~\ref{fig:controls} also seems to show quasi-periodic sinusoidal variations in $\widehat{C}_{T}'$ at a time period of approximately 50~s.
 This is an indication that, also for intermediate rows, vortex ring shedding could amount to part of the power increase observed in the optimal control simulations, albeit at specific moments in time, synchronized with the local flow conditions.

### 2018-02-14 getfiles.sh

@@ -1,15 +1,0 @@
-cp -v ../induction_control_wes/sweep_ct.eps .
-cp -v ../induction_control_wes/spectra_thrust_cthat.eps .
-cp -v ../induction_control_wes/multirow.eps .
-cp -v ../induction_control_wes/single_row_opt.eps .
-cp -v ../induction_control_wes/individual_gains.eps .
-cp -v ../induction_control_wes/fullfarm_singlerow.eps .
-cp -v ../induction_control_wes/scrambled.eps .
-cp -v ../induction_control_wes/vorticity.eps .
-cp -v ../induction_control_wes/setup_sinusoidal.eps .
-cp -v ../induction_control_wes/gains_turbulent_6D_6D_6D_wide_inphase2.eps .
-cp -v ../induction_control_wes/gains_turbulent_6D_6D_6D_wide_antiphase_z0_1m2.eps .
-cp -v ../induction_control_wes/pumpcases2.eps .
-cp -v ../induction_control_wes/gains_row22.eps .
-cp -v ../induction_control_wes/gains_fullscale.eps .
-cp -v ../induction_control_wes/frontviews_allc3t5.eps .

### 2018-02-14 draft.tex
Small aesthetics fix

@@
 \ref{fig:flowfield_MM17}.
-The wind farm was controlled over a total of $N_A = 15$ time windows with a prediction horizon $T = 240 s$ and a flow advancement time of $T_A = T/2 = 120 s$, resulting in a total control time $T_{\rm tot} = N_A T_A = 30$ minutes.
+The wind farm was controlled over a total of $N_A = 15$ time windows with a prediction horizon $T = 240$ s and a flow advancement time of $T_A = T/2 = 120$ s, resulting in a total control time $T_{\rm tot} = N_A T_A = 30$ minutes.
 
@@
 A conventionally (greedily) controlled wind farm with steady $C_T' = 2$ (corresponding to the Betz limit in momentum theory) was defined as a reference case.
-Several different optimal control cases were defined, based on the wind-turbine response time $\tau = $0, 5, or 30 s (instantaneous, fast, or slow response, see Eq.
-\ref{eq:ctihat_constraint}) and the maximal thrust coefficient $C_{T,\text{max}}' = 2$ or 3 (underinductive or overinductive, see Eq.
+Several different optimal control cases were defined, based on the wind-turbine response time $\tau = $ 0, 5, or 30 s (instantaneous, fast, or slow response, see Eq.
+\ref{eq:ctihat_constraint}) and the maximal thrust coefficient $C_{T,\text{max}}' = 2$ or~3 (underinductive or overinductive, see Eq.
 \ref{eq:boxct_constraint}).

### 2018-04-17 revision/reply_to_reviewers.tex
Started revision process

\documentclass[]{article}

\setlength{\parindent}{0em}
\setlength{\parskip}{1em}

\begin{document}

%\maketitle

\section*{Response to Reviewer 1}

We thank the reviewer for his/her constructive comments and are pleased with the positive assessment or our work. We have addressed the specific comments made by the reviewer as described below. We hope that the revised manuscript can now be accepted for publication.

\dotfill

\textbf{1. Reviewer:} \textit{One comparison that came to mind reading the paper, is that the results show very little value in optimizing the most-downstream turbines, and in general, improvement in power comes from modifications upstream, and that self-optimization is not possible. This stands in contrast to a result such as:}

\textit{Ciri, Umberto, Mario Rotea, Christian Santoni, and Stefano Leonardi. “Large Eddy Simulation for an Array of Turbines with Extremum Seeking Control.” In American Control Conference. Boston, MA, 2016. }

\textit{Where the TSR of downstream turbines is re-optimized for wake conditions (and the upstream turbine is left as is at the end of the optimization). It would seem the difference in modeling methods and/or how turbine control is implemented yields the different results, but I believe it would be worth discussing the difference, for example around the paragraph beginning with "The figure shows that the first row (R1)..." on page 12.}

\textbf{Response:} The reason for this different behavior of the downstream vs upstream turbines is twofold:
\begin{enumerate}
	\item The Extremum Seeking Control (ESC) approach by Ciri \emph{et al.} differs fundamentally from our approach. In our approach, we simultaneously optimize all control variables over a given time horizon, which allows for full dynamic cooperation and interaction between all turbines. In the ESC, controls are optimized one by one in the sense that ESC is initiated on the first row turbine and will be activated on the subsequent row only after its upstream neighbor has converged to a steady state solution. In this way, dynamic control and interaction between controls in different turbines are discarded a priori. Furthermore, although this is not an inherent limitation of the ESC approach, it seems that Ciri \emph{et al.} optimize for individual turbine power instead of total wind-farm power (see Eq. 10 $P_i$ is used instead of $\sum_i P_i$). Note that in Santoni \emph{et al.} (from the same research group as Ciri \emph{et al.}), where a Jensen model is used to optimize total wind-farm power using the individual tip speed ratios, it is also found that after full-farm optimization the downstream turbine operates at its individually optimal TSR (consistent with our observations).

	\item In our modeling approach, we make abstraction of detailed turbine dynamics in the sense that we directly control the thrust coefficient $C_T'$ instead of the generator torque or tip speed ratio. In this sense, the optimal thrust coefficient value of a turbine with no downstream neighbors can be analytically found to be $C_T' = 2$. In Ciri \emph{et al.}, the authors indicate that the reason why the TSR in the downstream rows (around 6.5) does not correspond to the individual optimal TSR (7.5) of individual turbines is that the TSR depends on a reference velocity and (direct quotation) ``As a consequence, the TSR is not a meaningful indicated of a waked turbine operating condition".

\end{enumerate}

Summarizing, the inverse behavior of the control dynamics compared to Ciri \emph{et al.} can be attributed to the difference in approach to wind-farm control (dynamic full-farm optimization vs steady one-by-one optimization) as well as the implementation of the turbine control ($C_T'$ vs $\lambda$). We believe that, if we were to constrain ourselves to steady control and would control a similarly defined $\lambda$ as in Ciri \emph{et al.}, our observations would be similar to the latter study. Note that Ciri \emph{et al.} do not increase wind-farm power over the case with turbines operating at locally optimal TSR, whereas our dynamic approach allows to improve power significantly over the case with locally optimal operation.

In our opinion, the current study (full-farm dynamic wind-farm optimization) and that of Ciri \emph{et al.} (providing a practical ESC method for optimizing individual turbines) cannot be directly compared. However, we understand the contrast in the results and therefore added a small comment to the revised manuscript (WM - WEET NIET OF WE DIT WEL ECHT MOETEN DOEN, COMMENTS KOMT WAT VERLOREN UIT IN MANUSCRIPT, MISSCHIEN GEWOON REVIEWER OVERTUIGEN DAT WE DIT NIET INCLUDEN?) (l. ??, p. ??):

``
... is not monotonous. Note that these observations are in contrast to the study of Ciri \emph{et al.} (REF), where Extremum Seeking Control is applied to optimize an array of three turbines and it was found that the set points in the first-row turbines remain untouched from their locally optimal values whereas downstream turbines are re-optimized for wake conditions. This contrast can mainly be attributed to our dynamic control approach, which allows for proper improvement of downstream flow conditions (which is not the case for set-point control as in Ciri \emph{et al.}), and that our ADM makes abstraction of any re-optimization of TSR.
''

WM - KAN WEL GOED ZIJN DAT ROTEA HIER DE REVIEWER IS - ZIEN DAT WE NIETS FOUT ZEGGEN. ANTWOORD IS NOG NIET ZO SCHERP ALS IK HET ZELF WIL. VOORAL IN ENUMERATE NOG WAT TE KNIPPEN

\dotfill

\textbf{2. Reviewer:} \textit{Figure 3/related text: Would another way to describe Ct2 vs Ct3 be that Ct2 can only lower the thrust, while Ct3 is allowed to raise it?}

\textbf{Response:} Indeed, this is correct. We have mentioned this explicitly in the revised manuscript as follows (p. ??, l. ??):

``
... and the maximal thrust coefficient $C_{T,\rm max}' = 2$ or 3, \textbf{with thrust forces that can respectively only be reduced (underinductive), or also increased (overinductive) compared to the Betz optimum at $C_T' = 2$} (see Eq. 5). 
''

\dotfill

\textbf{3. Reviewer:} \textit{``... NREL 5MW rotor with a 50\% increase in chord length ...'' does this imply the method is currently assuming the chord length is variable? Could this not be achieved by a change in pitch angle?}

\textbf{Response:} No, the method does not imply a variable chord length.

Current turbines are designed to approach maximum $C_T'$ values around 2, corresponding to the Betz limit. Although this maximum value can be increased by increasing the operational tip speed ratio (TSR), aiming to do so by adapting the pitch angle would inevitably lead to severe efficiency losses due to stall on the turbine blades. Therefore, we provide an example of how an alternative turbine design (i.e. with an increased chord length and operational TSR) could attain a maximum $C_T'$ of 3.5. Given such a turbine design, achieving thrust ratings $C_T' < 3.5$ is straightforward by pitching blades towards the feather position.

We have slightly modified the statement in the revised manuscript to avoid any confusion with regard to possibly having the chord length as a control variable. (p. ??, l. ??):
``
be \textbf{attained by the NREL 5MW turbine with a slight modification in the rotor design}, i.e. a 50\% increase in blade chord length and an operational tip speed ratio 25\% higher than 
the original design value (see Goit and Meyers, 2015, Appendix A)
''

\clearpage
\section*{Response to Reviewer 2}

\end{document}

### 2018-04-17 revision/reply_to_reviewers.tex
Added Rev2 comments

@@
 \section*{Response to Reviewer 2}
+We thank reviewer 2 for reading our work and for providing detailed feedback which has definitely improved the quality of the manuscript.
+We are pleased with his/her positive assessment of our research and have addressed the specific comments in the manuscript as described below.
+We hope that the revised manuscript can now be accepted for publication.
+
+\dotfill
+
+\textbf{Reviewer: } \textit{...LES setup is described/illustrated properly - except of the characteristics of the turbines/actuator disks considered.
+Would be nice to have the size of the disks explicitly noted (as they are somewhat hidden in Figure 2) to have a more clear scale of the considered wind farm.}
+
+\textbf{Response: } Indeed, we seem to have overlooked specifying the exact turbine dimensions in Section 2.2. We have updated the manuscript as follows:
+
+``
+... 12 rows by 6 columns. \textbf{The wind turbines have a hub height $z_h = 100$~m with a rotor diameter $D = 100$~m, and are spaced 6$D$ apart in both axial and transversal directions.}
+''
+
+\dotfill
+
+\textbf{Reviewer: } \textit{While describing the case setup in Section 2.2, the "flow advancement time", $T_A$ (also
+	referred in Figure 1) is considered as half of the prediction horizon $T$. Would $T_A$ (and
+	therefore $T$) be inflow dependent as the time delay (the time it takes for particles to
+	move from the upstream to downstream turbine(s))? Have you investigated if changing
+	$T$ (and/or $T_A$) has any effects on the resulting optimum $C_T$ set-points and on the power
+	gain?}
+
+\textbf{Response: }
+
+\dotfill
+
+\textbf{Reviewer: } \textit{As clearly seen in Figure 4c and 4d, there is a significant increase in turbulence
+	further downstream. In addition to the TKE and the transport, would be nice to have the
+	turbulence intensity TI values (as listed later on page 20, 10% for the baseline case),
+	both for the baseline case and the maximum added TI reported - possibly somewhere
+	around Figure 4. That again would give an indication on the applicability compared to
+	the field values observed. Also note the typo in the caption of Figure 4: after c) all the
+	subplots are marked to be continuously c).}
+
+\textbf{Response: }
+
+\dotfill
+
+\textbf{Reviewer: } \textit{On page 10, around line 10, the argument of "upstream actions do not require a
+	specific downstream response in order to increase power in that downstream row",
+	which is also paraphrased in the conclusions, needs to be elaborated. This rather
+	broad conclusion seem to oversee the probability of the curtailment of the downstream
+	turbine where down-regulation might be inevitable for certain CT set-points assigned
+	to downstream turbine(s) in the resulting optimization. Could be partially true for the
+	investigated C3t5 case since there observed very limited curtailment even at the most
+	upstream turbine (as in Figure 3b). However, also seen in Figure 8b (except of the
+	very last row as the authors indicated), there seem to be still a difference between on
+	the power gain at turbine R11 for the scenarios of R1-R10 and R1-R11. Narrowing the argument to the considered case or very little to no downstream curtailment CT
+	distributions is suggested.}
+
+\textbf{Response: }
+
+\dotfill
+
+\textbf{Reviewer: } \textit{On page 13, line 13, "the presence of the flow invariant features of the control signals"
+	needs further justification as Figure 11 would also depend on how variant the flow
+	features are in the simulations. That should include both the spatial and temporal
+	variance within the 30-min window. As far as the field measurements are concerned,
+	high spatial and temporal correlations are observed. For the former, Figure 2 gives a
+	brief idea about the wind speed range between the columns, that can be referred here.
+	For the latter, time series or relevant temporal statistics can be presented to assess the
+	randomness and strengthen the hypothesis.}
+
+\textbf{Response: }
+
+\dotfill
+
+\textbf{Reviewer: } \textit{On page 17, around line 5, a very nice example on how to implement the optimized
+	sinusoidal CT is presented. The practical examples can be further improved by a
+	short discussion on the expected response time of such increases in tip speed ratio
+	on a machine with high inertia. That would put the estimated sine wave period into
+	perspective as well.}
+
+\textbf{Response: }
+
+\dotfill
+
+\textbf{Reviewer: } \textit{For Section 4.2.3, the header "Full-scale wind farm test" is a bit misleading... Suggest
+	to change to "Full-scale wind farm simulations (in LES)" instead.}
+
+\textbf{Response: }
+
+\dotfill
+
+\textbf{Reviewer: } \textit{On Figure 19, why would the power decrease after Row 5 for the sinusoidal case?}
+
+\textbf{Response: }
+
+\dotfill
+
+\textbf{Reviewer: } \textit{Page 22 around line 5, the (inevitable) discussions on loads are included. In addition
+	to the loads on the controlled upstream turbine, Figure 20(b) indicates partial wakes
+	on the further downstream rows of turbines. Therefore, the section should be improved
+	by highlighting the probable increase in fatigue loading for not just turbine(s) R1 but for
+	the downstream rows as well, possibly starting as early as R3.}
+
+\textbf{Response: }
+
+\dotfill
+
+\textbf{Reviewer: } \textit{On the grammatical note, the manuscript is clear and easy to follow. The only comment
+	might be on the use of Sect. or Section; Fig. or Figure references.}
+
+\textbf{Response: }
 

### 2018-05-17 revision/reply_to_revs_2.tex
First draft of replies

\documentclass[]{article}

\usepackage{graphicx}

\setlength{\parindent}{0em}
\setlength{\parskip}{1em}

\begin{document}

%\maketitle

\section*{Response to Reviewer 1}

We thank the reviewer for his/her constructive comments and are pleased with the positive assessment or our work. We have addressed the specific comments made by the reviewer as described below. We hope that the revised manuscript can now be accepted for publication.

\dotfill

\textbf{1. Reviewer:} \textit{One comparison that came to mind reading the paper, is that the results show very little value in optimizing the most-downstream turbines, and in general, improvement in power comes from modifications upstream, and that self-optimization is not possible. This stands in contrast to a result such as:}

\textit{Ciri, Umberto, Mario Rotea, Christian Santoni, and Stefano Leonardi. “Large Eddy Simulation for an Array of Turbines with Extremum Seeking Control.” In American Control Conference. Boston, MA, 2016. }

\textit{Where the TSR of downstream turbines is re-optimized for wake conditions (and the upstream turbine is left as is at the end of the optimization). It would seem the difference in modeling methods and/or how turbine control is implemented yields the different results, but I believe it would be worth discussing the difference, for example around the paragraph beginning with "The figure shows that the first row (R1)..." on page 12.}

\textbf{Response:} The reason for this different behavior of the downstream vs upstream turbines is twofold:
\begin{enumerate}
	\item The Extremum Seeking Control (ESC) approach by Ciri \emph{et al.} differs fundamentally from our approach. In our approach, we simultaneously optimize all control variables over a given time horizon, which allows for full dynamic cooperation and interaction between all turbines. In the ESC, controls are optimized one by one in the sense that ESC is initiated on the first row turbine and will be activated on the subsequent row only after its upstream neighbor has converged to a steady state solution. In this way, dynamic control and interaction between controls in different turbines are discarded a priori. Furthermore, although this is not an inherent limitation of the ESC approach, Ciri \emph{et al.} optimize for individual turbine power instead of total wind-farm power (see Eq. 10 $P_i$ is used instead of $\sum_i P_i$). This is explicitly mentioned in the title of their follow up paper (Ciri \emph{et al.} Large-eddy simulations with extremum-seeking control for individual wind turbine power optimization)

	\item In our modeling approach, we make abstraction of detailed turbine dynamics in the sense that we directly control the thrust coefficient $C_T'$ instead of the generator torque or tip speed ratio. In this sense, the optimal thrust coefficient value of a turbine with no downstream neighbors can be analytically found to be $C_T' = 2$. In Ciri \emph{et al.}, the authors indicate that the reason why the TSR in the downstream rows (around 6.5) does not correspond to the individual optimal TSR (7.5) of individual turbines is that the TSR depends on a reference velocity and (direct quotation) ``As a consequence, the TSR is not a meaningful indicated of a waked turbine operating condition".

\end{enumerate}

Summarizing, the inverse behavior of the control dynamics compared to Ciri \emph{et al.} can be attributed to the difference in approach to wind-farm control (dynamic full-farm optimization vs steady sequential and individual optimization) as well as the implementation of the turbine control ($C_T'$ vs $\lambda$). In our opinion, the current study (full-farm dynamic wind-farm optimization) and that of Ciri \emph{et al.} (providing a practical ESC method for optimizing individual turbines) cannot be directly compared.

However, we understand the contrast in the results and therefore added a small comment to the revised manuscript  (l. ??, p. ??):

``
... is not monotonous. \textbf{Note that these observations are in contrast to the study of Ciri et al. 2017, where extremum seeking control is applied to optimize individual turbines in an array of three turbines. They found that set points in the first-row turbines remain untouched from their locally optimal values whereas downstream turbines are re-optimized for wake conditions. This contrast can be attributed to our dynamic and simultaneous wind-farm control approach, which differs from their sequential individual set-point control.  Furthermore, self-optimization in the latter study is achieved by recalibrating the torque gain. The baseline controller in our current actuator disk model makes abstraction of pitch and generator torque control and keeps $C_T'$ at its theoretical optimum instead.}
''

WM - WEET NIET OF WE DIT WEL ECHT MOETEN DOEN, MISSCHIEN GEWOON REVIEWER OVERTUIGEN DAT WE DIT NIET INCLUDEN? TE BESPREKEN

\dotfill

\textbf{2. Reviewer:} \textit{Figure 3/related text: Would another way to describe Ct2 vs Ct3 be that Ct2 can only lower the thrust, while Ct3 is allowed to raise it?}

\textbf{Response:} Indeed, this is correct. We have mentioned this explicitly in the revised manuscript as follows (p. ??, l. ??):

``
... and the maximal thrust coefficient $C_{T,\rm max}' = 2$ or 3, \textbf{with thrust forces that can respectively only be reduced (underinductive), or also increased (overinductive) compared to the Betz optimum at $C_T' = 2$} (see Eq. 5). 
''

\dotfill

\textbf{3. Reviewer:} \textit{``... NREL 5MW rotor with a 50\% increase in chord length ...'' does this imply the method is currently assuming the chord length is variable? Could this not be achieved by a change in pitch angle?}

\textbf{Response:} No, the method does not imply a variable chord length.

Current turbines are designed to approach maximum $C_T'$ values around 2, corresponding to the Betz limit. Although this maximum value can be increased by increasing the operational tip speed ratio (TSR), aiming to do so by adapting the pitch angle would inevitably lead to severe efficiency losses due to stall on the turbine blades. Therefore, we provide an example of how an alternative turbine design (i.e. with an increased chord length and operational TSR) could attain a maximum $C_T'$ of 3.5. Given such a turbine design, achieving thrust ratings $C_T' < 3.5$ is straightforward by pitching blades towards the feather position.

We have slightly modified the statement in the revised manuscript to avoid any confusion with regard to possibly having the chord length as a control variable. (p. ??, l. ??):

``
... be \textbf{attained by the NREL 5MW turbine with a slight modification in the rotor design}, i.e. a 50\% increase in blade chord length and an operational tip speed ratio 25\% higher than 
the original design value (see Goit and Meyers, 2015, Appendix A). \textbf{Furthermore, given this redesign, dynamic reductions from this maximum value could be realized through pitch control. More specifically, based on speculated specifications of similar turbines in practice, the NREL 5 MW is designed with a pitch rate of $8^{\circ}/$s. This implies that the rotor can go from power maximization ($0^\circ$ pitch) to total curtailment in fully feathered blade position ($90^\circ$ pitch) in 11.25 seconds (Jonkman et al. 2009).}
''

\clearpage
\section*{Response to Reviewer 2}
We thank reviewer 2 for reading our work and for providing detailed feedback which has improved the quality of the manuscript. We are pleased with his/her positive assessment of our research and have addressed the specific comments in the manuscript as described below. We hope that the revised manuscript can now be accepted for publication.

\dotfill

\textbf{Reviewer: } \textit{...LES setup is described/illustrated properly - except of the characteristics of the turbines/actuator disks considered. Would be nice to have the size of the disks explicitly noted (as they are somewhat hidden in Figure 2) to have a more clear scale of the considered wind farm.}

\textbf{Response: } Indeed, we seem to have overlooked specifying the exact turbine dimensions in Section 2.2. We have updated the manuscript as follows (l XX p YY):

``
... 12 rows by 6 columns. \textbf{The wind turbines have a hub height $z_h = 100$~m with a rotor diameter $D = 100$~m, and are spaced apart by 6$D$ in both axial and transversal directions.}
''

\dotfill

\textbf{Reviewer: } \textit{While describing the case setup in Section 2.2, the "flow advancement time", $T_A$ (also
	referred in Figure 1) is considered as half of the prediction horizon $T$. Would $T_A$ (and
	therefore $T$) be inflow dependent as the time delay (the time it takes for particles to
	move from the upstream to downstream turbine(s))? Have you investigated if changing
	$T$ (and/or $T_A$) has any effects on the resulting optimum $C_T$ set-points and on the power
	gain?}

\textbf{Response: } Indeed, $T$ and $T_A$ are important parameters in the receding-horizon approach. The optimization horizon $T$ can indeed be considered inflow-dependent. In our case, it is chosen as the time it takes for the flow to pass approximately four rows of turbines. In theory, it is desirable to have $T$ cover an entire wind-farm throughflow. In practice however, control over long time horizons is complicated by the chaotic nature of turbulence accompanied by adjoint gradient inaccuracies, and we choose $T$ as long as possible for our optimization to be unharmed by these inaccuracies (i.e. 240 s). The sensitivity of the power gains to the time horizon has not been formally quantified, yet it is reasonable to expect that the potential for beneficial interaction between turbines that are located more than three rows apart is limited.

The choice of $T_A$ is driven by conflicting incentives of reducing computational cost ($T_A \rightarrow T$) and mitigating finite-horizon effects ($T_A \rightarrow 0$). As a rule of thumb, we have been using $T_A = T/2$ in previous work. In the MM17 paper we have also tried $T_A = T/4$ and found that the effect on power gain was limited. (TODO: Perhaps include a figure here? )

Both concepts described above are discussed in more detail in the MM17 paper. To keep the description of the case setup concise in the current work, we simply add that $T = 240$ s corresponds to the time to pass four turbine rows, and further refer to MM17 for further elaboration of methodological choices. We have updated the revised manuscript as follows (l XX p XX):

``
... with a prediction horizon $T = 240$ s \textbf{(i.e. the time it takes for the flow to pass four rows of turbines)} and a flow ...
'',

and (l XX p XX, end of Section 2.2):

`` 
... and $\tau = 30$ s. \textbf{The choice of (and sensitivity to) setup parameters is further elaborated in MM17.}
''

\dotfill

\textbf{Reviewer: } \textit{As clearly seen in Figure 4c and 4d, there is a significant increase in turbulence
	further downstream. In addition to the TKE and the transport, would be nice to have the
	turbulence intensity TI values (as listed later on page 20, 10\% for the baseline case),
	both for the baseline case and the maximum added TI reported - possibly somewhere
	around Figure 4. That again would give an indication on the applicability compared to
	the field values observed. Also note the typo in the caption of Figure 4: after c) all the
	subplots are marked to be continuously c).}

\textbf{Response: }

We have added the requested TI values in the discussion around Figure 4 as follows (p ?? l ??):

``
.., for which an enhanced recovery was found as discussed above. \textbf{The turbulence intensity $TI \equiv (2k/3  )^{1/2}/(U_x^2+U_y^2+U_z^2)^{1/2}$ at hub height (not shown in the figure) is 10\% at the inlet for both the reference case and the controlled case. The combination of reduced near-wake mean velocities and increased velocity fluctuations in the controlled case increase local $TI$ in the turbine wakes (ranging from $\approx 2$\%-points in the wakes of middle rows to $\approx 12$\%-points in the first and last rows). This increase in turbulence intensity dissipates to below $1$\%-point difference at 10$D$ downstream of the last row.}
''

Furthermore, we thank the reviewer for pointing out the typos in the caption. This has been fixed in the revised manuscript.

\dotfill

\textbf{Reviewer: } \textit{On page 10, around line 10, the argument of "upstream actions do not require a
	specific downstream response in order to increase power in that downstream row",
	which is also paraphrased in the conclusions, needs to be elaborated. This rather
	broad conclusion seem to oversee the probability of the curtailment of the downstream
	turbine where down-regulation might be inevitable for certain CT set-points assigned
	to downstream turbine(s) in the resulting optimization. Could be partially true for the
	investigated C3t5 case since there observed very limited curtailment even at the most
	upstream turbine (as in Figure 3b). However, also seen in Figure 8b (except of the
	very last row as the authors indicated), there seem to be still a difference between on
	the power gain at turbine R11 for the scenarios of R1-R10 and R1-R11. Narrowing the argument to the considered case or very little to no downstream curtailment CT
	distributions is suggested.}

\textbf{Response: } WM - Ik vind de commentaar van de reviewer hier vrij verwarrend. Weet niet exact wat hij hier bedoelt. Hopelijk is onderstaand antwoord voldoende.

%We thank the reviewer for this comment. We believe that the statement referred to by the reviewer still holds, i.e.
%
%``An interesting observation from this figure is that , for any row $i$ except the last one, the power potential as observed in case C3t5 is almost fully recovered by
%only applying the optimal controls up to row $i-1$''. 
%
%This is a simple observation that stems directly from the data in the figure. Unless we misinterpret the point of the reviewer, we do not agree that a curtailment strategy would invalidate this statement. If, let's say, a curtailment strategy would have been incorporated into the control law for row $i$ (i.e. to the benefit of row $i+1$), the application of the control R1-R$_{i-1}$ would yield even greater power gains in row $i$ than when applying controls R1-R$_i$, so the statement would still hold. 
%
%However, we do agree that statements and conclusions made throughout section 3 are specific to the current case and should therefore be interpreted with care. We have updated the manuscript to be more explicit that results are not overall conclusions for wind-farm control in general, but should be interpreted as observations of the current optimal control case. This has been done at the beginning of Section 3, prior to the elaboration of analyses and experiments as follows (l XX p XX):
%
%`` 
%... to uncover some of the characteristics of these control signals. \textbf{Note that the conclusions drawn within this paragraph should be interpreted as observations of the current C3t5 optimal control cases, given specific wind-farm layout and flow conditions, and hence cannot just be generalized for any wind-farm control in general.}
%''
%
%We hope that the updates to the manuscript encompass the concerns expressed by the reviewer.

We thank the reviewer for this comment. We agree that statements and conclusions made throughout section 3 are specific to the current case and should therefore be interpreted with care. However, possible curtailment is not the only reason why certain statements in this section could be invalidated, e.g. possible phase-locked correlated control actions between turbines (which were found not to be present in the current case) would invalidate the same statement as referred to by the reviewer.

We have updated the manuscript to be more explicit in the sense that results are not overall conclusions for wind-farm control in general, but should be interpreted as observations of the current optimal control case. This has been done at the beginning of Section 3, prior to the elaboration of analyses and experiments as follows (l XX p XX):

`` 
... to uncover some of the characteristics of these control signals. \textbf{Note that the conclusions drawn within this section should be interpreted as observations of the current C3t5 optimal control cases, given specific wind-farm layout and flow conditions, and hence cannot just be generalized for any wind-farm control in general.}
''

However, we do believe that the sinusoidal thrust control strategy is something that can be generalized to other wind-farm setups (as evidenced by Section 4.2.2). We hope that the updates to the manuscript encompass the concerns expressed by the reviewer.

\dotfill

\textbf{Reviewer: } \textit{On page 13, line 13, "the presence of the flow invariant features of the control signals"
	needs further justification as Figure 11 would also depend on how variant the flow
	features are in the simulations. That should include both the spatial and temporal
	variance within the 30-min window. As far as the field measurements are concerned,
	high spatial and temporal correlations are observed. For the former, Figure 2 gives a
	brief idea about the wind speed range between the columns, that can be referred here.
	For the latter, time series or relevant temporal statistics can be presented to assess the
	randomness and strengthen the hypothesis.}

\textbf{Response: } We thank the reviewer for this useful comment. We have quantified spatial variance (between columns) as well as temporal variance (between time windows) of the flow field by investigating the correlation between the incoming disk-averaged velocity at 6$D$ upstream of the first-row turbines (illustrated for two columns of the wind farm in Figure \ref{fig:upstream_vel} of the current document). As shown in the Figure, there is considerable variation of the incoming velocities between columns and between time windows, and velocity fluctuations seem qualitatively uncorrelated. To keep the discussion in the manuscript concise, we do not include the figure below but instead report the average Pearson correlation coefficient between the incoming velocity fluctuations in different columns (for the full time horizon) and in different time windows (averaged over all columns), with low values of 0.12 and 0.07 respectively.

We have incorporated this in the revised manuscript as follows (p ??):

``
... the column swap is performed in 2 random independent ways. \textbf{The variability of flow conditions for different columns can be qualitatively observed in Figure 2. To further strengthen the hypothesis of the current experiment, we verified that the  correlation between flow conditions in different columns is small, i.e. with an average Pearson correlation coefficient of 0.12 between columns for the incoming velocity fluctuations 6$D$ upstream of the first row.}
''

and (p ??):

``
whereas the time synchronization of control actions to specific flow events is 10 eliminated. Similar to the first case, this is done in 2 random ways, \textbf{and the limited correlation between velocity fluctuations in different time windows was quantified at 0.07}. Figure 11b illustrates the row-averaged power for these...
cases.
''

\begin{figure}
	\includegraphics[width=\textwidth]{upstream_vel.eps}
	\caption{Disk-averaged axial velocity upstream of the first-row turbines for column 1 and column 2 in Figure 2 of the manuscript. Vertical dotted lines demarcate the optimization window boundaries (i.e. at integer multiples of $T_A$) \label{fig:upstream_vel}}
\end{figure}

\dotfill

\textbf{Reviewer: } \textit{On page 17, around line 5, a very nice example on how to implement the optimized
	sinusoidal CT is presented. The practical examples can be further improved by a
	short discussion on the expected response time of such increases in tip speed ratio
	on a machine with high inertia. That would put the estimated sine wave period into
	perspective as well.}

\textbf{Response: } Indeed, the high inertia of the rotor could be a compromise to achieving the desired sinusoidal thrust coefficient behavior. However, given the proposed redesign of the NREL 5MW rotor, dynamic reductions from the maximum $C_T' = 3.5$ could be achieved rapidly by using the fast pitch actuators with which modern turbines are equipped. We have added this comment to the discussion as follows:

``
... be \textbf{attained by the NREL 5MW turbine with a slight modification in the rotor design}, i.e. a 50\% increase in blade chord length and an operational tip speed ratio 25\% higher than 
the original design value (see Goit and Meyers, 2015, Appendix A). \textbf{Furthermore, given this redesign, dynamic reductions from this maximum value could be realized through pitch control. More specifically, based on speculated specifications of similar turbines in practice, the NREL 5 MW is designed with a pitch rate of $8^{\circ}/$s. This implies that the rotor can go from power maximization ($0^\circ$ pitch) to total curtailment in fully feathered blade position ($90^\circ$ pitch) in 11.25 seconds (Jonkman et al. 2009).}
''

\dotfill

\textbf{Reviewer: } \textit{For Section 4.2.3, the header ``Full-scale wind farm test" is a bit misleading... Suggest
	to change to "Full-scale wind farm simulations (in LES)" instead.}

\textbf{Response: } Agreed, we have changed the section header to ``Full-scale wind-farm LES"

\dotfill

\textbf{Reviewer: } \textit{On Figure 19, why would the power decrease after Row 5 for the sinusoidal case?}

\textbf{Response: } This can be explained based on Figure 20: it can be seen from the cross-section views that by sinusoidally varying first-row thrust, the time-averaged axial velocity at the turbine disk for R2 is increased, but the flow above the turbine disk is actually slowed down, hence indicating that momentum is entrained from the internal boundary layer above the turbine canopy. In downstream rows, this high momentum zone expands and mixes with the background flow field. Starting from R5, this zone is almost completely dissipated, and the disk velocity is actually slighter lower than in the reference case. In short, the sinusoidal control actions of the first row actually entrain momentum from the surrounding flow that would otherwise be entrained by natural turbulent mixing further downstream in the wind farm.

Moreover, it can be seen that the full optimal control (C3t5 in Figure 20) however is able to contain the high momentum zone by continuously improving the mixing process with the background boundary layer. The mechanisms thereof however remain elusive to date, and are subject of further investigation.

We have incorporated this discussion in the revised version of the manuscript as follows (p ?? l ??):

``
... with higher rotor velocities in the downstream as well.  \textbf{Note also that, for the fifth row, the disk velocity is slightly lower for the sinusoidal control case than for the reference case, consistent with the decreased power extraction observed in Fig. 19. This can be explained by the fact that first-row control actions cause enhanced entrainment of momentum from the internal boundary layer above the turbine canopy that would otherwise be entrained by natural turbulent mixing in passive downstream rows. In consequence, lesser entrainment occurs for downstream rows, resulting in a slight decrease in disk velocities from the fifth row onwards.}
''

\dotfill

\textbf{Reviewer: } \textit{Page 22 around line 5, the (inevitable) discussions on loads are included. In addition
	to the loads on the controlled upstream turbine, Figure 20(b) indicates partial wakes
	on the further downstream rows of turbines. Therefore, the section should be improved
	by highlighting the probable increase in fatigue loading for not just turbine(s) R1 but for
	the downstream rows as well, possibly starting as early as R3.}

\textbf{Response: } We thank the reviewer for this comment. Agreed, the downstream turbines could also be subjected to increased fatigue loading, and we have included this in the revised manuscript as follows (p ??, l ??):

``
... to fatigue loading \textbf{of the first-row turbines. Furthermore, partial wake alleviation and unsteady passing of abovementioned vortex rings could also increase fatigue loading in downstream rows.} Hence, structural aspects ...
''

\dotfill

\textbf{Reviewer: } \textit{On the grammatical note, the manuscript is clear and easy to follow. The only comment
	might be on the use of Sect. or Section; Fig. or Figure references.}

\textbf{Response: } We have followed the official guidelines for referring to figures and sections as provided by the publisher, paraphrasing from the website:\

\emph{The abbreviation ``Sect.'' should be used when it appears in running text and should be followed by a number unless it comes at the beginning of a sentence.}\\({\small \verb|https://www.wind-energy-science.net/for_authors/manuscript_preparation.html|})

\end{document}

### 2018-05-17 revision/revised_WM01.tex
First draft of replies

--- 2018-02-14 draft.tex
@@
 \usepackage{amsfonts}
+
+\newcommand{\revision}[1]{{\color{red} #1}}
 
@@
 
-\copyrightstatement{TEXT}
+%\copyrightstatement{TEXT}
 
@@
 \subsection{Case setup}
-The wind farm considered in MM17 has an aligned pattern of 12 rows by 6 columns, with 6 rotor diameters spacing in axial and tranversal directions.
-The flow is simulated on a domain of $10 \times 3.6 \times 1$ km$^3$, discretized on a simulation grid of $384 \times 192 \times 144$ grid points.
+The wind farm considered in MM17 has an aligned pattern of 12 rows by 6 columns.
+\revision{The wind turbines have a hub height $z_h = 100$~m with a rotor diameter $D = 100$~m, and are spaced apart by 6$D$ in both axial and transversal directions.} The flow is simulated on a domain of $10 \times 3.6 \times 1$ km$^3$, discretized on a simulation grid of $384 \times 192 \times 144$ grid points.
 A snapshot of the streamwise velocity field is shown in Fig.
 \ref{fig:flowfield_MM17}.
-The wind farm was controlled over a total of $N_A = 15$ time windows with a prediction horizon $T = 240$ s and a flow advancement time of $T_A = T/2 = 120$ s, resulting in a total control time $T_{\rm tot} = N_A T_A = 30$ minutes.
+The wind farm was controlled over a total of $N_A = 15$ time windows with a prediction horizon $T = 240$ s (\revision{i.e.
+the time it takes for the flow to pass four rows of turbines}) and a flow advancement time of $T_A = T/2 = 120$ s, resulting in a total control time $T_{\rm tot} = N_A T_A = 30$ minutes.
 
@@
 Several different optimal control cases were defined, based on the wind-turbine response time $\tau = $ 0, 5, or 30 s (instantaneous, fast, or slow response, see Eq.
-\ref{eq:ctihat_constraint}) and the maximal thrust coefficient $C_{T,\text{max}}' = 2$ or~3 (underinductive or overinductive, see Eq.
+\ref{eq:ctihat_constraint}) and the maximal thrust coefficient $C_{T,\text{max}}' = 2$ or~3 \revision{, with thrust forces that can respectively only be reduced (underinductive), or also increased (overinductive) compared to the Betz optimum at $C_T' = 2$} (see Eq.
 \ref{eq:boxct_constraint}).
@@
 C3t30  for the case with $C_{T,\text{max}}' = 3$ and $\tau = 30$ s.
+\revision{The choice of (and sensitivity to) setup parameters is further elaborated in MM17.}
 
@@
 
-Figures \ref{fig:flowfield_statistics}c shows sideviews of turbulence kinetic energy.
+Figure \ref{fig:flowfield_statistics}c shows sideviews of turbulence kinetic energy $k$.
 The figure shows an increase in turbulence throughout the entire wind farm, spreading to the internal boundary layer above the turbines.
 Note specifically the sharp increase in turbulence in the core wake region behind the first-row turbine, for which an enhanced recovery was found as discussed above.
+\revision{The turbulence intensity $TI \equiv (2k/3  )^{1/2}/(U_x^2+U_y^2+U_z^2)^{1/2}$ at hub height (not shown in the figure) is 10\% at the inlet for both the reference case and the controlled case.
+The combination of reduced near-wake mean velocities and increased velocity fluctuations in the controlled case increase local $TI$ in the turbine wakes (ranging from $\approx 2$\%-points in the wakes of middle rows to $\approx 12$\%-points in the first and last rows).
+This increase in turbulence intensity dissipates to below $1$\%-point difference at 10$D$ downstream of the last row.}
 
@@
 \emph{c):} Turbulence kinetic energy $k = (1/2)(\overline{\widetilde{u}'_x \widetilde{u}'_x} + \overline{\widetilde{u}'_y \widetilde{u}'_y} + \overline{\widetilde{u}'_z \widetilde{u}'_z})$ (sideview).
-\emph{c):} Turbulence top-down transport $ - \overline{\widetilde{u}'_x\widetilde{u}'_z}$.
-\emph{c):} Mean-flow top-down transport $ - \widetilde{U}_x \widetilde{U}_z$ (sideview).
-\emph{c):} Turbulence horizontal transport $\overline{\widetilde{u}'_x \widetilde{u}'_y}$ (planview).
-\emph{c):} Mean-flow horizontal transport $\widetilde{U}_x \widetilde{U}_y$ (planview).
+\emph{d):} Turbulence top-down transport $ - \overline{\widetilde{u}'_x\widetilde{u}'_z}$.
+\emph{e):} Mean-flow top-down transport $ - \widetilde{U}_x \widetilde{U}_z$ (sideview).
+\emph{f):} Turbulence horizontal transport $\overline{\widetilde{u}'_x \widetilde{u}'_y}$ (planview).
+\emph{g):} Mean-flow horizontal transport $\widetilde{U}_x \widetilde{U}_y$ (planview).
 Black lines indicate wind turbine locations.
@@
 The current section focuses on the optimal thrust coefficients generated by the optimal control simulations in MM17, and performs numerical experiments to uncover some of the characteristics of these control signals.
+\revision{Note that the conclusions drawn within this section should be interpreted as observations of the current C3t5 optimal control cases, given specific wind-farm layout and flow conditions, and hence cannot just be generalized for any wind-farm control in general.}
 
@@
 The intermediate rows (R2--R11) lie closer together, with the general trend being that the potential is somewhat decreased with downstream distance into the wind farm, although this decrease is not monotonous.
+\revision{Note that these observations are in contrast to the study of \cite{ciri2017large}, where extremum seeking control is applied to optimize individual turbines in an array of three turbines.
+They found that set points in the first-row turbines remain untouched from their locally optimal values whereas downstream turbines are self-optimized for wake conditions.
+This contrast can be attributed to our dynamic and simultaneous wind-farm control approach, which differs from their sequential individual set-point control.
+Furthermore, self-optimization in the latter study is achieved by recalibrating the torque gain.
+The baseline controller in our current actuator disk model makes abstraction of pitch and generator torque control and keeps $C_T'$ at its theoretical optimum instead.
+}
 
@@
 To avoid erroneous conclusions based on coincidence, the column swap is performed in 2 random independent ways.
-Row-averaged power for these cases is shown in Fig.~\ref{fig:scrambled}a.
+\revision{The variability of flow conditions for different columns can be qualitatively observed in Figure 2.
+To further strengthen the hypothesis of the current experiment, we verified that the  correlation between flow conditions in different columns is small, i.e.
+with an average Pearson correlation coefficient of 0.12 between columns for the incoming velocity fluctuations 6$D$ upstream of the first row.} Row-averaged power for these cases is shown in Fig.~\ref{fig:scrambled}a.
 
@@
 In this way, the spectral thrust characteristics for timescales smaller that the control horizon $T_A = 120$ s remain unchanged, whereas the time synchronization of control actions to specific flow events is eliminated.
-Similar to the first case, this is done in 2 random ways, and Fig.~\ref{fig:scrambled}b illustrates the row-averaged power for these cases.
+Similar to the first case, this is done in 2 random ways, \revision{and the limited correlation between velocity fluctuations in different time windows was quantified at 0.07}.
+Figure ~\ref{fig:scrambled}b illustrates the row-averaged power for these cases.
 
@@
 For instance, a Strouhal number $St = 0.25$ corresponds here to a sine wave period of $\approx$ 50 s for a turbine with diameter $D = 100$ m and a freestream velocity $U_\infty = 8.5$ m s$^{-1}$.
-The resulting maximum thrust coefficient of 3.5 can, e.g., be attained by an NREL 5MW rotor with a 50\% increase in chord length at a tip speed ratio that is 25\% higher than the greedy tip speed ratio (see \citealp{goit2015optimal}, Appendix A).
-Figure~\ref{fig:sinus_baseline}b illustrates normalized power extraction by row for the reference case, the best sinusoidal case, and the first four rows of the optimal control case C3t5 from MM17.
+The resulting maximum thrust coefficient of 3.5 can, e.g., be \revision{attained by the NREL 5MW turbine with a slight modification in the rotor design}, i.e.
+a 50\% increase in chord length at a tip speed ratio that is 25\% higher than the greedy tip speed ratio (see \citealp{goit2015optimal}, Appendix A).
+\revision{Furthermore, given this redesign, dynamic reductions from this maximum value could be realized through pitch control.
+More specifically, based on speculated specifications of similar turbines in practice, the NREL 5 MW is designed with a pitch rate of $8^{\circ}/$s.
+This implies that the rotor can go from power maximization ($0^\circ$ pitch) to total curtailment in fully feathered blade position ($90^\circ$ pitch) in 11.25 seconds \cite{jonkman2009definition}.} Figure~\ref{fig:sinus_baseline}b illustrates normalized power extraction by row for the reference case, the best sinusoidal case, and the first four rows of the optimal control case C3t5 from MM17.
 The figure shows that the power gain in the sinusoidal cases originates mostly from the second row, and that power in the first row is decreased by approximately 5\%.
@@
 
-\subsubsection{Full-scale wind-farm test}
-In the remainder of this section, the sinusoidal variation strategy will be tested in a full-scale wind farm, corresponding to the full $12 \times 6$ aligned wind farm of MM17.
+\subsubsection{\revision{Full-scale wind-farm LES}}
+In the remainder of this section, the sinusoidal variation strategy will be tested in a full-scale wind farm \revision{LES}, corresponding to the full $12 \times 6$ aligned wind farm of MM17.
 Simulations are performed for a reduced range of amplitudes and Strouhal numbers, corresponding to the most favorable region identified in the parameter sweeps above.
@@
 
-Figure~\ref{fig:sinus_fullscale} shows the power extraction of the full-scale test.
+Figure~\ref{fig:sinus_fullscale} shows the power extraction of the full-scale \revision{LES}.
 Figure~\ref{fig:sinus_fullscale}a shows the relative power gains over the reference case for the full wind farm.
@@
 In contrast, case C3t5, in which all turbines are actively controlled, succeeds to attain similar cross section characteristics with higher rotor velocities in the downstream as well.
+\revision{Note also that, for the fifth row, the disk velocity is slightly lower for the sinusoidal control case than for the reference case, consistent with the decreased power extraction observed in Fig.
+\ref{fig:sinus_fullscale}.
+This can be explained by the fact that first-row control actions cause enhanced entrainment of momentum from the internal boundary layer above the turbine canopy that would otherwise be entrained by natural turbulent mixing in passive downstream rows.
+In consequence, lesser entrainment occurs for downstream rows, resulting in a slight decrease in disk velocities from the fifth row onwards.}
 
@@
 However, important comments should be made.
-First, sustained sinusoidal thrust variations with a large amplitude could contribute significantly to turbine fatigue loading.
+First, sustained sinusoidal thrust variations with a large amplitude could contribute significantly to turbine fatigue loading \revision{of the first-row turbines.
+Furthermore, partial wake alleviation and unsteady passing of abovementioned vortex rings could also increase fatigue loading in downstream rows.}.
 Hence, structural aspects should be taken into account upon evaluating the practical viability of the approach.
@@
 
-\dataavailability{TEXT} %% use this section when having only data sets available
+%\dataavailability{TEXT} %% use this section when having only data sets available
 

### 2018-05-17 revision/reply_to_revs_2.tex
R2C1

@@
 
-\textbf{Reviewer: } \textit{While describing the case setup in Section 2.2, the "flow advancement time", $T_A$ (also
+\textbf{1. Reviewer: } \textit{While describing the case setup in Section 2.2, the "flow advancement time", $T_A$ (also
 	referred in Figure 1) is considered as half of the prediction horizon $T$. Would $T_A$ (and
@@
 
-\textbf{Response: } Indeed, $T$ and $T_A$ are important parameters in the receding-horizon approach.
+\textbf{Response: } It is true that $T$ and $T_A$ are important parameters in the receding-horizon approach.
 The optimization horizon $T$ can indeed be considered inflow-dependent.
@@
 In theory, it is desirable to have $T$ cover an entire wind-farm throughflow.
-In practice however, control over long time horizons is complicated by the chaotic nature of turbulence accompanied by adjoint gradient inaccuracies, and we choose $T$ as long as possible for our optimization to be unharmed by these inaccuracies (i.e.
+In practice however, control over long time horizons is complicated by the chaotic nature of turbulence accompanied by adjoint gradient inaccuracies, and we choose $T$ as long as possible without having the optimization be affected by these inaccuracies (i.e.
 240 s).
@@
 In the MM17 paper we have also tried $T_A = T/4$ and found that the effect on power gain was limited.
-(TODO: Perhaps include a figure here?
-)
 
@@
 
-\textbf{Reviewer: } \textit{As clearly seen in Figure 4c and 4d, there is a significant increase in turbulence
+\textbf{2. Reviewer: } \textit{As clearly seen in Figure 4c and 4d, there is a significant increase in turbulence
 	further downstream. In addition to the TKE and the transport, would be nice to have the
@@
 
-\textbf{Reviewer: } \textit{On page 10, around line 10, the argument of "upstream actions do not require a
+\textbf{3. Reviewer: } \textit{On page 10, around line 10, the argument of "upstream actions do not require a
 	specific downstream response in order to increase power in that downstream row",
@@
 
-\textbf{Reviewer: } \textit{On page 13, line 13, "the presence of the flow invariant features of the control signals"
+\textbf{4. Reviewer: } \textit{On page 13, line 13, "the presence of the flow invariant features of the control signals"
 	needs further justification as Figure 11 would also depend on how variant the flow
@@
 
-\textbf{Reviewer: } \textit{On page 17, around line 5, a very nice example on how to implement the optimized
+\textbf{5. Reviewer: } \textit{On page 17, around line 5, a very nice example on how to implement the optimized
 	sinusoidal CT is presented. The practical examples can be further improved by a
@@
 
-\textbf{Reviewer: } \textit{For Section 4.2.3, the header ``Full-scale wind farm test" is a bit misleading... Suggest
+\textbf{6. Reviewer: } \textit{For Section 4.2.3, the header ``Full-scale wind farm test" is a bit misleading... Suggest
 	to change to "Full-scale wind farm simulations (in LES)" instead.}
@@
 
-\textbf{Reviewer: } \textit{On Figure 19, why would the power decrease after Row 5 for the sinusoidal case?}
+\textbf{7. Reviewer: } \textit{On Figure 19, why would the power decrease after Row 5 for the sinusoidal case?}
 
@@
 
-\textbf{Reviewer: } \textit{Page 22 around line 5, the (inevitable) discussions on loads are included. In addition
+\textbf{8. Reviewer: } \textit{Page 22 around line 5, the (inevitable) discussions on loads are included. In addition
 	to the loads on the controlled upstream turbine, Figure 20(b) indicates partial wakes
@@
 
-\textbf{Reviewer: } \textit{On the grammatical note, the manuscript is clear and easy to follow. The only comment
+\textbf{9. Reviewer: } \textit{On the grammatical note, the manuscript is clear and easy to follow. The only comment
 	might be on the use of Sect. or Section; Fig. or Figure references.}

### 2018-05-17 revision/reply_to_revs_2.tex
R2C2 (replies + manuscript)

@@
 .., for which an enhanced recovery was found as discussed above.
-\textbf{The turbulence intensity $TI \equiv (2k/3  )^{1/2}/(U_x^2+U_y^2+U_z^2)^{1/2}$ at hub height (not shown in the figure) is 10\% at the inlet for both the reference case and the controlled case.
+\textbf{The turbulence intensity $TI \equiv (2k/3  )^{1/2}/U_\infty$ at hub height (not shown in the figure) is 10\% at the inlet for both the reference case and the controlled case.
 The combination of reduced near-wake mean velocities and increased velocity fluctuations in the controlled case increase local $TI$ in the turbine wakes (ranging from $\approx 2$\%-points in the wakes of middle rows to $\approx 12$\%-points in the first and last rows).

### 2018-05-17 revision/revised_WM02.tex
R2C2 (replies + manuscript)

--- 2018-05-17 revision/revised_WM01.tex
@@
 Note specifically the sharp increase in turbulence in the core wake region behind the first-row turbine, for which an enhanced recovery was found as discussed above.
-\revision{The turbulence intensity $TI \equiv (2k/3  )^{1/2}/(U_x^2+U_y^2+U_z^2)^{1/2}$ at hub height (not shown in the figure) is 10\% at the inlet for both the reference case and the controlled case.
+\revision{The turbulence intensity $TI \equiv (2k/3  )^{1/2}/U_\infty$ at hub height (not shown in the figure) is 10\% at the inlet for both the reference case and the controlled case.
 The combination of reduced near-wake mean velocities and increased velocity fluctuations in the controlled case increase local $TI$ in the turbine wakes (ranging from $\approx 2$\%-points in the wakes of middle rows to $\approx 12$\%-points in the first and last rows).

### 2018-05-17 revision/reply_to_revs_2.tex
Trivial changes to REv2

@@
 
-\textbf{Response: } WM - Ik vind de commentaar van de reviewer hier vrij verwarrend. Weet niet exact wat hij hier bedoelt. Hopelijk is onderstaand antwoord voldoende.
-
-%We thank the reviewer for this comment. We believe that the statement referred to by the reviewer still holds, i.e.
-%
-%``An interesting observation from this figure is that , for any row $i$ except the last one, the power potential as observed in case C3t5 is almost fully recovered by
-%only applying the optimal controls up to row $i-1$''. 
-%
-%This is a simple observation that stems directly from the data in the figure.
-Unless we misinterpret the point of the reviewer, we do not agree that a curtailment strategy would invalidate this statement.
-If, let's say, a curtailment strategy would have been incorporated into the control law for row $i$ (i.e.
-to the benefit of row $i+1$), the application of the control R1-R$_{i-1}$ would yield even greater power gains in row $i$ than when applying controls R1-R$_i$, so the statement would still hold.
-%
-%However, we do agree that statements and conclusions made throughout section 3 are specific to the current case and should therefore be interpreted with care.
-We have updated the manuscript to be more explicit that results are not overall conclusions for wind-farm control in general, but should be interpreted as observations of the current optimal control case.
-This has been done at the beginning of Section 3, prior to the elaboration of analyses and experiments as follows (l XX p XX):
-%
-%`` 
-%...
-to uncover some of the characteristics of these control signals.
-\textbf{Note that the conclusions drawn within this paragraph should be interpreted as observations of the current C3t5 optimal control cases, given specific wind-farm layout and flow conditions, and hence cannot just be generalized for any wind-farm control in general.}
-%''
-%
-%We hope that the updates to the manuscript encompass the concerns expressed by the reviewer.
-
-We thank the reviewer for this comment.
-We agree that statements and conclusions made throughout section 3 are specific to the current case and should therefore be interpreted with care.
-However, possible curtailment is not the only reason why certain statements in this section could be invalidated, e.g.
-possible phase-locked correlated control actions between turbines (which were found not to be present in the current case) would invalidate the same statement as referred to by the reviewer.
+\textbf{Response: } We find it difficult to follow the argument formulated by the reviewer here.
+However, we agree in general that statements and conclusions made throughout section 3 are specific to the current case and should therefore be interpreted with care.
 
 We have updated the manuscript to be more explicit in the sense that results are not overall conclusions for wind-farm control in general, but should be interpreted as observations of the current optimal control case.
-This has been done at the beginning of Section 3, prior to the elaboration of analyses and experiments as follows (l XX p XX):
+See the beginning of Section 3 (l XX p XX):
 
@@
 ''
-
-However, we do believe that the sinusoidal thrust control strategy is something that can be generalized to other wind-farm setups (as evidenced by Section 4.2.2).
-We hope that the updates to the manuscript encompass the concerns expressed by the reviewer.
 
@@
 ``
-whereas the time synchronization of control actions to specific flow events is 10 eliminated.
+whereas the time synchronization of control actions to specific flow events is eliminated.
 Similar to the first case, this is done in 2 random ways, \textbf{and the limited correlation between velocity fluctuations in different time windows was quantified at 0.07}.

### 2018-05-23 revision/reply_to_revs_2.tex
Feedback Johan part 1

@@
 \usepackage{graphicx}
+\usepackage{color}
 
@@
 
-\textbf{Response:} The reason for this different behavior of the downstream vs upstream turbines is twofold:
-\begin{enumerate}
-	\item The Extremum Seeking Control (ESC) approach by Ciri \emph{et al.} differs fundamentally from our approach.
-In our approach, we simultaneously optimize all control variables over a given time horizon, which allows for full dynamic cooperation and interaction between all turbines.
-In the ESC, controls are optimized one by one in the sense that ESC is initiated on the first row turbine and will be activated on the subsequent row only after its upstream neighbor has converged to a steady state solution.
-In this way, dynamic control and interaction between controls in different turbines are discarded a priori.
-Furthermore, although this is not an inherent limitation of the ESC approach, Ciri \emph{et al.} optimize for individual turbine power instead of total wind-farm power (see Eq.
-10 $P_i$ is used instead of $\sum_i P_i$).
-This is explicitly mentioned in the title of their follow up paper (Ciri \emph{et al.} Large-eddy simulations with extremum-seeking control for individual wind turbine power optimization)
+\textbf{Response:} This is an interesting comment.
+It is correct that this contrast results from the way turbines and their controls are modeled in the current study and that of Ciri et al.
+The reference disk-based thrust coefficient $C_T' = 2$ used here actually implies intrinsic self-optimization of blade pitch and generator torque to local flow conditions, even though these actions are not resolved by our current actuator disk formulation which directly controls $C_T'$.
+In case these degrees of freedom would be resolved, for instance using an actuator line model as in Ciri et al., the ESC from the latter study could e.g.
+be used to optimize torque controller gain in order to achieve an effective $C_T' \approx 2$.
 
-	\item In our modeling approach, we make abstraction of detailed turbine dynamics in the sense that we directly control the thrust coefficient $C_T'$ instead of the generator torque or tip speed ratio.
-In this sense, the optimal thrust coefficient value of a turbine with no downstream neighbors can be analytically found to be $C_T' = 2$.
-In Ciri \emph{et al.}, the authors indicate that the reason why the TSR in the downstream rows (around 6.5) does not correspond to the individual optimal TSR (7.5) of individual turbines is that the TSR depends on a reference velocity and (direct quotation) ``As a consequence, the TSR is not a meaningful indicated of a waked turbine operating condition".
-
-\end{enumerate}
-
-Summarizing, the inverse behavior of the control dynamics compared to Ciri \emph{et al.} can be attributed to the difference in approach to wind-farm control (dynamic full-farm optimization vs steady sequential and individual optimization) as well as the implementation of the turbine control ($C_T'$ vs $\lambda$).
-In our opinion, the current study (full-farm dynamic wind-farm optimization) and that of Ciri \emph{et al.} (providing a practical ESC method for optimizing individual turbines) cannot be directly compared.
-
-However, we understand the contrast in the results and therefore added a small comment to the revised manuscript  (l. ??, p. ??):
+We have included the fact that the reference case already implies self-optimization in the revised manuscript as follows (p 4, l. 21):
 
 ``
-...
-is not monotonous.
-\textbf{Note that these observations are in contrast to the study of Ciri et al.
-2017, where extremum seeking control is applied to optimize individual turbines in an array of three turbines.
-They found that set points in the first-row turbines remain untouched from their locally optimal values whereas downstream turbines are re-optimized for wake conditions.
-This contrast can be attributed to our dynamic and simultaneous wind-farm control approach, which differs from their sequential individual set-point control.
-Furthermore, self-optimization in the latter study is achieved by recalibrating the torque gain.
-The baseline controller in our current actuator disk model makes abstraction of pitch and generator torque control and keeps $C_T'$ at its theoretical optimum instead.}
+A conventionally (greedily) controlled wind farm with steady $C_T' = 2$ was defined as a reference case.
+\textbf{Note that this would correspond to a farm with ideal turbines for which generator torque is being controlled dynamically to track the maximum power point at the Betz limit perfectly.} Several different optimal control cases were defined, ...
 ''
 
-WM - WEET NIET OF WE DIT WEL ECHT MOETEN DOEN, MISSCHIEN GEWOON REVIEWER OVERTUIGEN DAT WE DIT NIET INCLUDEN? TE BESPREKEN
+However, even given the discussion above, it is not unimaginable that a turbine could temporarily outperform this greedy control setting, e.g.
+by timely reacting to incoming flow accelerations.
+This is what we would refer to as self-optimization.
+{\color{blue} We have explicitly defined this in the text as follows (p ??):
+
+``
+...This suggests that self-optimization \textbf{(i.e.
+temporarily outperforming conventional greedy control in response to incoming flow events)} is very limited: the optimal controls for a given turbine are designed to create ...
+'' }
+
+(IS DIT OKE?? Kan het deel in blauw misschien beter gewoon weglaten?)
 
@@
 Current turbines are designed to approach maximum $C_T'$ values around 2, corresponding to the Betz limit.
-Although this maximum value can be increased by increasing the operational tip speed ratio (TSR), aiming to do so by adapting the pitch angle would inevitably lead to severe efficiency losses due to stall on the turbine blades.
+Aiming to increase thrust by adapting the pitch angle would inevitably lead to severe efficiency losses due to stall on the turbine blades.
 Therefore, we provide an example of how an alternative turbine design (i.e.
@@
 ``
-... be \textbf{attained by the NREL 5MW turbine with a slight modification in the rotor design}, i.e. a 50\% increase in blade chord length and an operational tip speed ratio 25\% higher than 
+The resulting maximum thrust coefficient of 3.5 can, e.g., be \textbf{attained by the NREL 5MW turbine with a slight modification in the rotor design}, i.e.
+a 50\% increase in blade chord length and an operational tip speed ratio 25\% higher than
 the original design value (see Goit and Meyers, 2015, Appendix A).
-\textbf{Furthermore, given this redesign, dynamic reductions from this maximum value could be realized through pitch control.
-More specifically, based on speculated specifications of similar turbines in practice, the NREL 5 MW is designed with a pitch rate of $8^{\circ}/$s.
-This implies that the rotor can go from power maximization ($0^\circ$ pitch) to total curtailment in fully feathered blade position ($90^\circ$ pitch) in 11.25 seconds (Jonkman et al.
+\textbf{Furthermore, given such redesign, dynamic reductions from this value could be realized through blade pitch control, for which actuation in the order of $10^\circ/$s is possible (see, e.g., Jonkman et al.
 2009).}
 ''
+
+Note that this control strategy is not unique and is solely given as an indication for the technical feasibility of tracking the proposed $C_T'$ waveform.
+In practice, we expect that this can also be achieved through a combination of generator torque and blade pitch control.
+This is subject of ongoing investigation.
 
@@
 
-\textbf{Response: } Indeed, the high inertia of the rotor could be a compromise to achieving the desired sinusoidal thrust coefficient behavior.
-However, given the proposed redesign of the NREL 5MW rotor, dynamic reductions from the maximum $C_T' = 3.5$ could be achieved rapidly by using the fast pitch actuators with which modern turbines are equipped.
-We have added this comment to the discussion as follows:
+\textbf{Response: } We thank the reviewer for this suggestion.
+Dynamic reductions from the maximum $C_T' = 3.5$ could, for instance, be achieved by using the fast pitch actuators with which modern turbines are equipped.
+We have added this comment to the discussion as follows (p ??):
 
 ``
-... be \textbf{attained by the NREL 5MW turbine with a slight modification in the rotor design}, i.e. a 50\% increase in blade chord length and an operational tip speed ratio 25\% higher than 
+The resulting maximum thrust coefficient of 3.5 can, e.g., be \textbf{attained by the NREL 5MW turbine with a slight modification in the rotor design}, i.e.
+a 50\% increase in blade chord length and an operational tip speed ratio 25\% higher than
 the original design value (see Goit and Meyers, 2015, Appendix A).
-\textbf{Furthermore, given this redesign, dynamic reductions from this maximum value could be realized through pitch control.
-More specifically, based on speculated specifications of similar turbines in practice, the NREL 5 MW is designed with a pitch rate of $8^{\circ}/$s.
-This implies that the rotor can go from power maximization ($0^\circ$ pitch) to total curtailment in fully feathered blade position ($90^\circ$ pitch) in 11.25 seconds (Jonkman et al.
+\textbf{Furthermore, given such redesign, dynamic reductions from this value could be realized through blade pitch control, for which actuation in the order of $10^\circ/$s is possible (see, e.g., Jonkman et al.
 2009).}
 ''
+
+Note that this control strategy is not unique and is solely given as an indication for the technical feasibility of tracking the proposed $C_T'$ waveform.
+In practice, we expect that this can also be achieved through a combination of generator torque and blade pitch control.
+This is subject of ongoing investigation.
 
@@
 
-Moreover, it can be seen that the full optimal control (C3t5 in Figure 20) however is able to contain the high momentum zone by continuously improving the mixing process with the background boundary layer.
+Moreover, it can be seen that the full optimal control (C3t5 in Figure 20) is able to contain the high momentum zone by continuously improving the mixing process with the background boundary layer.
 The mechanisms thereof however remain elusive to date, and are subject of further investigation.

### 2018-05-23 revision/revised_WM02.tex
Feedback Johan part 1

@@
 
-\newcommand{\revision}[1]{{\color{red} #1}}
+\newcommand{\revision}[1]{{\color{blue} #1}}
 
@@
 
-A conventionally (greedily) controlled wind farm with steady $C_T' = 2$ (corresponding to the Betz limit in momentum theory) was defined as a reference case.
-Several different optimal control cases were defined, based on the wind-turbine response time $\tau = $ 0, 5, or 30 s (instantaneous, fast, or slow response, see Eq.
+A conventionally (greedily) controlled wind farm with steady $C_T' = 2$ was defined as a reference case.
+\revision{Note that this would correspond to a farm with ideal turbines for which generator torque is being controlled dynamically to track the maximum power point at the Betz limit perfectly.} Several different optimal control cases were defined, based on the wind-turbine response time $\tau = $ 0, 5, or 30 s (instantaneous, fast, or slow response, see Eq.
 \ref{eq:ctihat_constraint}) and the maximal thrust coefficient $C_{T,\text{max}}' = 2$ or~3 \revision{, with thrust forces that can respectively only be reduced (underinductive), or also increased (overinductive) compared to the Betz optimum at $C_T' = 2$} (see Eq.
@@
 An interesting observation from this figure is that , for any row $i$ except the last one, the power potential as observed in case C3t5 is almost fully recovered by only applying the optimal controls up to row $i-1$.
-This suggests that self-optimization is very limited: the optimal controls for a given turbine are designed to create favorable flow conditions in the downstream rows instead of increasing local power.
+This suggests that self-optimization \revision{(i.e.
+temporarily outperforming conventional greedy control in response to incoming flow events)} is very limited: the optimal controls for a given turbine are designed to create favorable flow conditions in the downstream rows instead of increasing local power.
 Furthermore, although the discussion in the previous paragraph has shown that downstream controls are optimized with the upstream actions in mind, the converse is not true: upstream control actions do not require a specific downstream response in order to increase power in that downstream row.
@@
 a 50\% increase in chord length at a tip speed ratio that is 25\% higher than the greedy tip speed ratio (see \citealp{goit2015optimal}, Appendix A).
-\revision{Furthermore, given this redesign, dynamic reductions from this maximum value could be realized through pitch control.
-More specifically, based on speculated specifications of similar turbines in practice, the NREL 5 MW is designed with a pitch rate of $8^{\circ}/$s.
-This implies that the rotor can go from power maximization ($0^\circ$ pitch) to total curtailment in fully feathered blade position ($90^\circ$ pitch) in 11.25 seconds \cite{jonkman2009definition}.} Figure~\ref{fig:sinus_baseline}b illustrates normalized power extraction by row for the reference case, the best sinusoidal case, and the first four rows of the optimal control case C3t5 from MM17.
+\revision{Furthermore, given such redesign, dynamic reductions from this value could be realized through blade pitch control, for which actuation rates in the order of $10^\circ/$s are possible (see, e.g., \citealp{jonkman2009definition}).} Figure~\ref{fig:sinus_baseline}b illustrates normalized power extraction by row for the reference case, the best sinusoidal case, and the first four rows of the optimal control case C3t5 from MM17.
 The figure shows that the power gain in the sinusoidal cases originates mostly from the second row, and that power in the first row is decreased by approximately 5\%.

### 2018-05-23 revision/reply_to_revs_WM2.tex
R1C1 finalized

--- 2018-05-23 revision/reply_to_revs_2.tex
@@
 \textbf{Response:} This is an interesting comment.
-It is correct that this contrast results from the way turbines and their controls are modeled in the current study and that of Ciri et al.
+It is correct that the difference originates from the way turbines and their controls are modeled in the current study compared to that of Ciri et al.
 The reference disk-based thrust coefficient $C_T' = 2$ used here actually implies intrinsic self-optimization of blade pitch and generator torque to local flow conditions, even though these actions are not resolved by our current actuator disk formulation which directly controls $C_T'$.
@@
 
-We have included the fact that the reference case already implies self-optimization in the revised manuscript as follows (p 4, l. 21):
+We have included the fact that the reference case already implies self-optimization in the revised manuscript as follows (p 4, line 21):
 
@@
 A conventionally (greedily) controlled wind farm with steady $C_T' = 2$ was defined as a reference case.
-\textbf{Note that this would correspond to a farm with ideal turbines for which generator torque is being controlled dynamically to track the maximum power point at the Betz limit perfectly.} Several different optimal control cases were defined, ...
+\textbf{Note that this would correspond to a farm with ideal turbines for which generator torque is being controlled dynamically to track the maximum power point at the Betz limit perfectly.
+In a real turbine controller this may, e.g., be achieved with the ESC proposed by Ciri et al.
+(2017).} Several different optimal control cases were defined, ...
 ''
-
-However, even given the discussion above, it is not unimaginable that a turbine could temporarily outperform this greedy control setting, e.g.
-by timely reacting to incoming flow accelerations.
-This is what we would refer to as self-optimization.
-{\color{blue} We have explicitly defined this in the text as follows (p ??):
-
-``
-...This suggests that self-optimization \textbf{(i.e.
-temporarily outperforming conventional greedy control in response to incoming flow events)} is very limited: the optimal controls for a given turbine are designed to create ...
-'' }
-
-(IS DIT OKE?? Kan het deel in blauw misschien beter gewoon weglaten?)
 
@@
 
-\textbf{Response:} Indeed, this is correct. We have mentioned this explicitly in the revised manuscript as follows (p. ??, l. ??):
+\textbf{Response:} Indeed, this is correct. We have mentioned this explicitly in the revised manuscript as follows (p. 4, line 25):
 
@@
 
-We have slightly modified the statement in the revised manuscript to avoid any confusion with regard to possibly having the chord length as a control variable. (p. ??, l. ??):
+We have slightly modified the statement in the revised manuscript to avoid any confusion with regard to possibly having the chord length as a control variable. (p. 17, line 29):
 
@@
 the original design value (see Goit and Meyers, 2015, Appendix A).
-\textbf{Furthermore, given such redesign, dynamic reductions from this value could be realized through blade pitch control, for which actuation in the order of $10^\circ/$s is possible (see, e.g., Jonkman et al.
+\textbf{Furthermore, given such redesign, dynamic reductions from this value could be realized through blade pitch control, for which actuation rates in the order of $10^\circ/$s are possible (see, e.g., Jonkman et al.
 2009).}
@@
 
-\textbf{Response: } Indeed, we seem to have overlooked specifying the exact turbine dimensions in Section 2.2. We have updated the manuscript as follows (l XX p YY):
+\textbf{Response: } Indeed, we seem to have overlooked specifying the exact turbine dimensions in Section 2.2. We have updated the manuscript as follows (p. 4, line 15):
 
@@
 To keep the description of the case setup concise in the current work, we simply add that $T = 240$ s corresponds to the time to pass four turbine rows, and further refer to MM17 for further elaboration of methodological choices.
-We have updated the revised manuscript as follows (l XX p XX):
+We have updated the revised manuscript as follows (p.
+4, line 20):
 
@@
 
-and (l XX p XX, end of Section 2.2):
+and (p. 4, line 27):
 
@@
 
-We have added the requested TI values in the discussion around Figure 4 as follows (p ?? l ??):
+We have added the requested TI values in the discussion around Figure 4 as follows (p. 6, line 19):
 
@@
 We have updated the manuscript to be more explicit in the sense that results are not overall conclusions for wind-farm control in general, but should be interpreted as observations of the current optimal control case.
-See the beginning of Section 3 (l XX p XX):
+See the beginning of Section 3 (p.
+8, line 5):
 
@@
 
-We have incorporated this in the revised manuscript as follows (p ??):
+We have incorporated this in the revised manuscript as follows (p. 14, line 7):
 
@@
 
-and (p ??):
+and (p. 14, line 14):
 
@@
 \begin{figure}
-	\includegraphics[width=\textwidth]{upstream_vel.eps}
+	\centering
+	\includegraphics[width=0.65\textwidth]{upstream_vel.eps}
 	\caption{Disk-averaged axial velocity upstream of the first-row turbines for column 1 and column 2 in Figure 2 of the manuscript.
@@
 Dynamic reductions from the maximum $C_T' = 3.5$ could, for instance, be achieved by using the fast pitch actuators with which modern turbines are equipped.
-We have added this comment to the discussion as follows (p ??):
+We have added this comment to the discussion as follows (p.
+17, line 29):
 
@@
 
-We have incorporated this discussion in the revised version of the manuscript as follows (p ?? l ??):
+We have incorporated this discussion in the revised version of the manuscript as follows (p. 23, line 11):
 
@@
 \textbf{Response: } We thank the reviewer for this comment.
-Agreed, the downstream turbines could also be subjected to increased fatigue loading, and we have included this in the revised manuscript as follows (p ??, l ??):
+Agreed, the downstream turbines could also be subjected to increased fatigue loading, and we have included this in the revised manuscript as follows (p.
+24, line 7):
 
@@
 
+\clearpage
+
 \textbf{9. Reviewer: } \textit{On the grammatical note, the manuscript is clear and easy to follow. The only comment

### 2018-05-23 revision/revised_WM02.tex
R1C1 finalized

@@
 A conventionally (greedily) controlled wind farm with steady $C_T' = 2$ was defined as a reference case.
-\revision{Note that this would correspond to a farm with ideal turbines for which generator torque is being controlled dynamically to track the maximum power point at the Betz limit perfectly.} Several different optimal control cases were defined, based on the wind-turbine response time $\tau = $ 0, 5, or 30 s (instantaneous, fast, or slow response, see Eq.
+\revision{Note that this would correspond to a farm with ideal turbines for which generator torque is being controlled dynamically to track the maximum power point at the Betz limit perfectly.
+In a real turbine controller this may, e.g., be achieved with the ESC proposed by \cite{ciri2017large}.} Several different optimal control cases were defined, based on the wind-turbine response time $\tau = $ 0, 5, or 30 s (instantaneous, fast, or slow response, see Eq.
 \ref{eq:ctihat_constraint}) and the maximal thrust coefficient $C_{T,\text{max}}' = 2$ or~3 \revision{, with thrust forces that can respectively only be reduced (underinductive), or also increased (overinductive) compared to the Betz optimum at $C_T' = 2$} (see Eq.
@@
 An interesting observation from this figure is that , for any row $i$ except the last one, the power potential as observed in case C3t5 is almost fully recovered by only applying the optimal controls up to row $i-1$.
-This suggests that self-optimization \revision{(i.e.
-temporarily outperforming conventional greedy control in response to incoming flow events)} is very limited: the optimal controls for a given turbine are designed to create favorable flow conditions in the downstream rows instead of increasing local power.
+This suggests that self-optimization is very limited: the optimal controls for a given turbine are designed to create favorable flow conditions in the downstream rows instead of increasing local power.
 Furthermore, although the discussion in the previous paragraph has shown that downstream controls are optimized with the upstream actions in mind, the converse is not true: upstream control actions do not require a specific downstream response in order to increase power in that downstream row.

### 2018-05-23 revision/reply_to_revs_WM2.tex
Final updates + change line numbers

@@
 \textbf{Note that this would correspond to a farm with ideal turbines for which generator torque is being controlled dynamically to track the maximum power point at the Betz limit perfectly.
-In a real turbine controller this may, e.g., be achieved with the ESC proposed by Ciri et al.
+In a real turbine controller this may, e.g., be achieved with the extremum seeking control proposed by Ciri et al.
 (2017).} Several different optimal control cases were defined, ...
@@
 
-\textbf{Response:} Indeed, this is correct. We have mentioned this explicitly in the revised manuscript as follows (p. 4, line 25):
+\textbf{Response:} Indeed, this is correct. We have mentioned this explicitly in the revised manuscript as follows (p. 4, line 26):
 
@@
 
-We have slightly modified the statement in the revised manuscript to avoid any confusion with regard to possibly having the chord length as a control variable. (p. 17, line 29):
+We have slightly modified the statement in the revised manuscript to avoid any confusion with regard to possibly having the chord length as a control variable. (p. 17, line 25):
 
 ``
-The resulting maximum thrust coefficient of 3.5 can, e.g., be \textbf{attained by the NREL 5MW turbine with a slight modification in the rotor design}, i.e.
-a 50\% increase in blade chord length and an operational tip speed ratio 25\% higher than
+\textbf{For instance, considering the NREL 5MW blade profiles, the maximum thrust coefficient of 3.5 can be attained by slightly changing the rotor design, e.g.
+using a 50\% increase in blade chord length and an operational tip speed ratio 25\% higher than
 the original design value (see Goit and Meyers, 2015, Appendix A).
-\textbf{Furthermore, given such redesign, dynamic reductions from this value could be realized through blade pitch control, for which actuation rates in the order of $10^\circ/$s are possible (see, e.g., Jonkman et al.
+Furthermore, given such redesign, dynamic reductions from this value could be realized through blade pitch control, for which actuation rates in the order of $10^\circ/$s are possible (see, e.g., Jonkman et al.
 2009).}
@@
 We have updated the revised manuscript as follows (p.
-4, line 20):
+4, line 19):
 
@@
 
-and (p. 4, line 27):
+and (p. 4, line 28):
 
@@
 
-We have added the requested TI values in the discussion around Figure 4 as follows (p. 6, line 19):
+We have added the requested TI values in the discussion around Figure 4 as follows (p. 5, line 20):
 
@@
 See the beginning of Section 3 (p.
-8, line 5):
+8, line 2):
 
@@
 
-We have incorporated this in the revised manuscript as follows (p. 14, line 7):
+We have incorporated this in the revised manuscript as follows (p. 14, line 4):
 
@@
 
-and (p. 14, line 14):
+and (p. 14, line 11):
 
@@
 We have added this comment to the discussion as follows (p.
-17, line 29):
+17, line 25):
 
 ``
-The resulting maximum thrust coefficient of 3.5 can, e.g., be \textbf{attained by the NREL 5MW turbine with a slight modification in the rotor design}, i.e.
-a 50\% increase in blade chord length and an operational tip speed ratio 25\% higher than
+\textbf{For instance, considering the NREL 5MW blade profiles, the maximum thrust coefficient of 3.5 can be attained by slightly changing the rotor design, e.g.
+using a 50\% increase in blade chord length and an operational tip speed ratio 25\% higher than
 	the original design value (see Goit and Meyers, 2015, Appendix A).
-\textbf{Furthermore, given such redesign, dynamic reductions from this value could be realized through blade pitch control, for which actuation rates in the order of $10^\circ/$s are possible (see, e.g., Jonkman et al.
+Furthermore, given such redesign, dynamic reductions from this value could be realized through blade pitch control, for which actuation rates in the order of $10^\circ/$s are possible (see, e.g., Jonkman et al.
 2009).}
@@
 
-We have incorporated this discussion in the revised version of the manuscript as follows (p. 23, line 11):
+We have incorporated this discussion in the revised version of the manuscript as follows (p. 23, line 8):
 
@@
 Agreed, the downstream turbines could also be subjected to increased fatigue loading, and we have included this in the revised manuscript as follows (p.
-24, line 7):
+24, line 4):
 

### 2018-05-23 revision/revised_WM02.tex
Final updates + change line numbers

@@
 \revision{Note that this would correspond to a farm with ideal turbines for which generator torque is being controlled dynamically to track the maximum power point at the Betz limit perfectly.
-In a real turbine controller this may, e.g., be achieved with the ESC proposed by \cite{ciri2017large}.} Several different optimal control cases were defined, based on the wind-turbine response time $\tau = $ 0, 5, or 30 s (instantaneous, fast, or slow response, see Eq.
+In a real turbine controller this may, e.g., be achieved with the extremum seeking control proposed by \cite{ciri2017large}.} Several different optimal control cases were defined, based on the wind-turbine response time $\tau = $ 0, 5, or 30 s (instantaneous, fast, or slow response, see Eq.
 \ref{eq:ctihat_constraint}) and the maximal thrust coefficient $C_{T,\text{max}}' = 2$ or~3 \revision{, with thrust forces that can respectively only be reduced (underinductive), or also increased (overinductive) compared to the Betz optimum at $C_T' = 2$} (see Eq.
@@
 For instance, a Strouhal number $St = 0.25$ corresponds here to a sine wave period of $\approx$ 50 s for a turbine with diameter $D = 100$ m and a freestream velocity $U_\infty = 8.5$ m s$^{-1}$.
-The resulting maximum thrust coefficient of 3.5 can, e.g., be \revision{attained by the NREL 5MW turbine with a slight modification in the rotor design}, i.e.
-a 50\% increase in chord length at a tip speed ratio that is 25\% higher than the greedy tip speed ratio (see \citealp{goit2015optimal}, Appendix A).
-\revision{Furthermore, given such redesign, dynamic reductions from this value could be realized through blade pitch control, for which actuation rates in the order of $10^\circ/$s are possible (see, e.g., \citealp{jonkman2009definition}).} Figure~\ref{fig:sinus_baseline}b illustrates normalized power extraction by row for the reference case, the best sinusoidal case, and the first four rows of the optimal control case C3t5 from MM17.
+\revision{For instance, considering the NREL 5MW blade profiles, the maximum thrust coefficient of 3.5 can be attained by slightly changing the rotor design, e.g.
+using a 50\% increase in blade chord length and an operational tip speed ratio 25\% higher than the original design value (see \citealp{goit2015optimal}, Appendix A).
+Furthermore, given such redesign, dynamic reductions from this value could be realized through blade pitch control, for which actuation rates in the order of $10^\circ/$s are possible (see, e.g., \citealp{jonkman2009definition}).}
+Figure~\ref{fig:sinus_baseline}b illustrates normalized power extraction by row for the reference case, the best sinusoidal case, and the first four rows of the optimal control case C3t5 from MM17.
 The figure shows that the power gain in the sinusoidal cases originates mostly from the second row, and that power in the first row is decreased by approximately 5\%.
